---
title: "Day 4 (AM): Data Visualization with ggplot2"
subtitle: "SC INBRE Biostatistics Summer Intensive 2026"
date: today
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    number-sections: true
    smooth-scroll: true
    theme: cosmo
    code-fold: false
    highlight-style: github
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---


# Overview

By the end of this session you will be able to:

- Describe the grammar of graphics and how `ggplot2` implements it
- Choose the right geom for the data and question at hand
- Visualize distributions, relationships, and group comparisons
- Use `labs()`, `theme()`, scales, and facets to produce publication-ready figures
- Annotate plots and save them to disk with `ggsave()`

We use two datasets throughout: the **Framingham Heart Study** (which you
already know) and the built-in `mpg` and `diamonds` datasets for some
general plotting concepts.

---

# The Grammar of Graphics

`ggplot2` is built on the idea that every plot can be described by the same
small set of components:

| Component | What it controls |
|---|---|
| **Data** | The tibble you are visualizing |
| **Aesthetics** (`aes()`) | How variables map to visual properties (x, y, color, size, shape) |
| **Geoms** | The geometric objects that represent the data (points, lines, bars, boxes) |
| **Scales** | How data values map to aesthetic values (axis limits, color palettes) |
| **Facets** | Splitting the plot into small multiples by a grouping variable |
| **Theme** | Non-data appearance (fonts, background, grid lines) |

Every `ggplot2` plot follows the same template:

```{r}
#| eval: false
ggplot(data = <DATA>) +
  <GEOM>(mapping = aes(<MAPPINGS>)) +
  <SCALES> +
  <FACETS> +
  <THEME>
```


---

# Exploratory Data Analysis: Two Fundamental Questions

Exploratory data analysis (EDA) is the practice of systematically examining your
data before modeling. Two questions drive EDA:

1. **What is the variation within each variable?** What is the typical value?
How spread out are the values? Are there outliers?
2. **What is the covariation between variables?** How do two or more variables
move together?

Your choice of plot depends on the types of variables involved:

| Variables | Plot type |
|---|---|
| 1 continuous | Histogram, density, boxplot |
| 1 categorical | Bar chart |
| 1 continuous + 1 categorical | Boxplot, violin, frequency polygon |
| 2 continuous | Scatterplot, smoothed line, 2D bin |
| 2 categorical | Mosaic, count heatmap |

---

```{r}
library(tidyverse)
library(this.path)
setwd(this.dir())

framingham <- read_csv("../../data/framingham.csv") |>
  mutate(
    male          = factor(male,          levels = c(0,1), labels = c("Female","Male")),
    currentSmoker = factor(currentSmoker, levels = c(0,1), labels = c("Non-smoker","Smoker")),
    prevalentHyp  = factor(prevalentHyp,  levels = c(0,1), labels = c("No","Yes")),
    diabetes      = factor(diabetes,      levels = c(0,1), labels = c("No","Yes")),
    TenYearCHD    = factor(TenYearCHD,    levels = c(0,1), labels = c("No","Yes")),
    education     = factor(education,     levels = 1:4,
                           labels = c("Some HS","HS Diploma","Some College","College Degree"))
  )
str(framingham)
```

# Visualizing Distributions

## Continuous Variables: Histograms

A histogram bins a continuous variable and counts observations per bin. Always
try several `binwidth` values --- the pattern can change substantially.

First, we'll use the default binwidth value, which ends up being 8 (I think):

```{r}
ggplot(framingham, aes(x = sysBP)) +
  geom_histogram(fill = "steelblue", color = "white") +
  labs(x = "Systolic Blood Pressure (mmHg)", y = "Count",
       title = "Distribution of Systolic Blood Pressure")
```


```{r}
ggplot(framingham, aes(x = sysBP)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "white") +
  labs(x = "Systolic Blood Pressure (mmHg)", y = "Count",
       title = "Distribution of Systolic Blood Pressure")
```

```{r}
# Narrower bins reveal more detail
ggplot(framingham, aes(x = sysBP)) +
  geom_histogram(binwidth = 2, fill = "steelblue", color = "white") +
  labs(x = "Systolic Blood Pressure (mmHg)", y = "Count")
```

The lower the binwidth, the more bumpy the histogram is going to look.

## Continuous Variables: Density Plots

A density plot is a smoothed histogram. It is better than a histogram for
overlaying multiple groups because lines are easier to compare than bars.

```{r}
ggplot(framingham, aes(x = sysBP, fill = male)) +
  # alpha will control how transparent the fill colors are.
  geom_density(alpha = 0.4) +
  labs(x = "Systolic Blood Pressure (mmHg)", y = "Density",
       fill = "Sex",
       title = "Systolic BP Distribution by Sex")
```

```{r}
ggplot(framingham, aes(x = sysBP, fill = male)) +
  # alpha will control how transparent the fill colors are.
  geom_density(alpha = 1) +
  labs(x = "Systolic Blood Pressure (mmHg)", y = "Density",
       fill = "Sex",
       title = "Systolic BP Distribution by Sex")
```

## Categorical Variables: Bar Charts

```{r}
ggplot(framingham, aes(x = education)) +
  geom_bar(fill = "steelblue") +
  labs(x = "Education Level", y = "Count",
       title = "Education Distribution in Framingham Cohort") +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
```

To plot proportions rather than counts, compute them first with `dplyr` and
use `geom_col()`:

```{r}
framingham |>
  count(education) |>
  mutate(pct = n / sum(n) * 100) |>
  ggplot(aes(x = education, y = pct)) +
  geom_col(fill = "steelblue") +
  labs(x = "Education Level", y = "Percent (%)",
       title = "Education Distribution (%)") +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
```

---

# Visualizing Covariation

## Continuous + Categorical: Boxplots

Boxplots show the median, IQR, and outliers compactly. They are ideal for
comparing a continuous variable across groups.

```{r}
ggplot(framingham, aes(x = male, y = sysBP)) +
  geom_boxplot(fill = "steelblue", alpha = 0.7) +
  labs(x = "Sex", y = "Systolic BP (mmHg)",
       title = "Systolic Blood Pressure by Sex")
```

When you have many groups, order them by the median to make comparisons easier:

```{r}
ggplot(framingham, aes(x = reorder(education, sysBP, FUN = median),
                       y = sysBP)) +
  geom_boxplot(fill = "steelblue", alpha = 0.7) +
  labs(x = "Education Level",
       y = "Systolic BP (mmHg)",
       title = "Systolic BP by Education Level",
       subtitle = "Groups ordered by median BP") +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
```

## Continuous + Categorical: Violin Plots

A violin plot shows the full distribution shape, not just the summary statistics.
Combine with a boxplot overlay for the best of both:

```{r}
ggplot(framingham, aes(x = male, y = sysBP)) +
  geom_violin(fill = "steelblue", alpha = 0.5) +
  geom_boxplot(width = 0.1, fill = "white", outlier.size = 0.8) +
  labs(x = "Sex", y = "Systolic BP (mmHg)",
       title = "Systolic BP by Sex: Violin + Boxplot")
```

## Continuous + Categorical: Frequency Polygons

`geom_freqpoly()` overlays distributions as lines rather than bars. 

```{r}
ggplot(framingham, aes(x = sysBP, color = male)) +
  geom_freqpoly(binwidth = 5, linewidth = 1) +
  labs(x = "Systolic BP (mmHg)", y = "Count", color = "Sex")
```

When groups differ greatly in size, use `y = after_stat(density)` to standardize:

```{r}
# Density version -- better for comparing groups of unequal size
ggplot(framingham, aes(x = sysBP, y = after_stat(density), color = male)) +
  geom_freqpoly(binwidth = 5, linewidth = 1) +
  labs(x = "Systolic BP (mmHg)", y = "Density", color = "Sex")
```

## Two Continuous Variables: Scatterplots

```{r}
ggplot(framingham, aes(x = age, y = sysBP)) +
  geom_point(alpha = 0.3) +
  labs(x = "Age (years)", y = "Systolic BP (mmHg)",
       title = "Systolic BP vs. Age")
```

Add a smooth trend line with `geom_smooth()`:

```{r}
#| message: false
ggplot(framingham, aes(x = age, y = sysBP)) +
  geom_point(alpha = 0.2) +
  # method = "loess" lines can show non-linear trends.
  geom_smooth(method = "loess", color = "firebrick", se = TRUE) +
  labs(x = "Age (years)", y = "Systolic BP (mmHg)",
       title = "Systolic BP vs. Age with Loess Smoother")
```

Map a third variable to color:

```{r}
#| message: false
ggplot(framingham, aes(x = age, y = sysBP, color = male)) +
  geom_point(alpha = 0.3) +
  # method = "lm" only shows linear trends.
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Age (years)", y = "Systolic BP (mmHg)",
       color = "Sex",
       title = "Systolic BP vs. Age by Sex")
```

### Log Transformations

When a variable is right-skewed, a log transform often reveals structure hidden
in the raw scale. Use `scale_x_log10()` to keep the original axis labels:

```{r}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_bin2d() +
  labs(title = "Diamond Price vs. Carat: Raw scale")

ggplot(diamonds, aes(x = carat, y = price)) +
  geom_bin2d() +
  scale_x_log10() +
  scale_y_log10() +
  labs(title = "Diamond Price vs. Carat: Log-log scale")
```

### Overplotting

Large datasets suffer from overplotting --- points pile on top of each other and
the density is invisible. Solutions:

```{r}
# Alpha transparency
ggplot(framingham, aes(x = age, y = sysBP)) +
  geom_point(alpha = 0.1) +
  labs(title = "Alpha transparency")
```

```{r}
# 2D bins
ggplot(framingham, aes(x = age, y = sysBP)) +
  geom_bin2d(bins = 30) +
  scale_fill_viridis_c() +
  labs(title = "2D bin plot")
```

---

# Faceting

Faceting splits a plot into small multiples by a variable --- often cleaner than
color coding when there are many groups.

```{r}
#| message: false
ggplot(framingham, aes(x = age, y = sysBP)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "lm", se = FALSE, color = "firebrick") +
  facet_wrap(~ male) +
  labs(x = "Age", y = "Systolic BP (mmHg)",
       title = "Systolic BP vs. Age, by Sex")
```

You can allow the x, y, or both scales to be unique across plots.

```{r}
#| message: false
ggplot(framingham, aes(x = age, y = sysBP)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "lm", se = FALSE, color = "firebrick") +
  facet_wrap(~ male, scales = "free") +
  labs(x = "Age", y = "Systolic BP (mmHg)",
       title = "Systolic BP vs. Age, by Sex")
```


`facet_grid()` facets on two variables simultaneously:

```{r}
#| message: false
ggplot(framingham, aes(x = age, y = sysBP)) +
  geom_point(alpha = 0.2, size = 0.8) +
  geom_smooth(method = "lm", se = FALSE, color = "firebrick") +
  facet_grid(currentSmoker ~ male) +
  labs(x = "Age", y = "Systolic BP (mmHg)",
       title = "Systolic BP vs. Age by Sex and Smoking Status")
```

You can do the same with `facet_wrap` but the labels are not as nice:

```{r}
#| message: false
ggplot(framingham, aes(x = age, y = sysBP)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "lm", se = FALSE, color = "firebrick") +
  facet_wrap(~ male + currentSmoker, scales = "free") +
  labs(x = "Age", y = "Systolic BP (mmHg)",
       title = "Systolic BP vs. Age, by Sex")
```

---

# Labels and Annotation

## `labs()`: Titles, Axes, and Legends

```{r}
framingham |> 
  filter(!is.na(education)) |> # Removing those with NA for eductation (you could include)
  ggplot(aes(x = reorder(education, sysBP, FUN = median),
                         y = sysBP, fill = male)) +
  geom_boxplot(alpha = 0.7) +
  labs(
    title    = "Systolic Blood Pressure by Education and Sex",
    subtitle = "Framingham Heart Study cohort",
    caption  = "Source: Framingham Heart Study",
    x        = "Education Level",
    y        = "Systolic BP (mmHg)",
    fill     = "Sex"
  ) +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
```

## `geom_text()` and `geom_label()`

Label individual points or groups directly on the plot. Here we label the car
model with the best highway mileage in each class:

```{r}
best_in_class <- mpg |>
  group_by(class) |>
  filter(row_number(desc(hwy)) == 1)

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_label(aes(label = model), data = best_in_class,
             nudge_y = 2, alpha = 0.6, size = 3) +
  labs(x = "Engine displacement (L)", y = "Highway mpg",
       color = "Vehicle class")
```

For overlapping labels, use `ggrepel` which automatically repositions them:

```{r}
# install.packages("ggrepel")
library(ggrepel)

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_point(size = 3, shape = 1, data = best_in_class) +
  geom_label_repel(aes(label = model), data = best_in_class, size = 3) +
  labs(x = "Engine displacement (L)", y = "Highway mpg",
       color = "Vehicle class")
```

## Reference Lines

Add horizontal or vertical reference lines to mark thresholds:

```{r}
ggplot(framingham, aes(x = age, y = sysBP)) +
  geom_point(alpha = 0.2) +
  geom_hline(yintercept = 140, color = "blue",
             linewidth = 1, linetype = "dashed") +
  annotate("text", x = 35, y = 143,
           label = "Hypertension threshold (140 mmHg)",
           color = "blue", hjust = 0, size = 7) +
  labs(x = "Age (years)", y = "Systolic BP (mmHg)",
       title = "Systolic BP with Hypertension Threshold")
```

---

# Scales

Scales control how data values map to visual properties. `ggplot2` names them
consistently: `scale_<aesthetic>_<type>()`.

## Axis Breaks and Labels

```{r}
ggplot(framingham, aes(x = age, y = sysBP)) +
  geom_point(alpha = 0.2) +
  scale_x_continuous(breaks = seq(30, 70, by = 10)) +
  scale_y_continuous(breaks = seq(80, 300, by = 40)) +
  labs(x = "Age (years)", y = "Systolic BP (mmHg)")
```

## Color Scales

The default color palette is not always ideal --- especially for color-blind
readers. ColorBrewer palettes are designed for this:

```{r}
ggplot(framingham, aes(x = age, y = sysBP, color = education)) +
  geom_point(alpha = 0.4) +
  scale_color_brewer(palette = "Set2") +
  labs(x = "Age", y = "Systolic BP", color = "Education")
```

For continuous color scales, use viridis which is both colorblind-friendly and
perceptually uniform:

```{r}
ggplot(framingham, aes(x = age, y = sysBP, color = BMI)) +
  geom_point(alpha = 0.5) +
  scale_color_viridis_c(option = "plasma") +
  labs(x = "Age", y = "Systolic BP", color = "BMI")
```

For a diverging scale (e.g., positive/negative, above/below a midpoint):

```{r}
framingham |>
  mutate(sysBP_centered = sysBP - mean(sysBP, na.rm = TRUE)) |>
  ggplot(aes(x = age, y = sysBP_centered, color = sysBP_centered)) +
  geom_point(alpha = 0.5) +
  scale_color_gradient2(low = "steelblue", mid = "white",
                        high = "firebrick", midpoint = 0) +
  labs(x = "Age", y = "Systolic BP (centered)", color = "BP\n(centered)")
```

## Zooming with `coord_cartesian()`

To zoom in without dropping data (which `xlim`/`ylim` inside `scale_` would do):

```{r}
#| message: false
ggplot(framingham, aes(x = age, y = sysBP)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm") +
  coord_cartesian(xlim = c(40, 60), ylim = c(100, 200)) +
  labs(title = "Zoomed: ages 40-60, BP 100-200 mmHg")
```

---

# Themes

Themes control all non-data elements: background, grid lines, fonts, and text size.

```{r}
#| message: false
base <- ggplot(framingham, aes(x = age, y = sysBP, color = male)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Age", y = "Systolic BP", color = "Sex")

base + theme_gray()    # default
base + theme_bw()      # white background, grid lines
base + theme_minimal() # minimal: no background, light grid
base + theme_classic() # no grid, classic axes
```

Fine-tune individual elements with `theme()`:

```{r}
#| message: false
base +
  theme_minimal() +
  theme(
    plot.title       = element_text(size = 14, face = "bold"),
    axis.title       = element_text(size = 12),
    legend.position  = "bottom",
    panel.grid.minor = element_blank()
  ) +
  labs(title = "Systolic BP vs. Age by Sex")
```

---

# Saving Plots

Use `ggsave()` to write the most recent plot to disk. Always specify `width` and
`height` for reproducibility:

```{r}
#| eval: false
ggplot(framingham, aes(x = age, y = sysBP, color = male)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal() +
  labs(x = "Age (years)", y = "Systolic BP (mmHg)", color = "Sex")

ggsave("figures/sysBP_vs_age.png", width = 7, height = 5, dpi = 300)
ggsave("figures/sysBP_vs_age.pdf", width = 7, height = 5)
```

::: {.callout-tip}
Use `.pdf` for vector graphics (scalable, best for publication) and `.png` for
raster graphics (best for slides and web). Always set `dpi = 300` for `.png`
to ensure print quality.
:::

When working in Quarto, chunk options control figure output directly --- no need
for `ggsave()` in most cases:

````markdown
```{{r}}
#| fig-width: 7
#| fig-height: 5
#| fig-cap: "Systolic BP vs. Age by Sex"
#| fig-alt: "Scatterplot of systolic blood pressure on the y-axis and age on
#|            the x-axis, with separate trend lines by sex."
```
````

---

# Putting It Together: A Publication-Ready Figure

Here is a complete example building a figure from scratch with all the elements
covered today:

```{r}
#| message: false
#| fig-width: 8
#| fig-height: 5
framingham |>
  filter(!is.na(education)) |>
  ggplot(aes(x = age, y = sysBP, color = education)) +
  geom_point(alpha = 0.15, size = 0.8) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1) +
  geom_hline(yintercept = 140, linetype = "dashed",
             color = "gray40", linewidth = 0.8) +
  annotate("text", x = 32, y = 143,
           label = "Hypertension threshold", color = "gray40",
           hjust = 0, size = 3) +
  scale_color_brewer(palette = "Set2") +
  scale_y_continuous(breaks = seq(80, 300, by = 40)) +
  facet_wrap(~ male) +
  theme_minimal() +
  theme(
    legend.position  = "bottom",
    panel.grid.minor = element_blank(),
    strip.text       = element_text(face = "bold")
  ) +
  labs(
    title    = "Systolic Blood Pressure vs. Age by Sex and Education Level",
    subtitle = "Framingham Heart Study cohort — linear trends per education group",
    caption  = "Dashed line indicates hypertension threshold (140 mmHg)",
    x        = "Age (years)",
    y        = "Systolic BP (mmHg)",
    color    = "Education level"
  )
```


If you think this is crazy, here's some actual code of mine that I recently used to create a figure:

```{r}
#| eval: false
cols4methods <- c("black","lightsteelblue4"   ,
                  "firebrick1" ,         "palevioletred2",
                  "cornflowerblue","blue1", "darkgoldenrod4" , 
                  "maroon2",               "chartreuse4" , "coral")

shape_vals <- c(15:24)

pos_wid <- 0.88 ## controls the spread of the points and the width of the bars
pos_siz <- 2.5 ## size of the points

# ..... Lots of code missing ......

## Creating the plot. The 'pdf' can be uncommented when creating a pdf of the plot.
mse_plot_fpc <- ggplot(d_plot_mse_fpc %>% filter(M == 5625), aes(x=as.factor(n.basis), y=mean_mse, color = Method)) + 
  scale_color_manual(values=cols4methods[c(1,4,5,6)]) +
  geom_errorbar(aes(ymin = min_mser, ymax = max_mser), 
                width = pos_wid, 
                position=position_dodge(width=pos_wid)) + 
  facet_wrap(~sparse_gr) +
  theme_bw() + 
  theme_light() +
  geom_point(aes(shape=Method), size=pos_siz, position=position_dodge(width=pos_wid)) + 
  scale_shape_manual(values=shape_vals[c(1,4,5,6)]) +
  xlab("Number of Basis Functions") +
  ylab("") +
  theme_bw(base_size = 10, base_family = "Helvetica") +
  theme(strip.background = element_blank(),
        strip.placement = "outside",
        axis.text=element_text(color="#666666",size=8),
        axis.title = element_text(color="#666666",
                                  face="bold", size=11),
        strip.text.x = element_text(size=10),
        strip.text.y = element_text(size=10))  

# ..... Lots of code missing ......


mse_plot <- ggplot(d_plot_mse, aes(x=re, y=mean_mse, color = Method)) + 
  scale_color_manual(values=cols4methods) +
  geom_errorbar(aes(ymin = min_mser, ymax = max_mser), 
                width = pos_wid, 
                position=position_dodge(width=pos_wid)) + 
  facet_wrap(~M_gr) +
  theme_bw() + 
  theme_light() +
  geom_point(aes(shape=Method), size=pos_siz, position=position_dodge(width=pos_wid)) + 
  scale_shape_manual(values=shape_vals) +
  xlab("Number of random effects") +
  ylab("Scaled RMSE") +
  theme_bw(base_size = 10, base_family = "Helvetica") +
  theme(strip.background = element_blank(),
        strip.placement = "outside",
        axis.text=element_text(color="#666666",size=8),
        axis.title = element_text(color="#666666",
                                  face="bold", size=11),
        strip.text.x = element_text(size=10),
        strip.text.y = element_text(size=10))  


library(ggpubr)

final_plot <- ggarrange(
  mse_plot, mse_plot_fpc, 
  labels = c("A", "B"),
  common.legend = TRUE,
  legend = "right",
  nrow = 1)

ggsave("Mse_figA_all.pdf", width = 7, height = 3)
```


```{r}
#| echo: false
#| fig-cap: "Here's the final figure"
#| out-width: "100%"
#| out-height: 3in
knitr::include_graphics("figures/Mse_figA_all.pdf")
```

---

# Exercises

**Exercise 1.** Make a histogram of BMI in the Framingham data. Try at least
three different `binwidth` values. Which reveals the most useful structure?
Add a vertical line at BMI = 30 (the obesity threshold) using `geom_vline()`
and annotate it with `annotate()`.

**Exercise 2.** Compare the distribution of total cholesterol (`totChol`) across
education levels using (a) overlapping density plots with `geom_density()` and
(b) boxplots ordered by median cholesterol. Which display is more informative
and why? Add proper labels and a theme to both.

**Exercise 3.** Make a scatterplot of `age` vs. `totChol`, colored by
`currentSmoker`. Add separate linear trend lines for smokers and non-smokers
using `geom_smooth(method = "lm")`. Facet by sex. Use a ColorBrewer palette and
`theme_bw()`. What do you notice about the relationship between age and
cholesterol for smokers vs. non-smokers?

**Exercise 4.** Using the `diamonds` dataset, make a scatterplot of `carat` vs.
`price`. The raw scale will suffer from severe overplotting --- solve this with
`geom_bin2d()` or `alpha`. Then apply `scale_x_log10()` and `scale_y_log10()`.
Label the axes with the original units (carats and dollars) and add a title
summarizing the main finding.

**Exercise 5 (Challenge).** Build a single publication-ready figure using
Framingham data that shows the 10-year CHD rate (proportion with
`TenYearCHD == "Yes"`) by age group and sex. Steps: (a) create an age group
variable with `cut()` or `case_when()` (e.g., 30--39, 40--49, 50--59, 60+),
(b) compute CHD rate per group using `group_by()` and `summarize()`, (c) plot
with `geom_col()` or `geom_line()`, (d) facet or color by sex, (e) add all
labels, a sensible theme, and `ggsave()` it to a file.
