---
title: "Day 4 (PM) Lab: Data Visualization with ggplot2"
subtitle: "SC INBRE Biostatistics Summer Intensive 2026"
date: today
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    number-sections: true
    smooth-scroll: true
    theme: cosmo
    code-fold: false
    highlight-style: github
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

# Overview

This lab covers everything from Day 4 AM:

- **Distributions:** histograms, density plots, bar charts
- **Covariation:** boxplots, violin plots, scatterplots, frequency polygons
- **Polish:** `labs()`, `theme()`, scales, facets, annotations, `ggsave()`

**How this lab works:**

- **Part 1 (instructor demo, ~20 min):** Follow along as the instructor builds
  a complete figure from raw data to publication-ready output.
- **Part 2 (on your own, ~90 min):** Six exercises using the Framingham data
  and the built-in `nycflights13` dataset. Create a new `.qmd` file, work
  through the exercises, render when done.

The **Quick Reference Card** in Section 3 has the key functions.

---

# Part 1: Instructor Demo

## Setup

```{r}
library(tidyverse)
library(this.path)
setwd(this.dir())

framingham <- read_csv("../../data/framingham.csv") |>
  mutate(
    male          = factor(male,          levels = c(0,1), labels = c("Female","Male")),
    currentSmoker = factor(currentSmoker, levels = c(0,1), labels = c("Non-smoker","Smoker")),
    prevalentHyp  = factor(prevalentHyp,  levels = c(0,1), labels = c("No","Yes")),
    diabetes      = factor(diabetes,      levels = c(0,1), labels = c("No","Yes")),
    TenYearCHD    = factor(TenYearCHD,    levels = c(0,1), labels = c("No","Yes")),
    education     = factor(education,     levels = 1:4,
                           labels = c("Some HS","HS Diploma","Some College","College Degree"))
  )
```

## Step 1: Start Simple

Always start with the simplest version of the plot and build from there.

```{r}
ggplot(framingham, aes(x = age, y = sysBP)) +
  geom_point()
```

## Step 2: Reduce Overplotting, Add a Trend

```{r}
#| message: false
ggplot(framingham, aes(x = age, y = sysBP)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "lm", color = "firebrick", se = TRUE)
```

## Step 3: Add a Grouping Variable

```{r}
#| message: false
ggplot(framingham, aes(x = age, y = sysBP, color = male)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "lm", se = FALSE)
```

## Step 4: Facet

```{r}
#| message: false
ggplot(framingham, aes(x = age, y = sysBP, color = currentSmoker)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ male)
```

## Step 5: Polish — Labels, Theme, Color Scale, Annotation

```{r}
#| message: false
#| fig-width: 8
#| fig-height: 5
ggplot(framingham, aes(x = age, y = sysBP, color = currentSmoker)) +
  geom_point(alpha = 0.15, size = 0.8) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1) +
  geom_hline(yintercept = 140, linetype = "dashed",
             color = "gray40", linewidth = 0.8) +
  annotate("text", x = 32, y = 144,
           label = "Hypertension threshold (140 mmHg)",
           hjust = 0, size = 3, color = "gray40") +
  facet_wrap(~ male) +
  scale_color_brewer(palette = "Set1") +
  theme_minimal() +
  theme(
    legend.position  = "bottom",
    panel.grid.minor = element_blank(),
    strip.text       = element_text(face = "bold")
  ) +
  labs(
    title    = "Systolic Blood Pressure vs. Age by Sex and Smoking Status",
    subtitle = "Framingham Heart Study — linear trends per group",
    caption  = "Dashed line indicates hypertension threshold (140 mmHg)",
    x        = "Age (years)",
    y        = "Systolic Blood Pressure (mmHg)",
    color    = "Smoking status"
  )
```

This is the workflow for every figure: start plain, layer in complexity, polish
last. The plot code becomes your record of every decision you made.

---

# Part 2: Exercises

## Instructions

**Step 1.** Open a new Quarto document: File → New File → Quarto Document.
Title it `"Day 4 Lab: Data Visualization"`, select HTML, click **Create**.

**Step 2.** Replace the template with this YAML and setup chunk:

````markdown
---
title: "Day 4 Lab: Data Visualization"
subtitle: "SC INBRE Biostatistics Summer Intensive 2026"
date: today
format:
  html:
    toc: true
    theme: cosmo
execute:
  echo: true
  warning: false
  message: false
---

```{{r}}
library(tidyverse)
library(this.path)
library(nycflights13)
setwd(this.dir())

framingham <- read_csv("../../data/framingham.csv") |>
  mutate(
    male          = factor(male,          levels = c(0,1), labels = c("Female","Male")),
    currentSmoker = factor(currentSmoker, levels = c(0,1), labels = c("Non-smoker","Smoker")),
    prevalentHyp  = factor(prevalentHyp,  levels = c(0,1), labels = c("No","Yes")),
    diabetes      = factor(diabetes,      levels = c(0,1), labels = c("No","Yes")),
    TenYearCHD    = factor(TenYearCHD,    levels = c(0,1), labels = c("No","Yes")),
    education     = factor(education,     levels = 1:4,
                           labels = c("Some HS","HS Diploma","Some College","College Degree"))
  )
```
````

**Step 3.** Work through Exercises 1--6 in order. For each one:

- Add a new code chunk (`Ctrl+Alt+I` / `Cmd+Option+I`)
- Build the plot incrementally --- start plain, then add polish
- Write 2--3 sentences below the plot describing what it shows and what you
  notice

**Step 4.** Render (`Ctrl+Shift+K` / `Cmd+Shift+K`) when finished and fix
any errors.

::: {.callout-tip}
**Build incrementally.** Write the minimal `ggplot()` + geom first and confirm
it runs. Then add one element at a time. This is far easier to debug than
writing the full polished version all at once.
:::

::: {.callout-note}
Exercises 1--4 cover the core material. Exercise 5 requires joining data before
plotting (a Day 3 skill). Exercise 6 is the challenge --- a complete figure
built from scratch. Don't worry if you don't reach it.
:::

---

**Exercise 1 — Distributions.**

Using the Framingham data:

a. Make a histogram of BMI. Experiment with at least two `binwidth` values and
   pick the one that reveals the most useful structure. Add a vertical line at
   BMI = 30 (obesity threshold) using `geom_vline()` and label it with
   `annotate()`.

b. Overlay density plots of BMI separately for diabetic and non-diabetic
   participants using `geom_density()` with `fill` mapped to `diabetes` and
   `alpha = 0.4`. Use `scale_fill_brewer(palette = "Set1")`. What do you
   notice about the distributions?

c. Make a bar chart showing the count of participants at each education level.
   Tilt the x-axis labels so they don't overlap (`angle = 20, hjust = 1` inside
   `element_text()`). Then make a second version showing the **percentage**
   at each level rather than the count (compute percentages first with `dplyr`,
   then use `geom_col()`).

```{r}
# your code here
```

---

**Exercise 2 — Comparing Groups.**

a. Make a boxplot of glucose (`glucose`) by education level. Order the boxes by
   median glucose using `reorder()`. Add `coord_flip()` to make the labels
   readable. Apply `theme_bw()` and proper axis labels.

b. Repeat using `geom_violin()` overlaid with a narrow `geom_boxplot()`
   (`width = 0.1`). Which display do you prefer and why?

c. Make a frequency polygon (`geom_freqpoly()`) of systolic BP (`sysBP`) with
   one line per education level (`color = education`). Use
   `y = after_stat(density)` so groups of different sizes are comparable.
   Which education group has the broadest distribution?

```{r}
# your code here
```

---

**Exercise 3 — Scatterplots and Smoothers.**

a. Make a scatterplot of age vs. total cholesterol (`totChol`). With over 4,000
   points there will be overplotting --- use `alpha` to address it. Add a loess
   smoother (`method = "loess"`) in red with `se = TRUE`. Describe the overall
   trend.

b. Repeat but color points by `currentSmoker` and add **separate** linear trend
   lines for each group (`method = "lm"`, `se = FALSE`). Does the relationship
   between age and cholesterol differ between smokers and non-smokers?

c. Facet the plot from (b) by sex (`facet_wrap(~ male)`). Add a complete set
   of labels and apply `theme_minimal()`. Are there any sex differences in the
   smoking–cholesterol relationship?

```{r}
# your code here
```

---

**Exercise 4 — Scales and Polish.**

a. Make a scatterplot of age vs. sysBP, with BMI mapped to `color`. Use
   `scale_color_viridis_c(option = "plasma")` for the color scale. What does
   the color gradient reveal about the relationship between BMI and BP?

b. Take any plot you have made so far and fully polish it to publication
   quality. This means: informative title, subtitle, and caption; axis labels
   with units; a clean theme; a sensible color palette; and axis breaks set
   explicitly with `scale_x_continuous(breaks = ...)` or
   `scale_y_continuous(breaks = ...)`. Save it to a `figures/` subfolder using
   `ggsave("figures/my_plot.png", width = 7, height = 5, dpi = 300)`.

```{r}
# your code here
```

---

**Exercise 5 — Joining Before Plotting.**

This exercise combines Day 3 joins with Day 4 visualization using `nycflights13`.

a. Join `flights` with `weather` on `origin` and `time_hour` to add visibility
   (`visib`) and temperature (`temp`) to each flight. Call this `flights_wx`.

b. Compute the mean arrival delay by rounded visibility level
   (`mutate(visib = round(visib))`), grouped by origin airport. Then plot mean
   delay (y) vs. visibility (x) with a separate colored line per origin airport
   using `geom_line()` and `geom_point()`. Use `scale_color_brewer(palette = "Dark2")`.
   Is there a relationship between visibility and delays? Does it differ by airport?

c. Make a boxplot of departure delay (`dep_delay`) by `origin` airport, limited
   to delays between --20 and 120 minutes (use `filter()` first, or
   `coord_cartesian()`). Order the boxes by median delay, apply
   `theme_classic()`, and add proper labels.

```{r}
# your code here
```

---

**Exercise 6 (Challenge) — A Complete Figure.**

Build a single polished figure that answers this question: **How does the
10-year coronary heart disease rate in the Framingham cohort vary by age group,
sex, and hypertension status?**

Steps:

a. Create an age group variable using `case_when()`:
   "40--44", "45--49", "50--54", "55--59", "60--64", "65+"
   (or similar grouping of your choosing). Make it a factor with levels in
   the right order.

b. Compute the CHD rate (proportion with `TenYearCHD == "Yes"`) for each
   combination of age group, sex, and hypertension status using
   `group_by()` and `summarize()`. Keep only groups with at least 20
   participants.

c. Plot the result. A grouped bar chart (`geom_col()` with `position = "dodge"`)
   or a line plot (`geom_line()` + `geom_point()`) with age group on the x-axis
   works well here. Color by hypertension status, facet by sex.

d. Add a full set of labels, a meaningful color palette, and a clean theme.

e. Save the figure with `ggsave()` at 300 dpi.

f. Write a 3--4 sentence caption summarizing the main findings as if this were
   going into a manuscript.

```{r}
# your code here
```

---

# Quick Reference Card

## The ggplot2 Template

```r
ggplot(data, aes(x = var1, y = var2, color = var3)) +
  geom_*() +
  scale_*() +
  facet_wrap(~ var4) +
  theme_*() +
  theme(...) +
  labs(title = "...", x = "...", y = "...", color = "...")
```

## Geoms

| Geom | Use case |
|---|---|
| `geom_histogram(binwidth)` | Distribution of one continuous variable |
| `geom_density(alpha)` | Smoothed distribution; good for overlaying groups |
| `geom_bar()` | Counts of a categorical variable |
| `geom_col()` | Pre-computed values as bars |
| `geom_boxplot()` | Distribution summary by group |
| `geom_violin()` | Full distribution shape by group |
| `geom_freqpoly(binwidth)` | Overlapping distributions as lines |
| `geom_point(alpha, size)` | Scatterplot |
| `geom_smooth(method, se)` | Trend line: `"lm"`, `"loess"` |
| `geom_line()` | Connected points (time series, summaries) |
| `geom_bin2d()` | 2D density for large scatterplots |
| `geom_hline(yintercept)` | Horizontal reference line |
| `geom_vline(xintercept)` | Vertical reference line |
| `geom_text(aes(label))` | Text labels at data points |
| `geom_label(aes(label))` | Text labels with background box |

## Aesthetics (inside `aes()`)

| Aesthetic | Controls |
|---|---|
| `x`, `y` | Position |
| `color` / `colour` | Point/line color |
| `fill` | Bar/box/violin fill color |
| `alpha` | Transparency (0 = invisible, 1 = solid) |
| `size` | Point size or line width |
| `shape` | Point shape |
| `linetype` | Line type (solid, dashed, dotted) |
| `group` | Grouping without visual change |

## Scales

| Function | Use case |
|---|---|
| `scale_x_continuous(breaks, limits)` | Customize x-axis ticks |
| `scale_y_continuous(breaks, limits)` | Customize y-axis ticks |
| `scale_x_log10()` | Log-transform x-axis |
| `scale_color_brewer(palette)` | ColorBrewer categorical palette |
| `scale_color_viridis_c()` | Viridis continuous color scale |
| `scale_color_gradient2(low, mid, high)` | Diverging color scale |
| `scale_fill_brewer(palette)` | ColorBrewer fill palette |
| `coord_cartesian(xlim, ylim)` | Zoom without dropping data |
| `coord_flip()` | Flip x and y axes |

## Labels and Annotation

| Function | Use case |
|---|---|
| `labs(title, subtitle, caption, x, y, color)` | Add/change all labels |
| `annotate("text", x, y, label)` | Add a text annotation at a fixed position |
| `geom_hline(yintercept, linetype, color)` | Horizontal reference line |
| `geom_vline(xintercept, linetype, color)` | Vertical reference line |

## Themes

| Function | Style |
|---|---|
| `theme_gray()` | Default gray background |
| `theme_bw()` | White background, grid lines |
| `theme_minimal()` | Minimal — no background, light grid |
| `theme_classic()` | No grid, classic axis lines |
| `theme_void()` | Empty — only geoms visible |
| `theme(legend.position = "bottom")` | Move legend |
| `theme(axis.text.x = element_text(angle = 20, hjust = 1))` | Tilt x labels |
| `theme(panel.grid.minor = element_blank())` | Remove minor grid lines |

## Faceting

```r
facet_wrap(~ var)              # one variable, automatic layout
facet_wrap(~ var, ncol = 2)    # force 2 columns
facet_grid(row_var ~ col_var)  # two variables, row × column grid
```

## Saving

```r
ggsave("figures/myplot.png", width = 7, height = 5, dpi = 300)
ggsave("figures/myplot.pdf", width = 7, height = 5)
```
