---
title: "Day 2 (PM) Lab: Working with Data in R"
subtitle: "SC INBRE Biostatistics Summer Intensive 2026"
date: today
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    number-sections: true
    smooth-scroll: true
    theme: cosmo
    code-fold: false
    highlight-style: github
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

# Overview

This lab draws on everything covered so far:

- **Day 1:** R basics, object types, indexing with `$` and `[ ]`, mathematical
  functions, reading data, inspecting with `str()` / `summary()` / `head()`
- **Day 2 AM:** Variable types, factors, strings, `dplyr` verbs (`filter()`,
  `mutate()`, `select()`, `arrange()`, `group_by()`, `summarize()`), and pipes

The dataset is the **Framingham Heart Study**, a landmark longitudinal study of
cardiovascular disease risk factors. The file `framingham.csv` is in the `data/`
folder. It contains 4,240 participants and 16 variables.

**How this lab works:**

- **Part 1 (instructor demo, ~25 min):** Follow along as the instructor works
  through the setup, data inspection, and variable type correction. Run the
  code on your own machine as you go.
- **Part 2 (on your own, ~90 min):** Complete the exercises independently. The
  quick reference card at the bottom of this document has the key functions you
  need. The instructor is in the room for questions.

---

# Part 1: Instructor Demo

## Setup

```{r}
#| echo: false
library(tidyverse)
library(this.path)

this.dir <- this.dir()
setwd(this.dir)



framingham <- read_csv('../../data/framingham.csv')
```

```{r}
#| eval: false
library(tidyverse)
library(rstudioapi) ## you may need to install
this.dir <- dirname(rstudioapi::getSourceEditorContext()$path)
setwd(this.dir)

# The following does the same and sometimes works better with .qmd
# library(this.path) 
# this.dir <- this.dir()
# setwd(this.dir)

framingham <- read_csv("../../data/framingham.csv")
```

::: {.callout-note}
We use `read_csv()` (tidyverse) rather than base R's `read.csv()`. It returns
a tibble, prints more cleanly, and never silently converts strings to factors.
:::

## Getting to Know the Data

```{r}
dim(framingham)     # rows and columns
names(framingham)   # variable names
```

```{r}
head(framingham)    # first 6 rows
```

```{r}
str(framingham)     # structure: type of every variable + example values
```

```{r}
summary(framingham) # five-number summary for numeric; NA counts visible here
```

```{r}
#| eval: false
View(framingham)    # open full spreadsheet in RStudio viewer
```

**Look for:** Which variables have `NA` counts? Which numeric variables are
clearly categorical (e.g., coded 0/1 or 1--4)?

### Missing data

```{r}
colSums(is.na(framingham))                                  # count per column
round(colSums(is.na(framingham)) / nrow(framingham) * 100, 1) # percent missing
```

## Fixing Variable Types

Several variables are stored as numbers but represent categories. Left as-is,
a regression model will treat them as continuous --- almost always wrong.

| Variable | Coding |
|---|---|
| `male` | 0 = Female, 1 = Male |
| `currentSmoker` | 0 = Non-smoker, 1 = Smoker |
| `BPMeds` | 0 = No, 1 = Yes |
| `prevalentStroke` | 0 = No, 1 = Yes |
| `prevalentHyp` | 0 = No, 1 = Yes |
| `diabetes` | 0 = No, 1 = Yes |
| `TenYearCHD` | 0 = No event, 1 = CHD within 10 years |
| `education` | 1 = Some HS, 2 = HS Diploma, 3 = Some College, 4 = College Degree |

```{r}
framingham <- framingham |>
  mutate(
    male            = factor(male,            levels = c(0, 1),
                                              labels = c("Female", "Male")),
    currentSmoker   = factor(currentSmoker,   levels = c(0, 1),
                                              labels = c("Non-smoker", "Smoker")),
    BPMeds          = factor(BPMeds,          levels = c(0, 1),
                                              labels = c("No", "Yes")),
    prevalentStroke = factor(prevalentStroke, levels = c(0, 1),
                                              labels = c("No", "Yes")),
    prevalentHyp    = factor(prevalentHyp,    levels = c(0, 1),
                                              labels = c("No", "Yes")),
    diabetes        = factor(diabetes,        levels = c(0, 1),
                                              labels = c("No", "Yes")),
    TenYearCHD      = factor(TenYearCHD,      levels = c(0, 1),
                                              labels = c("No", "Yes")),
    education       = factor(education,       levels = 1:4,
                                              labels = c("Some HS", "HS Diploma",
                                                         "Some College",
                                                         "College Degree"))
  )
```

Run `str()` again to confirm the columns now show as `<fct>`, and `summary()`
to see that categorical variables now show counts per level:

```{r}
str(framingham)
summary(framingham)
```

## One Worked Example

Here is a complete pipeline combining what we have done. Read each step before
running:

```{r}
# Among participants NOT on BP medication,
# compute mean systolic BP and CHD rate by hypertension status and sex
framingham |>
  filter(BPMeds == "No") |>
  group_by(prevalentHyp, male) |>
  summarize(
    n          = n(),
    mean_sysBP = mean(sysBP, na.rm = TRUE),
    sd_sysBP   = sd(sysBP, na.rm = TRUE),
    chd_pct    = mean(TenYearCHD == "Yes", na.rm = TRUE) * 100,
    .groups    = "drop"
  ) |>
  arrange(desc(chd_pct))
```

::: {.callout-tip}
This is the template for most of the exercises below: `filter()` to narrow the
data, `mutate()` to create new variables, `group_by()` + `summarize()` to
compute group statistics, then `arrange()` to sort the result.
:::

---

# Part 2: Exercises

## Instructions

The demo is over --- this part is yours. Here is how to get started:

**Step 1.** Open a new Quarto document: File → New File → Quarto Document.
Give it a title like `"Day 2 Lab"` and select HTML. Click **Create**.

**Step 2.** Delete the template content that Quarto pre-fills and replace it
with a YAML header and setup chunk matching the one from Part 1:

````markdown
---
title: "Day 2 Lab: Working with Data"
subtitle: "SC INBRE Biostatistics Summer Intensive 2026"
date: today
format:
  html:
    toc: true
    theme: cosmo
execute:
  echo: true
  warning: false
  message: false
---

```{{r}}
library(tidyverse)
library(this.path) 
this.dir <- this.dir()
setwd(this.dir)

framingham <- read_csv("../../data/framingham.csv")
```
````

**Step 3.** Re-run the variable type correction block from Part 1 in your new
document. All six exercises depend on the factor conversions being in place ---
if you skip this, your code will not behave as expected.

**Step 4.** Work through Exercises 1--6 in order. For each one:

- Add a new code chunk (`Ctrl+Alt+I` / `Cmd+Option+I`)
- Write and run your code
- Add a sentence or two below the chunk describing what you found in plain
  language --- not just what the numbers are, but what they mean

**Step 5.** When you have finished all exercises, click **Render**
(`Ctrl+Shift+K` / `Cmd+Shift+K`) and confirm the document compiles to HTML
without errors. Fix any errors before moving on.

::: {.callout-tip}
**Before asking for help**, check two things: (1) the **Quick Reference Card**
in the next section of this document, and (2) whether you have `na.rm = TRUE` where
needed --- this is the most common reason summaries return `NA` unexpectedly.
:::

::: {.callout-note}
It is fine if you do not finish all six exercises during the session. Exercises
1--5 cover the core material. Exercise 6 is a challenge problem for those who
finish early. Prioritize understanding over speed --- a working solution you can
explain is worth more than a finished document you cannot.
:::

---

**Exercise 1.** Using `filter()` and `summarize()`, compute the mean, standard
deviation, and median of total cholesterol (`totChol`) separately for diabetic
and non-diabetic participants. Do diabetic participants tend to have higher
cholesterol?

```{r}
# your code here
```

---

**Exercise 2.** Create two new variables using `mutate()`:

- `pulse_pressure`: systolic BP minus diastolic BP (`sysBP - diaBP`)
- `bmi_category`: a factor with four levels using `case_when()`:
  `"Underweight"` (BMI < 18.5), `"Normal"` (18.5--24.9),
  `"Overweight"` (25--29.9), `"Obese"` (≥ 30)

After creating these, compute the mean pulse pressure by BMI category.

```{r}
# your code here
```

---

**Exercise 3.** Using **base R indexing** (`$` and `[ ]`), extract the subset
of participants who are female, non-smokers, and over age 55. How many are
there? What is their mean glucose level? Now repeat using `filter()` and
confirm you get the same row count.

```{r}
# base R version

# dplyr version
```

---

**Exercise 4.** Among current smokers only, use `summarize()` to compute the
mean, standard deviation, and range of `cigsPerDay`. Then use `mutate()` to
add a logical variable `heavy_smoker` (TRUE if `cigsPerDay > 20`) to the full
dataset and count how many heavy smokers there are.

```{r}
# your code here
```

---

**Exercise 5.** Compute the 10-year CHD rate (percentage with
`TenYearCHD == "Yes"`) by education level. Sort from highest to lowest rate.
Does education appear to be associated with CHD risk?

```{r}
# your code here
```

---

**Exercise 6 (Challenge).** Build a single pipeline that:

(a) Keeps only participants aged 40--65
(b) Creates a new variable `hyp_smoker`: `"Yes"` if the participant has
    prevalent hypertension **and** is a current smoker, `"No"` otherwise
    (use `if_else()` or `case_when()`)
(c) Computes, for each combination of `hyp_smoker` and `male`: number of
    participants, mean systolic BP, and 10-year CHD rate
(d) Keeps only groups with at least 50 participants
(e) Sorts by CHD rate descending

What combination of characteristics is associated with the highest CHD risk
in this age group?

```{r}
# your code here
```

---

# Quick Reference Card

Use this while working on the exercises. For more detail, check the Day 1 and
Day 2 AM documents.

## Reading and Inspecting Data

| Function | What it does |
|---|---|
| `read_csv("file.csv")` | Read a CSV file; returns a tibble |
| `dim(df)` | Number of rows and columns |
| `names(df)` | Column names |
| `str(df)` | Structure: type and example values of every column |
| `summary(df)` | Summary statistics; shows NA counts |
| `head(df, n)` | First `n` rows (default 6) |
| `View(df)` | Open spreadsheet viewer in RStudio |
| `colSums(is.na(df))` | Count of missing values per column |

## Indexing (Base R)

| Syntax | What it does |
|---|---|
| `df$col` | Extract column as a vector |
| `df[i, ]` | Row `i`, all columns |
| `df[, j]` | All rows, column `j` |
| `df[condition, ]` | Rows where condition is TRUE |
| `df[["col"]]` | Extract column by name (useful when name is in a variable) |

## Variable Types

| Function | What it does |
|---|---|
| `factor(x, levels, labels)` | Convert to factor with named levels |
| `levels(f)` | Show the levels of a factor |
| `class(x)` | Check the type of an object |
| `is.na(x)` | Logical vector: TRUE where values are missing |

## dplyr Verbs

| Function | What it does |
|---|---|
| `filter(df, condition)` | Keep rows where condition is TRUE |
| `select(df, col1, col2)` | Keep specified columns |
| `mutate(df, new = expr)` | Add or modify columns |
| `arrange(df, col)` | Sort rows; `desc(col)` for descending |
| `group_by(df, col)` | Group rows for subsequent summaries |
| `summarize(df, stat = expr)` | Collapse groups to summary rows |
| `n()` | Count rows (use inside `summarize()`) |

## Useful Expressions Inside mutate / summarize

| Expression | What it does |
|---|---|
| `mean(x, na.rm = TRUE)` | Mean, ignoring NAs |
| `sd(x, na.rm = TRUE)` | Standard deviation |
| `median(x, na.rm = TRUE)` | Median |
| `sum(x, na.rm = TRUE)` | Sum |
| `min(x, na.rm = TRUE)` / `max(x, na.rm = TRUE)` | Min / max |
| `sum(is.na(x))` | Count missing values |
| `mean(var == "Yes")` | Proportion where condition is TRUE |
| `case_when(cond1 ~ val1, cond2 ~ val2, TRUE ~ val3)` | Recode with multiple conditions |
| `if_else(condition, true_val, false_val)` | Simple two-way recode |

## Pipes

```r
df |> filter(...) |> mutate(...) |> summarize(...)
# read as: "take df, then filter, then mutate, then summarize"
```
