---
title: "Day 2 (AM): R Programming Fundamentals"
subtitle: "SC INBRE Biostatistics Summer Intensive 2026"
date: today
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    number-sections: true
    smooth-scroll: true
    theme: cosmo
    code-fold: false
    highlight-style: github
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| echo: false
library(tidyverse)
```

# Overview

By the end of this session you will be able to:

- Recognize the main object types in R and know when each arises
- Understand the variable types you will encounter in biological datasets
- Manipulate and transform data using core `dplyr` verbs
- Understand what a function is and read a simple one

---

# Object Types in R: A Quick Tour

R has several object types. We will cover them briefly here because you will see
them mentioned in help files and error messages. In practice, **you will almost
always be working with tibbles** (see below), so don't worry about memorizing
these --- just know they exist.

## Vectors

A vector is the simplest object: a sequence of values all of the same type.
Everything else in R is built on top of vectors.

```{r}
nums <- c(1.5, 2.3, 7.1, 4.8)
chars <- c("control", "treatment", "control", "treatment")
logicals <- c(TRUE, FALSE, TRUE, TRUE)
```

## Matrices

A matrix is a 2D vector --- all values must be the same type. Useful for linear
algebra, but rarely used to store biological data directly.

```{r}
m <- matrix(1:9, nrow = 3, ncol = 3)
m
```

## Lists

A list is a collection of objects that can be of different types and lengths ---
think of it as a container that can hold anything. Model output in R (e.g., from
`lm()`) is returned as a list.

```{r}
my_list <- list(
  name   = "Patient A",
  values = c(1.2, 3.4, 5.6),
  flag   = TRUE
)
my_list
my_list$name
my_list$values
```

## Data Frames and Tibbles

A data frame is a rectangular table where each column is a vector. A **tibble**
is the tidyverse version of a data frame with some sensible defaults: it prints
more neatly, never converts strings to factors automatically, and gives cleaner
error messages.

::: {.callout-important}
**This is the object type you will use for virtually everything in this course.**
When you read in a dataset with `readr` (e.g., `read_csv()`), you get a tibble.
When you manipulate data with `dplyr`, you get a tibble back. You will rarely
need to think about matrices or bare lists for your data. However, R functions will commonly give back results as lists. So, knowing how to work with them is important.
:::

```{r}
# A small tibble created by hand
patient_data <- tibble(
  id        = 1:5,
  treatment = c("A", "B", "A", "B", "A"),
  response  = c(12.1, 8.4, 14.3, 9.1, 11.7)
)
patient_data
```

Notice what prints above the column values: `<int>`, `<chr>`, `<dbl>`. These are
the **variable types**, which we cover next.

---

# Indexing in R

Indexing means extracting specific elements from an object --- a single value, a
column, a row, or a subset. Knowing how to do this in base R is important because
you will encounter it constantly in help files, Stack Overflow answers, and other
people's code, even if you end up preferring tidyverse syntax for most tasks.

## Indexing Vectors with `[ ]`

Use single square brackets to extract elements from a vector by position:

```{r}
x <- c(10, 20, 30, 40, 50)

x[1]        # first element (R is 1-indexed, not 0-indexed)
x[3]        # third element
x[c(1, 3)]  # first and third elements
x[2:4]      # elements 2 through 4
x[-1]       # everything except the first element
```

You can also index by a logical vector of the same length. Elements where the
condition is `TRUE` are kept:

```{r}
x[x > 25]           # elements greater than 25
x[c(TRUE, FALSE, TRUE, FALSE, TRUE)]  # explicit logical index
```

Named vectors can be indexed by name:

```{r}
bp <- c(systolic = 120, diastolic = 80)
bp["systolic"]
bp[c("systolic", "diastolic")]
```

## Indexing Data Frames with `[ , ]`

Data frames use two-dimensional indexing: `[rows, columns]`. Leaving one
dimension blank means "all":

```{r}
patient_data <- data.frame(
  id        = 1:5,
  treatment = c("A", "B", "A", "B", "A"),
  response  = c(12.1, 8.4, 14.3, 9.1, 11.7)
)
patient_data             # all data
patient_data[1, ]        # first row, all columns
patient_data[, 2]        # all rows, second column (returns a vector)
patient_data[1, 2]       # first row, second column (single value)
patient_data[1:3, ]      # rows 1 through 3
patient_data[, c(1, 3)]  # columns 1 and 3
```

Columns can also be referenced by name inside `[ ]`:

```{r}
patient_data[, "response"]
patient_data[patient_data$treatment == "A", ]  # rows where treatment is "A"
```

## The `$` Operator

`$` is the most common way to extract a single column from a data frame or
element from a list. It returns the column as a vector:

```{r}
patient_data$response        # the response column as a vector
patient_data$treatment       # the treatment column as a vector
mean(patient_data$response)  # pass directly to a function
```

::: {.callout-tip}
In RStudio, typing `patient_data$` will trigger autocomplete showing all
available column names. This is one of the most useful features of the IDE
for data exploration.
:::

`$` works identically on lists:

```{r}
my_list <- list(name = "Patient A", values = c(1.2, 3.4, 5.6))
my_list$name
my_list$values[2]  # chain indexing: second element of the values vector
```

## Double Brackets `[[ ]]`

Double brackets extract a single element by position or name, stripping the
container. The difference from `$` is subtle but matters for lists and
programmatic access:

```{r}
patient_data[["response"]]   # same result as patient_data$response
my_list[["values"]]          # same as my_list$values

# Useful when the column name is stored in a variable
col_name <- "response"
patient_data[[col_name]]     # $ does not work here; [[ ]] does
```

## Putting It Together: Subsetting a Data Frame

A common task is extracting a subset of rows and columns based on a condition:

```{r}
# Patients in treatment A with response above 12
patient_data[patient_data$treatment == "A" & patient_data$response > 12, ]
```

::: {.callout-note}
This base R approach is verbose but important to recognize. In the next section
we will use `dplyr::filter()` and `dplyr::select()` to do the same thing more
readably. Both approaches are valid --- you will see both in the wild.
:::

## Quick Reference

| Goal | Syntax |
|------|--------|
| Vector element by position | `x[2]` |
| Vector elements by condition | `x[x > 5]` |
| Data frame column as vector | `df$col` or `df[["col"]]` |
| Data frame rows by condition | `df[df$col == "A", ]` |
| Data frame single cell | `df[2, 3]` |
| List element | `lst$name` or `lst[["name"]]` |

---

# Variable Types

The type of a variable determines what operations you can do with it and how R
and statistical functions will treat it. Getting variable types right is one of
the most important --- and most commonly skipped --- steps in data preparation.

We will use the `flights` dataset from the `nycflights13` package throughout.
Install it once if needed: `install.packages("nycflights13")`.

```{r}
library(nycflights13)
flights
```

The abbreviations under each column name tell you the type:

| Abbreviation | Type | Example |
|---|---|---|
| `<int>` | Integer | `year`, `month`, `day` |
| `<dbl>` | Double (numeric, decimal) | `dep_delay`, `air_time` |
| `<chr>` | Character (string) | `carrier`, `origin` |
| `<lgl>` | Logical (TRUE/FALSE) | e.g., `is_delayed` |
| `<fct>` | Factor (categorical with levels) | e.g., severity grades |
| `<dttm>` | Date-time | `time_hour` |

## Numeric Variables (`<dbl>`, `<int>`)

Numeric variables represent measured quantities. Arithmetic and summary functions
apply directly.

```{r}
mean(flights$dep_delay, na.rm = TRUE)
sd(flights$dep_delay, na.rm = TRUE)
range(flights$air_time, na.rm = TRUE)
```

::: {.callout-note}
`na.rm = TRUE` tells R to ignore missing values when computing the summary.
Without it, a single `NA` in the vector will return `NA` for the whole
calculation.
:::

## Character Variables (`<chr>`)

Character variables (strings) hold text. They cannot be used in arithmetic but
can be searched, split, and recoded.

```{r}
# How many unique carriers are there?
unique(flights$carrier)

# How many flights per carrier?
table(flights$carrier)
```

Common string operations from `stringr` (all start with `str_`):

```{r}
x <- c("Control Group", "Treatment Group A", "Treatment Group B")

str_to_lower(x)          # convert to lowercase
str_to_upper(x)          # convert to uppercase
str_length(x)            # number of characters
str_detect(x, "Control") # does the string contain "Control"?
str_replace(x, "Group", "Arm") # replace a pattern
```

`str_detect()` is especially useful for filtering:

```{r}
# Keep only carriers whose code starts with "U"
table(str_detect(flights$carrier, "^U"))
```

## Logical Variables (`<lgl>`)

Logicals are `TRUE` or `FALSE`. They arise naturally from comparisons and are
useful for filtering and counting.

```{r}
# Create a logical variable: was the flight delayed on arrival?
table(flights$arr_delay > 0)
```

`sum()` and `mean()` treat `TRUE` as 1 and `FALSE` as 0, so you can count and
compute proportions easily:

```{r}
sum(flights$arr_delay > 0, na.rm = TRUE)
mean(flights$arr_delay > 0, na.rm = TRUE)*100
```

## Factors (`<fct>`)

Factors represent **categorical variables with a fixed set of levels**. They are
essential in statistical modeling --- a factor in a regression model is handled
as a categorical predictor (with a reference level), while a character variable
or numeric code is not.

```{r}
# Create a factor
severity <- factor(c("mild", "severe", "moderate", "mild", "severe"),
                   levels = c("mild", "moderate", "severe"))
severity
levels(severity)
table(severity)
```

::: {.callout-important}
**When does this matter for you?** If you code a categorical variable as a number
(e.g., 1 = mild, 2 = moderate, 3 = severe) and put it in a model without
converting it to a factor, R will treat it as a continuous variable and fit a
single slope. This is almost never what you want. Always check whether your
grouping variables are stored as factors. **This is one of the most common errors when running models in R.**
:::

Use `factor()` to convert, and `fct_recode()` from the `forcats` package
(part of the tidyverse) to rename levels:

```{r}
flights$origin_fct = factor(flights$origin)
flights$origin_fct[1:5]
```

The first level is the reference level, to change use `relevel` (important for model interpretation):

```{r}
severity_reref <- relevel(severity, ref = "moderate")
levels(severity_reref)
```

### Recoding and Collapsing Levels

`fct_recode()` lets you rename levels; `fct_collapse()` merges them:

```{r}
severity_recoded <- fct_recode(severity,
  "Grade 1" = "mild",
  "Grade 2" = "moderate",
  "Grade 3" = "severe"
)
levels(severity_recoded)
```

```{r}
# Collapse moderate and severe into a single "high" level
fct_collapse(severity,
  "low"  = "mild",
  "high" = c("moderate", "severe")
)
```

---

# Data Transformations with dplyr

`dplyr` is the workhorse package for data manipulation. Its functions are called
**verbs** because each one does one thing clearly. We will cover the ones you will
use in almost every analysis.

## The Pipe

The pipe (`|>`) passes the output of one function as the first argument of the
next. Read it as "then":

```{r}
#| eval: false
flights |>           # start with flights, then
  filter(...) |>     # keep certain rows, then
  mutate(...) |>     # create new columns, then
  summarize(...)     # collapse to a summary
```

You may also see `%>%` from the `magrittr` package --- it behaves identically
for our purposes.

## Filtering Rows: `filter()`

Keep rows that satisfy a condition:

```{r}
# Flights that arrived more than 2 hours late
flights |>
  filter(arr_delay > 120)

# Flights departing in January on United or American
flights |>
  filter(month == 1 & 
           carrier %in% c("UA", "AA")
         )
```

Common operators: `>`, `>=`, `<`, `<=`, `==`, `!=`, `%in%`, `is.na()`.

::: {.callout-note}
`=` assigns a value; `==` tests for equality. Using `=` inside `filter()` is
one of the most common beginner errors in R.
:::

```{r}
# Flights departing in January not on United or American
flights |>
  filter(month == 1 & 
           !(carrier %in% c("UA", "AA")) 
         )
```

## Selecting and Renaming Columns: `select()` and `rename()`

```{r}
# Keep only a few columns
flights |>
  select(year, month, day, carrier, dep_delay, arr_delay)

# Select a range of columns
flights |>
  select(year:dep_time)

# Rename a column (**new name = old name**)
flights |>
  rename(tail_number = tailnum) |>
  select(tail_number, carrier)
```

Useful helpers inside `select()`: `starts_with()`, `ends_with()`,
`contains()`, `where(is.numeric)`.

```{r}
# Keep only columns that contain "time"
flights |>
  select(contains("time"))
```

## Creating and Modifying Columns: `mutate()`

`mutate()` adds new columns or overwrites existing ones:

```{r}
flights |>
  mutate(
    gain       = dep_delay - arr_delay,      # time made up in air
    speed_mph  = distance / air_time * 60,   # speed in mph
    late       = arr_delay > 0               # logical flag
  ) |>
  select(carrier, gain, speed_mph, late)
```

`mutate()` is also where you will do type conversions:

```{r}
flights |>
  mutate(
    month_fct = factor(month,
                       ## The original month variable is 1:12.
                       ## We tell R what the original levels are here.
                       levels = 1:12, 
                       labels = month.abb)  # month.abb = Jan, Feb, ... (already exists)
  ) |>
  count(month_fct)
```

## Recoding Variables: `if_else()` and `case_when()`

Two functions come up constantly when creating new variables inside `mutate()`.

`if_else()` handles a simple two-way recode:

```{r}
flights |>
  mutate(long_flight = if_else(air_time > 300, "Long", "Short")) |>
  count(long_flight)
```

`case_when()` handles three or more conditions. Each condition is written as
`condition ~ value`, and `.default` catches anything not matched:

```{r}
flights |>
  mutate(
    delay_category = case_when(
      dep_delay <= 0          ~ "On time or early",
      dep_delay <= 30         ~ "Minor delay (<= 30 min)",
      dep_delay <= 120        ~ "Moderate delay (31-120 min)",
      .default                = "Major delay (> 120 min)"
    )
  ) |>
  count(delay_category)
```

Conditions are evaluated top to bottom --- the first match wins. Make sure
your most specific conditions come first.

## Sorting: `arrange()`

```{r}
# Most delayed departures first
flights |>
  arrange(desc(dep_delay)) |>
  select(carrier, origin, dest, dep_delay)
```

What days have the most flights?

```{r}
flights |>
  mutate(
    # Create a date variable
    date = make_date(year = year, month = month, day = day) 
  ) |>
  # Calculate the number of flights per day.
  count(date) |> 
  # Display in decreasing order by the number of flights.
  arrange(desc(n))
```

## Grouped Summaries: `group_by()` + `summarize()`

This combination is how you compute group-level statistics:

```{r}
# Mean arrival delay by carrier
flights |>
  group_by(carrier) |>
  summarize(
    mean_delay = mean(arr_delay, na.rm = TRUE),
    n_flights  = n()
  ) |>
  arrange(desc(mean_delay))
```

::: {.callout-note}
Always include `na.rm = TRUE` inside summary functions like `mean()`, `sd()`,
and `median()`. Missing values in the input will return `NA` for the whole
summary otherwise.
:::

Multiple grouping variables:

```{r}
flights |>
  group_by(origin, carrier) |>
  summarize(
    mean_delay = mean(arr_delay, na.rm = TRUE),
    n          = n(),
    .groups    = "drop"    # good practice: drop grouping after summarizing
  )
```

## Putting It Together: A Pipeline

Here is a realistic pipeline: compute the on-time arrival rate by carrier for
flights longer than 500 miles, and keep only carriers with at least 1,000 such
flights.

```{r}
flights |>
  filter(distance > 500) |>
  mutate(on_time = arr_delay <= 0) |>
  group_by(carrier) |>
  summarize(
    n           = n(),
    pct_on_time = mean(on_time, na.rm = TRUE) * 100
  ) |>
  filter(n >= 1000) |>
  arrange(desc(pct_on_time))
```

---

# Introduction to Functions

## What Is a Function?

You have been using functions since Day 1: `mean()`, `filter()`, `mutate()`.
A function is a reusable block of code that takes **inputs** (arguments) and
returns an **output**.

Writing your own functions becomes valuable once you find yourself copying and
pasting the same block of code more than twice. Instead of repeating code ---
and risking subtle errors each time --- you write it once, name it, and call it.

::: {.callout-note}
**You will not be writing many functions in this workshop.** The goal here is
to understand the concept so that you can read functions when you encounter them
and recognize when your own code might benefit from one eventually.
:::

## Function Anatomy

Every function has three parts:

```{r}
#| eval: false
my_function <- function(argument1, argument2) {
  # body: code that uses the arguments
  result <- argument1 + argument2
  result   # the last evaluated expression is returned
}
```

A simple example --- a function that converts Fahrenheit to Celsius:

```{r}
f_to_c <- function(fahrenheit) {
  (fahrenheit - 32) * 5 / 9
}

f_to_c(98.6)   # normal body temperature
f_to_c(212)    # boiling point of water
f_to_c(c(32, 72, 98.6, 104))  # works on a vector too
```

## Default Arguments

Functions can have default values for arguments. The caller only needs to supply
a value if they want to override the default:

```{r}
round_to <- function(x, digits = 2) {
  round(x, digits)
}

round_to(3.14159)       # uses the default: 2 digits
round_to(3.14159, 4)    # override to 4 digits
```

## A Practical Example: A Custom Summary

Imagine you are computing the same set of summaries repeatedly on different
variables. Rather than typing it out each time, you can wrap it in a function:

```{r}
quick_summary <- function(x) {
  tibble(
    n       = sum(!is.na(x)),
    n_miss  = sum(is.na(x)),
    mean    = mean(x, na.rm = TRUE),
    sd      = sd(x, na.rm = TRUE),
    min     = min(x, na.rm = TRUE),
    max     = max(x, na.rm = TRUE)
  )
}

quick_summary(flights$dep_delay)
quick_summary(flights$air_time)
```

The same code, applied to any numeric vector, with one function call. This is
the core benefit: write once, use many times, update in one place.

## When to Write a Function

A useful rule of thumb: if you have copied and pasted a block of code three
times, it is time to write a function. The benefits are:

1. **Readability** --- a well-named function makes code self-documenting
2. **Maintainability** --- one change propagates everywhere
3. **Fewer errors** --- no risk of updating a value in one copy but not another

For now, focus on reading and understanding functions when you encounter them.
As you get more comfortable with R, writing your own will start to feel natural.

---

# Exercises

1. Using the `flights` dataset, create a new column called `dep_hour` that
   extracts the hour of departure from `dep_time` (hint: integer division ---
   `dep_time %/% 100` gives the hour). Then compute the mean departure delay
   by hour of day. Which hour has the worst average delays?

2. The `carrier` column uses two-letter codes. Use `filter()` and `%in%` to
   keep only Delta (`"DL"`), United (`"UA"`), and American (`"AA"`). Then
   compute the mean and standard deviation of arrival delay for each.

3. Create a factor version of `month` with abbreviated month names as labels
   (use `month.abb` as the `labels` argument to `factor()`). Count flights
   by month and confirm the levels print in calendar order.

4. Write a function called `pct_missing` that takes a vector `x` and returns
   the percentage of values that are `NA`. Test it on `flights$dep_delay` and
   `flights$arr_delay`.

5. Build a single pipeline on `flights` that: (a) keeps only flights in the
   summer months (June, July, August), (b) creates a logical column indicating
   whether the flight was more than 30 minutes late on arrival, (c) computes
   the percentage of such flights by origin airport, and (d) sorts the result
   from worst to best.
