---
title: "Day 3 (AM): Importing, Tidying, and Joining Data"
subtitle: "SC INBRE Biostatistics Summer Intensive 2026"
date: today
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    number-sections: true
    smooth-scroll: true
    theme: cosmo
    code-fold: false
    highlight-style: github
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| echo: false
library(tidyverse)
library(nycflights13)
```

# Overview

By the end of this session you will be able to:

- Read CSV, Excel, and other file formats into R
- Understand what "tidy data" means and why it matters
- Pivot data between wide and long formats using `pivot_longer()` and `pivot_wider()`
- Separate and unite columns
- Understand explicit vs. implicit missing values
- Join two data frames together using mutating and filtering joins

---

# Importing Data

## Key Terms

Before reading data into R, three concepts matter:

- **Header:** The first row of a file is usually variable names. R needs to know
  whether to treat it as names or as data values.
- **Delimiter:** The character that separates entries in each row (comma in `.csv`,
  tab in `.tsv`, etc.). R uses this to know where one value ends and the next begins.
- **Missing value:** How your file represents missingness. R recognizes `NA`, but
  files often use `.`, `""`, `N/A`, `999`, or other codes that need to be declared.

## Reading CSV Files with `readr`

```{r}
#| eval: false
library(tidyverse)  # loads readr
library(readxl)     # for Excel files
library(haven)      # for SAS, Stata, SPSS files
```

`readr` is the tidyverse package for reading rectangular data. The most common
functions:

| Function | Reads |
|---|---|
| `read_csv()` | Comma-separated (`.csv`) |
| `read_csv2()` | Semicolon-separated (common in Europe) |
| `read_tsv()` | Tab-separated |
| `read_delim()` | Any delimiter you specify |

All have the same interface --- once you know `read_csv()` the rest are trivial.

## Working Directories and File Paths

R looks for files relative to the **working directory**. Check yours with `getwd()`:

```{r}
getwd()
```

Rather than hard-coding a full file path (which breaks on other people's machines),
use `rstudioapi` or `this.path` to set the working directory to wherever the script lives. Note, `rstudioapi` only works inside of RStudio and can give error messages when used with .qmd files (and in other situations). `this.path` is usually safer.

```{r}
#| eval: false
library(rstudioapi)
this.dir <- dirname(rstudioapi::getSourceEditorContext()$path)
setwd(this.dir)

library(this.path) 
this.dir <- this.dir()
setwd(this.dir)
```

Then reference files with short relative paths: `"data/myfile.csv"` rather than
`"/Users/yourname/Documents/.../myfile.csv"`. 

You can use `"../data/framingham.csv"` to tell R to go back one folder, find the data folder, and then look for framingham.csv. We used this in the previous lab with:

```{r}
#| eval: false
framingham <- read_csv("../../data/framingham.csv")
```

which said go back two folders, find the data folder, and then look for framingham.csv.

::: {.callout-tip}
**Windows users:** File paths use `\` in Windows Explorer, but R requires `/` or
`\\`. Copy a path from Explorer and replace each `\` with `/` before using it in R.
:::

## Reading a CSV

```{r}
#| eval: false
students <- read_csv("data/students.csv")
```

When you run `read_csv()` it prints a message showing the number of rows and
columns, the delimiter used, and how each column was parsed. This is useful for
catching problems early.

### Common Arguments

**Skipping rows** --- useful when files have metadata at the top:

```{r}
#| message: false
read_csv("The first line of metadata
  The second line of metadata
  x,y,z
  1,2,3", skip = 2)
```

**No column names** --- tell R to generate them:

```{r}
#| message: false
read_csv("1,2,3
         4,5,6", col_names = FALSE)
```

**Supply your own column names:**

```{r}
#| message: false
read_csv("1,2,3
         4,5,6", col_names = c("id", "weight", "height"))
```

**Declare missing value codes** --- files often use `.` or `N/A` instead of `NA`:

```{r}
#| message: false
read_csv("a,b,c
         1,2,.", na = ".")
```

### Post-Import Cleaning

After reading data in, two immediate tasks often arise:

**1. Fix column names with spaces.** Column names with spaces require backtick
quoting (`\`Full Name\``) which is error-prone. Use `janitor::clean_names()` to
convert everything to snake_case automatically:

```{r}
library(janitor)
x <- data.frame(caseID = 1, DOB = 2, Other = 3)
x |> 
  clean_names()

# if you prefer camelCase variable names:
x %>%
  clean_names(., "lower_camel")
```

Or rename specific columns with `rename()`:

```{r}
x |> rename(case_id = `caseID`, dob = DOB, other = Other)
```

**2. Fix variable types.** `read_csv()` guesses column types from the data. Always
check with `str()` and correct anything that was misread (e.g., a numeric column
read as character because of a stray word like `"five"`).

## Reading Excel Files

Use `readxl::read_excel()`. You can specify which sheet to read:

```{r}
#| eval: false
library(readxl)
data_xl <- read_excel("data/myfile.xlsx", sheet = "data")
dict_xl <- read_excel("data/myfile.xlsx", sheet = "dictionary")
```

The same arguments apply: `col_names`, `na`, `skip`.

## Writing Data

```{r}
#| eval: false
write_csv(students, "data/students_clean.csv")
```

::: {.callout-note}
**Important:** When you write a tibble to CSV and read it back, factor levels and
other type information are lost --- the file only stores plain text. If you need to
preserve types across sessions, use `saveRDS()` / `readRDS()` to save as an R
binary file instead.
:::

## Other Formats

The `haven` package reads SAS (`.sas7bdat`), Stata (`.dta`), and SPSS (`.sav`)
files with the same interface:

```{r}
#| eval: false
library(haven)
spss_data <- read_sav("data/myfile.sav") #SPSS data
sas_data  <- read_sas("data/myfile.sas7bdat") #SAS data
stata_data <- read_dta("data/myfile.dta") #Stata data
```

SPSS and Stata files often carry variable labels and value labels that `haven`
preserves as attributes --- useful when you are handed someone else's data and
need to understand the coding.

::: {.callout-tip}
You can read in SAS data, you can even write SAS data (using `write_sas`), but SAS data exported from R will not properly be read into SAS. Exported SPSS (`write_sav`) and Stata (`write_dta`) files can be read into their respected software programs.  
:::


SPSS data has particularly nice features:

```{r}
library(haven)
x2 <- labelled_spss(
  1:10,
  c(Good = 1, Okay = 4, Bad = 8),
  na_range = c(9, Inf),
  label = "Quality rating"
)
x2
is.na(x2)
```

SPSS data (with labels) can be read into SPSS or to Stata. To read into SAS you have to export to Stata, export from Stata, and then import into SAS.

---

# Tidy Data

## What Does "Tidy" Mean?

The same underlying data can be represented in many different ways. Consider these
four representations of the same TB case data:

```{r}
table1  # tidy
table2  # observation spread across two rows
table3  # two variables crammed into one column
table4a # cases only, years as columns
table4b # population only, years as columns
```

Only `table1` is **tidy**. Tidy data follows three rules:

1. Each variable is a column; each column is a variable
2. Each observation is a row; each row is an observation
3. Each value is a cell; each cell is a single value

<!-- FIGURE: Diagram showing the three tidy data rules visually
     (variables = columns, observations = rows, values = cells) -->

Why tidy? Because every function in the tidyverse --- `ggplot2`, `dplyr`,
`tidyr`, modeling functions --- expects tidy data. When your data is tidy, tools
work naturally. When it is not, you spend your time fighting the data instead of
analyzing it.

Here is how naturally `table1` works once it is tidy:

```{r}
# Compute rate per 10,000
table1 |>
  mutate(rate = cases / population * 10000)

# Cases per year
table1 |>
  count(year, wt = cases)
```

## Pivoting: Wide to Long with `pivot_longer()`

Real data is often **wide**: observations spread across many columns, with column
names encoding a variable rather than labeling it. The `billboard` dataset is a
classic example --- 76 columns (`wk1` through `wk76`) where each column is
really a value of the variable `week`:

```{r}
billboard
```

`pivot_longer()` collapses these columns into rows. Three key arguments:

- `cols`: which columns to pivot (uses `select()` syntax)
- `names_to`: what to call the new column that holds the old column names
- `values_to`: what to call the new column that holds the values

```{r}
billboard |>
  pivot_longer(
    cols        = starts_with("wk"),
    names_to    = "week",
    values_to   = "rank",
    values_drop_na = TRUE   # drop rows where the song wasn't charting
  )
```

::: {.callout-note}
`values_drop_na = TRUE` removes rows that only exist because of the wide
structure --- songs that weren't charting in a given week don't have a meaningful
`NA`, they simply weren't ranked.
:::

## Pivoting: Long to Wide with `pivot_wider()`

`pivot_wider()` is the inverse --- it spreads rows into columns. Use it when an
observation is scattered across multiple rows, as in `table2`:

```{r}
table2
```

Each country-year observation is split across two rows (one for `cases`, one for
`population`). Fix it with:

```{r}
table2 |>
  pivot_wider(names_from = type, values_from = count)
```

`pivot_wider()` is also useful for creating **display tables** --- not tidy, but
human-readable:

```{r}
diamonds |>
  count(clarity, color) |>
  pivot_wider(names_from = color, values_from = n)
```

## Separating and Uniting Columns

### Separating Columns

`separate_wider_delim()` splits a column on a delimiter character.

```{r}
table3  # rate column contains "cases/population"

table3 |>
  separate_wider_delim(rate, delim = "/", names = c("cases", "population"))
```

The resulting columns are character by default. Pipe into `mutate()` to
convert types after splitting:

```{r}
table3 |>
  separate_wider_delim(rate, delim = "/", names = c("cases", "population")) |>
  mutate(cases = as.integer(cases), population = as.integer(population))
```

`separate_wider_position()` splits at a fixed character position. Supply a
named integer vector where the name becomes the column name and the integer is
the field width. Unnamed entries are dropped silently:

```{r}
# Split the 4-digit year into century (first 2 chars) and year (last 2 chars)
table3 |>
  separate_wider_position(year, widths = c(century = 2, year = 2))
```

Combine both to split `rate` by delimiter and `year` by position in one
pipeline:

```{r}
table3 |>
  separate_wider_delim(rate, delim = "/", names = c("cases", "population")) |>
  mutate(cases = as.integer(cases), population = as.integer(population)) |>
  separate_wider_position(year, widths = c(century = 2, year = 2))
```


### Uniting Columns

`unite()` is the inverse --- it combines multiple columns into one:

```{r}
# Rebuild the year column from century + year
table3 |>
  separate(rate, into = c("cases", "population"), convert = TRUE) |>
  separate(year, into = c("century", "year"), sep = 2) |>
  unite(full_year, century, year, sep = "")
```

## Missing Values: Explicit vs. Implicit

A value can be missing in two distinct ways:

- **Explicit:** the cell exists and contains `NA`
- **Implicit:** the row simply does not exist in the data

```{r}
stocks <- tibble(
  year   = c(2015, 2015, 2015, 2015, 2016, 2016, 2016),
  qtr    = c(   1,    2,    3,    4,    2,    3,    4),
  return = c(1.88, 0.59, 0.35,   NA, 0.92, 0.17, 2.66)
)
stocks
```

The return for Q4 2015 is explicitly missing (`NA`). The return for Q1 2016 is
implicitly missing --- the row does not exist at all.

Pivoting can make implicit missings explicit:

```{r}
stocks |>
  pivot_wider(names_from = year, values_from = return)
```

Use `complete()` to make implicit missing values explicit while keeping long
format:

```{r}
stocks |>
  complete(year, qtr)
```

Use `fill()` when missing values mean "same as the previous row" (last
observation carried forward):

```{r}
treatment <- tribble(
  ~person,            ~treatment, ~response,
  "Derrick Whitmore", 1,          7,
  NA,                 2,          10,
  NA,                 3,          9,
  "Katherine Burke",  1,          4
)

treatment |> fill(person)
```

---

# Joining Data

## Why Join?

It is rare that all the information you need lives in a single data frame.
Relational data --- data spread across multiple linked tables --- is the norm in
practice. Joins let you combine them.

We will use the `nycflights13` package, which has five related tables:

```{r}
library(nycflights13)
```

- `flights`: every flight departing NYC in 2013
- `airlines`: carrier codes and full names
- `airports`: airport codes and metadata
- `planes`: plane metadata by tail number
- `weather`: hourly weather at each NYC airport

<!-- FIGURE: Relationship diagram showing how the five nycflights13 tables
     connect to each other via shared key variables -->

## Keys

A **primary key** uniquely identifies each row in a table. A **foreign key** is
the matching variable in another table that refers back to the primary key.

```{r}
airlines  # primary key: carrier
airports  # primary key: faa
planes    # primary key: tailnum
```

`weather` has a **compound key** --- you need both `origin` and `time_hour`
together to uniquely identify a row:

```{r}
weather
```

Always verify that a claimed primary key actually uniquely identifies rows:

```{r}
# Confirm tailnum is unique in planes
planes |>
  count(tailnum) |>
  filter(n > 1)

# Confirm the compound key in weather
weather |>
  count(origin, time_hour) |>
  filter(n > 1)
```

Empty results confirm both keys are valid.

## Mutating Joins

A **mutating join** adds columns from one table to another by matching on a key.
We will work with a slim version of `flights`:

```{r}
flights2 <- flights |>
  select(year, time_hour, origin, dest, tailnum, carrier)
flights2
```

### `left_join()` — the join you will use 90% of the time

`left_join(x, y)` keeps every row in `x` and fills in matching columns from `y`.
Rows in `x` with no match in `y` get `NA` for the new columns.

Add full carrier names:

```{r}
flights2 |>
  left_join(airlines)
```

Add weather conditions at departure:

```{r}
flights2 |>
  left_join(weather |> select(origin, time_hour, temp, wind_speed))
```

Add plane size:

```{r}
flights2 |>
  left_join(planes |> select(tailnum, type, engines, seats))
```

::: {.callout-important}
**Why `left_join()` almost always?** Because it preserves every row in your
primary dataset. You will not silently lose observations the way `inner_join()`
can. If a join key is missing from `y`, you get `NA` --- which is visible and
diagnosable. If `inner_join()` drops rows, you may not notice.
:::

### Specifying the Join Key

By default, `left_join()` joins on all columns with matching names. This is
called a **natural join** and works well when column names are consistent across
tables. When they are not, use `by`:

```{r}
# both tables have a "year" column but mean different things
# only join on tailnum
flights2 |>
  left_join(planes, by = "tailnum")
```

When the key has different names in each table, use `by = c("x_name" = "y_name")`:

```{r}
# join flights destination to airport metadata
flights2 |>
  left_join(airports, by = c("dest" = "faa"))

# join flights origin to airport metadata
flights2 |>
  left_join(airports, by = c("origin" = "faa"))
```

### Other Mutating Joins

| Join | Keeps |
|---|---|
| `left_join(x, y)` | All rows of `x`; fills NAs where no match in `y` |
| `right_join(x, y)` | All rows of `y`; fills NAs where no match in `x` |
| `inner_join(x, y)` | Only rows with matches in **both** `x` and `y` |
| `full_join(x, y)` | All rows from both; NAs where no match on either side |

In practice: use `left_join()` as your default. Use `inner_join()` only when
you explicitly want to drop non-matching rows and have verified that is safe.

## Filtering Joins

Filtering joins do not add columns --- they use `y` to decide which rows of `x`
to keep or drop.

### `semi_join()` — keep rows in `x` that have a match in `y`

```{r}
# Keep only airports that appear as origins in flights
airports |>
  semi_join(flights2, by = c("faa" = "origin"))
```

### `anti_join()` — keep rows in `x` that have NO match in `y`

`anti_join()` is the most useful diagnostic tool for join problems. It finds
rows that would be silently dropped by an inner join.

```{r}
# Which destination codes in flights have no matching airport entry?
flights2 |>
  anti_join(airports, by = c("dest" = "faa")) |>
  distinct(dest)
```

```{r}
# Which tail numbers in flights have no plane record?
flights2 |>
  anti_join(planes, by = "tailnum") |>
  distinct(tailnum) |>
  nrow()
```

::: {.callout-tip}
**Always run `anti_join()` before a major join.** If it returns rows you did not
expect to drop, you have a data quality problem to investigate before proceeding
with the analysis.
:::

---

# Exercises

**Exercise 1.** Read the `framingham.csv` file from the `data/` folder using
`read_csv()`. Check `str()` and `summary()`. How many variables were read in as
the wrong type? Use `mutate()` to correct at least three of them.

**Exercise 2.** The `table4a` dataset has years as column names. Use
`pivot_longer()` to convert it to tidy format with columns `country`, `year`,
and `cases`. Then join it with the tidy version of `table4b` (population data)
using `left_join()` to create a single tidy table with `country`, `year`,
`cases`, and `population`.

**Exercise 3.** Using `flights` and `airlines`, add the full airline name to the
flights dataset. Then compute the mean arrival delay by airline name (not code)
and sort from worst to best.

**Exercise 4.** Add weather data (temperature and wind speed) to `flights2`
using `left_join()`. Then: how many flights are missing weather data? Use
`anti_join()` to identify which `origin`-`time_hour` combinations in `flights2`
have no match in `weather`.

**Exercise 5 (Challenge).** Start with `flights`. Join in plane information
(`planes`) and airport names for both `origin` and `dest` (requires two separate
joins to `airports`, using different `by` arguments and different suffixes).
Then compute, for each origin airport name and aircraft engine type, the mean
air time. Keep only groups with at least 500 flights and sort by mean air time
descending.
