---
title: "Day 3 (PM) Lab: Importing, Tidying, and Joining Data"
subtitle: "SC INBRE Biostatistics Summer Intensive 2026"
date: today
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    number-sections: true
    smooth-scroll: true
    theme: cosmo
    code-fold: false
    highlight-style: github
execute:
  echo: true
  warning: false
  message: false
---

# Overview

This lab covers everything from Day 3 AM:

- **Importing:** reading CSV and Excel files, handling missing value codes,
  cleaning column names, writing output
- **Tidy data:** identifying untidy data, pivoting with `pivot_longer()` and
  `pivot_wider()`, separating and uniting columns
- **Joining:** keys, `left_join()`, `semi_join()`, `anti_join()`

**How this lab works:**

- **Part 1 (instructor demo, ~20 min):** Follow along as the instructor works
  through a short data import and join pipeline using the Framingham and
  `nycflights13` data.
- **Part 2 (on your own, ~90 min):** Six exercises of increasing difficulty.
  Create a new `.qmd` file, work through the exercises, and render when done.

The **Quick Reference Card** in Section 3 has the key functions.

---

# Part 1: Instructor Demo

## Setup

```{r}
library(tidyverse)
library(readxl)
library(nycflights13)
library(this.path)
setwd(this.dir())
```

## Reading a CSV with Non-Standard Missing Values

Real data files often encode missing values in non-standard ways --- `"N/A"`,
`"."`, `-99`, or a blank cell. Always declare these when reading in:

```{r}
#| eval: false
# tell read_csv() that both "N/A" and blank cells should be treated as NA
students <- read_csv("data/students.csv", na = c("N/A", ""))
students
```

After reading, clean column names with `janitor::clean_names()` and fix
variable types with `mutate()` --- the same workflow you used on Day 2:

```{r}
#| eval: false
library(janitor)
students <- students |>
  clean_names() |>
  mutate(meal_plan = factor(meal_plan))
```

## What Tidy Looks Like

These two tables hold the same data. Only one is tidy:

```{r}
table1  # tidy: one row per country-year, one column per variable
table2  # untidy: cases and population are stacked into one column
```

In `table1`, computing a rate per 10,000 is one line:

```{r}
table1 |> mutate(rate = cases / population * 10000)
```

In `table2`, you cannot do this directly --- you would have to separate cases
from population first, then rejoin them. That is the cost of untidy data.

## Pivoting: Wide to Long

```{r}
# table4a has years as column names --- year is a variable, not a label
table4a

table4a |>
  pivot_longer(
    cols      = c(`1999`, `2000`),
    names_to  = "year",
    values_to = "cases"
  )
```

## A Join: Adding Airline Names to Flights

```{r}
flights2 <- flights |>
  select(year, time_hour, origin, dest, tailnum, carrier)

# left_join keeps every row in flights2 and adds the name column from airlines
flights2 |>
  left_join(airlines, by = "carrier") |>
  select(carrier, name, origin, dest)
```

## An Anti-Join: Finding Data Quality Problems

```{r}
# Which tail numbers in flights have no matching record in planes?
flights2 |>
  anti_join(planes, by = "tailnum") |>
  distinct(tailnum) |>
  nrow()
```

Over 700 flights have tail numbers that don't appear in `planes`. These would
be silently dropped by an `inner_join()`. `anti_join()` makes the problem
visible before it causes silent data loss.

---

# Part 2: Exercises

## Instructions

**Step 1.** Open a new Quarto document: File → New File → Quarto Document.
Title it `"Day 3 Lab"`, select HTML, click **Create**.

**Step 2.** Replace the template content with this YAML and setup chunk:

````markdown
---
title: "Day 3 Lab: Importing, Tidying, and Joining"
subtitle: "SC INBRE Biostatistics Summer Intensive 2026"
date: today
format:
  html:
    toc: true
    theme: cosmo
execute:
  echo: true
  warning: false
  message: false
---

```{{r}}
library(tidyverse)
library(readxl)
library(nycflights13)
library(this.path)
this.dir <- this.dir()
setwd(this.dir)
```
````

**Step 3.** Work through Exercises 1--6 in order. For each one:

- Add a new code chunk (`Ctrl+Alt+I` / `Cmd+Option+I`)
- Write and run your code
- Add 1--2 sentences below the chunk describing what you found

**Step 4.** When finished, click **Render** (`Ctrl+Shift+K` / `Cmd+Shift+K`)
and fix any errors before you are done.

::: {.callout-tip}
Check the **Quick Reference Card** in Section 3 before asking for help. The
most common issues are: (1) forgetting backticks around numeric column names
like `` `1999` ``, (2) `by` argument mismatches in joins, and (3) `na.rm = TRUE`
missing from summary functions.
:::

::: {.callout-note}
Exercises 1--4 cover the core material. Exercise 5 is harder and requires
combining joins with `group_by()` / `summarize()`. Exercise 6 is a challenge
--- don't worry if you don't reach it.
:::

---

**Exercise 1 — Reading and Cleaning.**
Read `framingham.csv` from the `data/` folder using `read_csv()`. The dataset
uses blank cells for missing values, which `read_csv()` handles automatically.
After reading in:

a. Use `str()` to check column types. Which variables are stored as `<dbl>` but
   should be factors?
b. Convert at least four binary or categorical variables to factors with
   meaningful labels using `mutate()` and `factor()` (you did this on Day 2 ---
   copy that block if you still have it).
c. Write the cleaned dataset to a new file called `framingham_clean.csv` using
   `write_csv()`.

```{r}
# your code here
```

---

**Exercise 2 — Identifying Untidy Data.**
Look at `table2` and `table4a` (built into the tidyverse).

a. In plain English (1--2 sentences each), describe what makes `table2` untidy
   and what makes `table4a` untidy. What would a tidy version of each look like?

b. Tidy `table4a` using `pivot_longer()` so it has three columns: `country`,
   `year`, and `cases`. Note: the column names `1999` and `2000` are numbers,
   not strings --- you need to wrap them in backticks (`` `1999` ``).

c. Tidy `table4b` the same way, producing columns `country`, `year`, and
   `population`.

d. Join the two tidy tables together using `left_join()` to produce a single
   tidy table with `country`, `year`, `cases`, and `population`. Add a `rate`
   column (cases per 10,000 population).

```{r}
# your code here
```

---

**Exercise 3 — Pivoting Real Data.**
The `who` dataset (built into the tidyverse) is a World Health Organization
tuberculosis dataset with diagnosis counts spread across many columns:

```{r}
who
```

a. How many columns currently hold what is really the `count` variable (i.e.,
   how many columns need to be pivoted into rows)? Use `ncol(who)` and count
   the non-variable columns to figure this out.

b. Use `pivot_longer()` to gather all columns from `new_sp_m014` through
   `newrel_f65` into two columns: `key` (the old column name) and `count`
   (the value). Drop rows where `count` is `NA` using `values_drop_na = TRUE`.

c. How many rows does the resulting dataset have? How does this compare to the
   original?

d. The `key` column mashes together diagnosis type, sex, and age group. Use
   `mutate()` and `str_replace()` to fix the inconsistent `"newrel"` → `"new_rel"`
   prefix, then use `separate()` twice to split `key` into `new`, `type`, and
   `sexage`, and then `sexage` into `sex` and `age`. What does the final tidy
   dataset look like?

```{r}
# your code here
```

---

**Exercise 4 — Left Joins.**
Using `flights` and the other `nycflights13` tables:

a. Add the full airline name (from `airlines`) to `flights`. Then compute the
   mean departure delay by airline name. Which airline has the worst average
   departure delay?

b. Add the plane's number of seats (from `planes`) to `flights` using
   `left_join()` with `by = "tailnum"`. What proportion of flights are missing
   seat information (i.e., had no matching record in `planes`)?

c. Add the destination airport's full name and time zone (from `airports`) to
   `flights`. Which time zone has the most flight destinations from NYC?

```{r}
# your code here
```

---

**Exercise 5 — Joins and Grouped Summaries.**
This exercise combines joining with the `group_by()` / `summarize()` workflow
from Day 2.

a. Join `flights` with `weather` (using `origin` and `time_hour` as the
   compound key) to add temperature (`temp`) and visibility (`visib`) to each
   flight.

b. Compute the mean arrival delay by visibility level. Round visibility to the
   nearest integer first using `mutate(visib = round(visib))`. Sort from worst
   to best delay. Is there a relationship between visibility and delays?

c. Repeat (b) but compute separately for each origin airport. Use
   `group_by(origin, visib)`. Describe in 1--2 sentences what you find.

```{r}
# your code here
```

---

**Exercise 6 (Challenge) — Building a Spatial Dataset.**

a. Create a dataset called `flight_avg_delay` by:

   - Grouping `flights` by destination (`dest`)
   - Computing the mean arrival delay (`avg_delay`) for each destination,
     ignoring NAs
   - Joining in the latitude (`lat`), longitude (`lon`), and full airport name
     (`name`) from `airports` using `left_join()` with `by = c("dest" = "faa")`
   - Dropping destinations with no matching airport record using `filter()`
     and `!is.na(lat)`

b. How many destinations have no match in `airports`? Use `anti_join()` to
   find them before building the dataset.

c. Once you have `flight_avg_delay`, run this code to plot average arrival delay
   on a US map:

```{r}
#| eval: false

# This requires the maps package.
install.packages("maps") 
library(maps)

flight_avg_delay |>
  ggplot(aes(lon, lat, color = avg_delay)) +
  borders("state") +
  geom_point(size = 2) +
  coord_quickmap() +
  scale_color_gradient2(low = "blue", mid = "white", high = "red",
                        midpoint = 0) +
  labs(title = "Average Arrival Delay by Destination Airport",
       color  = "Avg delay (min)")
```

Which regions of the country have the worst average arrival delays?

```{r}
# your code here
```

---

# Quick Reference Card

## Reading Data

| Function | What it does |
|---|---|
| `read_csv("file.csv")` | Read CSV; returns a tibble |
| `read_csv("file.csv", na = c("N/A", ""))` | Declare missing value codes |
| `read_csv("file.csv", skip = n)` | Skip first `n` rows |
| `read_excel("file.xlsx", sheet = "name")` | Read specific Excel sheet |
| `write_csv(df, "output.csv")` | Write tibble to CSV |
| `clean_names(df)` | Convert column names to snake_case (`janitor`) |

## Tidy Data

| Function | What it does |
|---|---|
| `pivot_longer(cols, names_to, values_to)` | Wide → long |
| `pivot_wider(names_from, values_from)` | Long → wide |
| `separate(col, into, sep)` | Split one column into multiple |
| `unite(new_col, col1, col2, sep)` | Merge multiple columns into one |
| `complete(var1, var2)` | Make implicit NAs explicit |
| `fill(col)` | Carry last non-NA value forward |
| `values_drop_na = TRUE` | Drop NA rows inside `pivot_longer()` |

## Joins

| Function | Keeps |
|---|---|
| `left_join(x, y, by)` | All rows of `x`; NAs where no match in `y` |
| `inner_join(x, y, by)` | Only rows with matches in both |
| `full_join(x, y, by)` | All rows from both; NAs where no match |
| `semi_join(x, y, by)` | Rows of `x` that have a match in `y` |
| `anti_join(x, y, by)` | Rows of `x` that have NO match in `y` |

**Specifying keys:**

```r
# Same name in both tables
left_join(x, y, by = "key")

# Different names in each table
left_join(x, y, by = c("x_col" = "y_col"))

# Compound key (multiple columns)
left_join(x, y, by = c("col1", "col2"))
```

## Useful Reminders

- Numeric column names need backticks: `` `1999` ``, `` `2000` ``
- Check for join problems with `anti_join()` **before** your main join
- `na.rm = TRUE` is needed in `mean()`, `sd()`, `sum()` etc. whenever NAs are present
- After `pivot_longer()`, always check the result with `str()` and `head()`
