---
title: "Course Datasets"
subtitle: "SC INBRE Biostatistics Summer Intensive 2026"
date: today
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    number-sections: true
    smooth-scroll: true
    theme: cosmo
    code-fold: false
    highlight-style: github
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| echo: false
library(tidyverse)
library(this.path)
setwd(this.dir())
```

# Overview

This document describes the datasets used throughout the course. Each entry
includes a description, the relevant course days, a load test, exploratory
checks, and notes on any cleaning steps needed before analysis.

All datasets are either built into R (no file needed), available through an R
package, or stored as `.csv` files in the `data/` folder of the course
repository.

---

# Framingham Heart Study

## Source

The Framingham Heart Study is a landmark longitudinal cohort study of
cardiovascular disease risk factors, initiated in 1948 in Framingham,
Massachusetts. The teaching dataset used here is a publicly available
subset widely used in statistics education.

**File:** `data/framingham.csv` (in the course GitHub repository)

**Relevant course days:** Days 2--10 (used as the primary dataset
throughout Weeks 1 and 2)

## Description

The dataset contains demographic and cardiovascular risk factor information
on 4,240 participants. The primary outcome of interest is the 10-year risk
of coronary heart disease (`TenYearCHD`).

| Variable | Type | Description |
|---|---|---|
| `male` | Binary | Sex (0 = Female, 1 = Male) |
| `age` | Continuous | Age in years |
| `education` | Categorical | 1 = Some HS, 2 = HS Diploma, 3 = Some College, 4 = College Degree |
| `currentSmoker` | Binary | Current smoker (0 = No, 1 = Yes) |
| `cigsPerDay` | Continuous | Cigarettes smoked per day |
| `BPMeds` | Binary | On blood pressure medication (0 = No, 1 = Yes) |
| `prevalentStroke` | Binary | Prior stroke (0 = No, 1 = Yes) |
| `prevalentHyp` | Binary | Prevalent hypertension (0 = No, 1 = Yes) |
| `diabetes` | Binary | Diabetes status (0 = No, 1 = Yes) |
| `totChol` | Continuous | Total cholesterol (mg/dL) |
| `sysBP` | Continuous | Systolic blood pressure (mmHg) |
| `diaBP` | Continuous | Diastolic blood pressure (mmHg) |
| `BMI` | Continuous | Body mass index (kg/m²) |
| `heartRate` | Continuous | Heart rate (beats per minute) |
| `glucose` | Continuous | Blood glucose level (mg/dL) |
| `TenYearCHD` | Binary | 10-year coronary heart disease event (0 = No, 1 = Yes) |

## Load

```{r}
framingham <- read_csv("framingham.csv") |>
  mutate(
    male          = factor(male,          levels = c(0,1), labels = c("Female","Male")),
    currentSmoker = factor(currentSmoker, levels = c(0,1), labels = c("Non-smoker","Smoker")),
    BPMeds        = factor(BPMeds,        levels = c(0,1), labels = c("No","Yes")),
    prevalentStroke = factor(prevalentStroke, levels = c(0,1), labels = c("No","Yes")),
    prevalentHyp  = factor(prevalentHyp,  levels = c(0,1), labels = c("No","Yes")),
    diabetes      = factor(diabetes,      levels = c(0,1), labels = c("No","Yes")),
    TenYearCHD    = factor(TenYearCHD,    levels = c(0,1), labels = c("No","Yes")),
    education     = factor(education,     levels = 1:4,
                           labels = c("Some HS","HS Diploma","Some College","College Degree"))
  )

glimpse(framingham)
```

## Exploratory Checks

```{r}
summary(framingham)
```

```{r}
# Missing values per column
colSums(is.na(framingham))
```

```{r}
# Missing as a percentage
round(colSums(is.na(framingham)) / nrow(framingham) * 100, 1)
```

## Cleaning Notes

- The binary and categorical variables are stored as integers in the raw file
  and must be converted to factors using `mutate()` before analysis. The load
  code above does this automatically.
- Missing values are present in seven variables: `education`, `cigsPerDay`,
  `BPMeds`, `totChol`, `BMI`, `heartRate`, and `glucose`. Always use
  `na.rm = TRUE` in summary functions and be aware of how your modeling
  function handles missing data (most will drop rows with any `NA` by default).
- After applying the `filter(!is.na(...))` approach used in the Day 7 lab,
  the complete-case sample is approximately 3,658 participants.

---

# Airway (RNA-Seq)

## Source

Bioconductor package: `airway`

Himes et al. (2014). *RNA-Seq Transcriptome Profiling Identifies CRISPLD2 as
a Glucocorticoid Responsive Gene that Modulates Cytokine Function in Airway
Smooth Muscle Cells.* PLOS ONE.

**Relevant course days:** Days 11--14 (genomics, differential expression,
multiple testing, dimensionality reduction)

## Description

This dataset contains read counts from an RNA-sequencing experiment on human
airway smooth muscle cell lines. Four cell lines were treated with
dexamethasone (a corticosteroid) and four were untreated controls.

| Component | Description |
|---|---|
| Dimensions | 63,677 genes × 8 samples |
| Samples | 4 treated (dexamethasone) + 4 untreated controls |
| Format | `RangedSummarizedExperiment` object |
| Count matrix | Access with `assay(airway)` |
| Sample metadata | Access with `colData(airway)` |
| Gene metadata | Access with `rowData(airway)` |

## Install and Load

```{r}
#| eval: false
# Install Bioconductor and the airway package (run once)
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("airway")
```

```{r}
library(airway)
data(airway)
airway
```

## Exploratory Checks

```{r}
# Dimensions: genes x samples
dim(airway)

# Sample metadata
colData(airway)
```

```{r}
# First few rows of the count matrix
airway_counts <- assay(airway)
head(airway_counts)
```

```{r}
# Missing values (should be none in count data)
sum(is.na(airway_counts))
```

```{r}
# Distribution of counts for the first sample
summary(airway_counts[, 1])
```

## Cleaning Notes

- `assay(airway)` extracts the raw integer count matrix. Sample metadata
  (`colData`) and gene metadata (`rowData`) must be extracted separately if
  needed as plain data frames.
- Many genes have zero counts across all samples. DESeq2 handles this
  automatically through its filtering steps and does not require manual
  removal of zero-count genes.
- The `dex` column in `colData(airway)` records treatment status: `"trt"`
  (dexamethasone treated) vs. `"untrt"` (control). This is the grouping
  variable used in the Day 12 differential expression analysis.

---

# InsectSprays

## Source

Built-in base R dataset (`datasets` package --- no install required)

**Relevant course days:** Day 8 (GLMs --- Poisson regression for count
outcomes)

## Description

This dataset records the effectiveness of six different insecticide sprays
(A through F) applied to agricultural experimental units. It is a classic
example for count data analysis using Poisson or negative binomial
regression.

| Variable | Type | Description |
|---|---|---|
| `count` | Integer | Number of insects in each experimental unit |
| `spray` | Factor | Insecticide type (A, B, C, D, E, or F) |

72 observations total (12 per spray type).

## Load

```{r}
data(InsectSprays)
head(InsectSprays)
```

## Exploratory Checks

```{r}
str(InsectSprays)
summary(InsectSprays)
```

```{r}
# Counts per spray type
InsectSprays |>
  group_by(spray) |>
  summarize(
    n        = n(),
    mean     = mean(count),
    sd       = sd(count),
    min      = min(count),
    max      = max(count)
  )
```

```{r}
# Visualize
ggplot(InsectSprays, aes(x = spray, y = count, fill = spray)) +
  geom_boxplot(alpha = 0.7, show.legend = FALSE) +
  labs(x = "Insecticide", y = "Insect count",
       title = "InsectSprays: Count by Insecticide Type") +
  theme_minimal()
```

```{r}
# Missing values
colSums(is.na(InsectSprays))
```

## Cleaning Notes

No missing values. No factor conversion needed --- `spray` is already
stored as a factor with six levels (A through F). Spray type C, D, and E
appear to be substantially more effective (lower counts) than A, B, and F,
making this a clear case for demonstrating group comparisons with count
outcomes.

---

# sleepstudy (Longitudinal)

## Source

Package: `lme4`

Belenky et al. (2003). *Patterns of performance degradation and restoration
during sleep restriction and subsequent recovery: A sleep dose-response
study.* Journal of Sleep Research.

**Relevant course days:** Day 10 (mixed models for repeated/clustered data)

## Description

This dataset contains average reaction times from a psychomotor vigilance
test recorded daily from 18 subjects over 10 days in a sleep deprivation
study. On Day 0, subjects had a normal night of sleep. Beginning that night,
they were restricted to 3 hours of sleep per night for 9 consecutive nights.

| Variable | Type | Description |
|---|---|---|
| `Reaction` | Continuous | Average reaction time (milliseconds) |
| `Days` | Integer | Number of days of sleep deprivation (0--9) |
| `Subject` | Factor | Subject identifier (18 subjects) |

180 observations total (18 subjects × 10 days).

## Load

```{r}
library(lme4)
data("sleepstudy")
head(sleepstudy, 12)
```

## Exploratory Checks

```{r}
str(sleepstudy)
summary(sleepstudy)
```

```{r}
# Reaction time trajectories per subject
ggplot(sleepstudy, aes(x = Days, y = Reaction, group = Subject,
                       color = Subject)) +
  geom_line(alpha = 0.7, show.legend = FALSE) +
  geom_point(alpha = 0.5, show.legend = FALSE) +
  labs(x = "Days of sleep deprivation",
       y = "Reaction time (ms)",
       title = "Reaction Time over Sleep Deprivation by Subject") +
  theme_minimal()
```

```{r}
# Missing values
colSums(is.na(sleepstudy))
```

## Cleaning Notes

No missing values. `Subject` is already stored as a factor. The key feature
of this dataset for Day 10 is that each subject has a different baseline
reaction time (random intercept) and a different rate of decline with sleep
deprivation (random slope) --- making it an ideal demonstration of the
random intercept-and-slope mixed model.

---

# lung (Survival Analysis)

## Source

Package: `survival`

Data from the North Central Cancer Treatment Group (NCCTG). Originally
published in Loprinzi et al. (1994). *Prospective evaluation of prognostic
variables from patient-completed questionnaires.* Journal of Clinical
Oncology.

**Relevant course days:** Day 9 (survival analysis --- Kaplan-Meier,
log-rank test, Cox proportional hazards)

## Description

This dataset contains clinical and performance data from 228 patients with
advanced-stage lung cancer. The primary outcome is time to death, with some
patients censored (lost to follow-up or alive at end of study).

| Variable | Type | Description |
|---|---|---|
| `inst` | Integer | Institution code |
| `time` | Integer | Survival time (days) |
| `status` | Binary | Censoring status (1 = censored, 2 = dead) |
| `age` | Integer | Age in years |
| `sex` | Binary | Sex (1 = Male, 2 = Female) |
| `ph.ecog` | Integer | ECOG performance score (0 = good, 4 = bedridden) |
| `ph.karno` | Integer | Karnofsky score rated by physician (0--100) |
| `pat.karno` | Integer | Karnofsky score rated by patient (0--100) |
| `meal.cal` | Integer | Calories consumed at meals |
| `wt.loss` | Integer | Weight loss in past 6 months (lbs) |

::: {.callout-note}
In the `lung` dataset, `status = 1` means **censored** and `status = 2`
means **dead**. This is the opposite of the convention used in many other
survival datasets, where 1 = event. The `survival` package handles this
correctly when you use `Surv(time, status == 2)`.
:::

## Load

```{r}
library(survival)
head(lung)
```

## Exploratory Checks

```{r}
str(lung)
summary(lung)
```

```{r}
# Missing values per column
colSums(is.na(lung))
round(colSums(is.na(lung)) / nrow(lung) * 100, 1)
```

```{r}
# Event rate
table(lung$status)
prop.table(table(lung$status))
```

```{r}
# Quick Kaplan-Meier curve
km_fit <- survfit(Surv(time, status == 2) ~ sex, data = lung)
plot(km_fit, col = c("steelblue", "firebrick"),
     xlab = "Days", ylab = "Survival probability",
     main = "Kaplan-Meier: Survival by Sex")
legend("topright", legend = c("Male", "Female"),
       col = c("steelblue", "firebrick"), lty = 1)
```

## Cleaning Notes

- Missing values are present in `inst`, `ph.ecog`, `ph.karno`, `pat.karno`,
  `meal.cal`, and `wt.loss`. These are handled automatically by `survfit()`
  and `coxph()`, which use complete cases for each model.
- `sex` is coded 1 = Male and 2 = Female --- convert to a factor with
  informative labels for plots: `mutate(sex = factor(sex, levels = c(1,2), labels = c("Male","Female")))`.
- The `status` coding (1 = censored, 2 = dead) is non-standard. Always use
  `Surv(time, status == 2)` to define the survival object correctly.

---

# Summary Table

| Dataset | Source | Rows | Columns | Primary Use | Days |
|---|---|---|---|---|---|
| `framingham` | CSV file | 4,240 | 16 | Linear/logistic regression, EDA | 2--10 |
| `airway` | Bioconductor | 63,677 genes × 8 samples | --- | RNA-seq, DESeq2, PCA | 11--14 |
| `InsectSprays` | Base R | 72 | 2 | Poisson regression, count data | 8 |
| `sleepstudy` | `lme4` | 180 | 3 | Mixed models, longitudinal data | 10 |
| `lung` | `survival` | 228 | 10 | Survival analysis, KM, Cox PH | 9 |
