---
title: "01 - Introduction to R"
subtitle: "Biostatistics 735 — Machine Learning in Public Health"
author: "Alexander McLain"
date: today
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    number-sections: true
    smooth-scroll: true
    theme: cosmo
    code-fold: false
    highlight-style: github
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| include: false
library(tidyverse)
library(nycflights13)
library(janitor)
library(ggrepel)
library(haven)
library(this.path)
setwd(this.dir())

framingham <- read_csv("framingham.csv") |>
  mutate(
    male          = factor(male,          levels = c(0,1), labels = c("Female","Male")),
    currentSmoker = factor(currentSmoker, levels = c(0,1), labels = c("Non-smoker","Smoker")),
    prevalentHyp  = factor(prevalentHyp,  levels = c(0,1), labels = c("No","Yes")),
    diabetes      = factor(diabetes,      levels = c(0,1), labels = c("No","Yes")),
    TenYearCHD    = factor(TenYearCHD,    levels = c(0,1), labels = c("No","Yes")),
    education     = factor(education,     levels = 1:4,
                           labels = c("Some HS","HS Diploma","Some College","College Degree"))
  )
```

# Overview

By the end of this document you will be able to:

- Install and load R packages
- Set the working directory directly and with `this.path`
- Read CSV, Excel, and other common file formats into R
- Recognize the main variable types in R and convert between them
- Manipulate and transform data using core `dplyr` verbs
- Combine data frames using mutating and filtering joins
- Build exploratory and publication-ready plots with `ggplot2`
- Create and render a Quarto document

---

# Packages

Base R is powerful, but most of what we will do in this course uses **packages**
— collections of functions contributed by the R community. There are over
20,000 available.

**Install once** (downloads to your machine):

```{r}
#| eval: false
install.packages("tidyverse")
```

**Load at the start of every session** (makes functions available):

```{r}
library(tidyverse)
```

::: {.callout-important}
`install.packages()` and `library()` do different things. Installing downloads
the package; loading activates it. You only install once, but you load every
session.

Also: use straight quotes (`"tidyverse"`), not curly/smart quotes. Smart quotes
will cause an error.
:::

`tidyverse` loads several packages at once:

| Package | Purpose |
|---------|---------|
| `ggplot2` | Data visualization |
| `dplyr` | Data manipulation |
| `tidyr` | Reshaping data |
| `readr` | Reading/writing files |
| `stringr` | Working with text |
| `forcats` | Working with categorical variables |

We will use all of these throughout the course. Additional packages used in this
document:

```{r}
#| eval: false
install.packages(c("nycflights13", "janitor", "ggrepel", "haven", "this.path"))
```

---

# Working Directory

## Setting It Directly

R looks for files relative to the **working directory**. Check yours with `getwd()`:

```{r}
getwd()
```

You can set it manually with `setwd()`:

```{r}
#| eval: false
setwd("/Users/yourname/Documents/my-project")
```

Hard-coded paths like this break on other people's machines, so they should be
avoided whenever possible.

## Using `this.path`

Rather than hard-coding a full file path, use `this.path` to set the working
directory to wherever the script lives. Note, `rstudioapi` only works inside of
RStudio and can give error messages when used with `.qmd` files (and in other
situations). `this.path` is usually safer.

```{r}
#| eval: false
library(rstudioapi)
this.dir <- dirname(rstudioapi::getSourceEditorContext()$path)
setwd(this.dir)

library(this.path)
this.dir <- this.dir()
setwd(this.dir)
```

Then reference files with short relative paths: `"data/myfile.csv"` rather than
`"/Users/yourname/Documents/.../myfile.csv"`.

You can use `"../data/framingham.csv"` to tell R to go back one folder, find the
data folder, and then look for `framingham.csv`.

::: {.callout-tip}
**Windows users:** File paths use `\` in Windows Explorer, but R requires `/` or
`\\`. Copy a path from Explorer and replace each `\` with `/` before using it in R.
:::

---

# Reading in Data

## Key Terms

Before reading data into R, three concepts matter:

- **Header:** The first row of a file is usually variable names. R needs to know
  whether to treat it as names or as data values.
- **Delimiter:** The character that separates entries in each row (comma in `.csv`,
  tab in `.tsv`, etc.). R uses this to know where one value ends and the next begins.
- **Missing value:** How your file represents missingness. R recognizes `NA`, but
  files often use `.`, `""`, `N/A`, `999`, or other codes that need to be declared.

## Reading CSV Files with `readr`

`readr` is the tidyverse package for reading rectangular data. The most common
functions:

| Function | Reads |
|---|---|
| `read_csv()` | Comma-separated (`.csv`) |
| `read_csv2()` | Semicolon-separated (common in Europe) |
| `read_tsv()` | Tab-separated |
| `read_delim()` | Any delimiter you specify |

All have the same interface — once you know `read_csv()` the rest are trivial.

```{r}
#| eval: false
library(this.path)
setwd(this.dir())
framingham <- read_csv("framingham.csv")
```

When you run `read_csv()` it prints a message showing the number of rows and
columns, the delimiter used, and how each column was parsed. This is useful for
catching problems early.

### Common Arguments

**Skipping rows** — useful when files have metadata at the top:

```{r}
#| message: false
read_csv("The first line of metadata
  The second line of metadata
  x,y,z
  1,2,3", skip = 2)
```

**No column names** — tell R to generate them:

```{r}
#| message: false
read_csv("1,2,3
         4,5,6", col_names = FALSE)
```

**Declare missing value codes** — files often use `.` or `N/A` instead of `NA`:

```{r}
#| message: false
read_csv("a,b,c
         1,2,.", na = ".")
```

### Post-Import Cleaning

After reading data in, two immediate tasks often arise:

**1. Fix column names with spaces.** Column names with spaces require backtick
quoting (`` `Full Name` ``) which is error-prone. Use `janitor::clean_names()` to
convert everything to snake_case automatically:

```{r}
library(janitor)
x <- data.frame(caseID = 1, DOB = 2, Other = 3)
x |>
  clean_names()
```

Or rename specific columns with `rename()`:

```{r}
x |> rename(case_id = `caseID`, dob = DOB, other = Other)
```

**2. Fix variable types.** `read_csv()` guesses column types from the data. Always
check with `str()` and correct anything that was misread (e.g., a numeric column
read as character because of a stray word like `"five"`).

## Reading Excel Files

Use `readxl::read_excel()`. You can specify which sheet to read:

```{r}
#| eval: false
library(readxl)
data_xl <- read_excel("data/myfile.xlsx", sheet = "data")
dict_xl <- read_excel("data/myfile.xlsx", sheet = "dictionary")
```

The same arguments apply: `col_names`, `na`, `skip`.

## Writing Data

```{r}
#| eval: false
write_csv(framingham, "framingham_clean.csv")
```

::: {.callout-note}
**Important:** When you write a tibble to CSV and read it back, factor levels and
other type information are lost — the file only stores plain text. If you need to
preserve types across sessions, use `saveRDS()` / `readRDS()` to save as an R
binary file instead.
:::

## Other Formats

The `haven` package reads SAS (`.sas7bdat`), Stata (`.dta`), and SPSS (`.sav`)
files with the same interface:

```{r}
#| eval: false
library(haven)
spss_data  <- read_sav("data/myfile.sav")
sas_data   <- read_sas("data/myfile.sas7bdat")
stata_data <- read_dta("data/myfile.dta")
```

SPSS and Stata files often carry variable labels and value labels that `haven`
preserves as attributes — useful when you are handed someone else's data and
need to understand the coding.

::: {.callout-tip}
You can read in SAS data, you can even write SAS data (using `write_sas`), but SAS
data exported from R will not properly be read into SAS. Exported SPSS
(`write_sav`) and Stata (`write_dta`) files can be read into their respective
software programs.
:::

---

# Variable Types (Brief)

The type of a variable determines what operations you can do with it and how R
and statistical functions will treat it. Getting variable types right is one of
the most important — and most commonly skipped — steps in data preparation.

We will use the `flights` dataset from the `nycflights13` package throughout.

```{r}
library(nycflights13)
flights
```

The abbreviations under each column name tell you the type:

| Abbreviation | Type | Example |
|---|---|---|
| `<int>` | Integer | `year`, `month`, `day` |
| `<dbl>` | Double (numeric, decimal) | `dep_delay`, `air_time` |
| `<chr>` | Character (string) | `carrier`, `origin` |
| `<lgl>` | Logical (TRUE/FALSE) | e.g., `is_delayed` |
| `<fct>` | Factor (categorical with levels) | e.g., severity grades |
| `<dttm>` | Date-time | `time_hour` |

## Numeric Variables (`<dbl>`, `<int>`)

Numeric variables represent measured quantities. Arithmetic and summary functions
apply directly.

```{r}
mean(flights$dep_delay, na.rm = TRUE)
sd(flights$dep_delay, na.rm = TRUE)
range(flights$air_time, na.rm = TRUE)
```

::: {.callout-note}
`na.rm = TRUE` tells R to ignore missing values when computing the summary.
Without it, a single `NA` in the vector will return `NA` for the whole
calculation.
:::

## Character Variables (`<chr>`)

Character variables (strings) hold text. They cannot be used in arithmetic but
can be searched, split, and recoded.

```{r}
unique(flights$carrier)
table(flights$carrier)
```

## Logical Variables (`<lgl>`)

Logicals are `TRUE` or `FALSE`. They arise naturally from comparisons and are
useful for filtering and counting. `sum()` and `mean()` treat `TRUE` as 1 and
`FALSE` as 0, so you can count and compute proportions easily:

```{r}
sum(flights$arr_delay > 0, na.rm = TRUE)
mean(flights$arr_delay > 0, na.rm = TRUE) * 100
```

## Factors (`<fct>`)

Factors represent **categorical variables with a fixed set of levels**. They are
essential in statistical modeling — a factor in a regression model is handled
as a categorical predictor (with a reference level), while a character variable
or numeric code is not.

```{r}
severity <- factor(c("mild", "severe", "moderate", "mild", "severe"),
                   levels = c("mild", "moderate", "severe"))
severity
levels(severity)
table(severity)
```

::: {.callout-important}
**When does this matter for you?** If you code a categorical variable as a number
(e.g., 1 = mild, 2 = moderate, 3 = severe) and put it in a model without
converting it to a factor, R will treat it as a continuous variable and fit a
single slope. This is almost never what you want. Always check whether your
grouping variables are stored as factors. **This is one of the most common errors when running models in R.**
:::

The first level is the reference level. To change it use `relevel` (important for
model interpretation):

```{r}
severity_reref <- relevel(severity, ref = "moderate")
levels(severity_reref)
```

`fct_recode()` lets you rename levels; `fct_collapse()` merges them:

```{r}
severity_recoded <- fct_recode(severity,
  "Grade 1" = "mild",
  "Grade 2" = "moderate",
  "Grade 3" = "severe"
)
levels(severity_recoded)
```

```{r}
fct_collapse(severity,
  "low"  = "mild",
  "high" = c("moderate", "severe")
)
```

---

# Data Transformations with dplyr

`dplyr` is the workhorse package for data manipulation. Its functions are called
**verbs** because each one does one thing clearly. We will cover the ones you will
use in almost every analysis.

## The Pipe

The pipe (`|>`) passes the output of one function as the first argument of the
next. Read it as "then":

```{r}
#| eval: false
flights |>           # start with flights, then
  filter(...) |>     # keep certain rows, then
  mutate(...) |>     # create new columns, then
  summarize(...)     # collapse to a summary
```

You may also see `%>%` from the `magrittr` package — it behaves identically
for our purposes.

## Filtering Rows: `filter()`

Keep rows that satisfy a condition:

```{r}
flights |>
  filter(arr_delay > 120)

flights |>
  filter(month == 1 &
           carrier %in% c("UA", "AA")
         )
```

Common operators: `>`, `>=`, `<`, `<=`, `==`, `!=`, `%in%`, `is.na()`.

::: {.callout-note}
`=` assigns a value; `==` tests for equality. Using `=` inside `filter()` is
one of the most common beginner errors in R.
:::

## Selecting and Renaming Columns: `select()` and `rename()`

```{r}
flights |>
  select(year, month, day, carrier, dep_delay, arr_delay)

flights |>
  rename(tail_number = tailnum) |>
  select(tail_number, carrier)
```

Useful helpers inside `select()`: `starts_with()`, `ends_with()`,
`contains()`, `where(is.numeric)`.

```{r}
flights |>
  select(contains("time"))
```

## Creating and Modifying Columns: `mutate()`

`mutate()` adds new columns or overwrites existing ones:

```{r}
flights |>
  mutate(
    gain       = dep_delay - arr_delay,
    speed_mph  = distance / air_time * 60,
    late       = arr_delay > 0
  ) |>
  select(carrier, gain, speed_mph, late)
```

`mutate()` is also where you will do type conversions:

```{r}
flights |>
  mutate(
    month_fct = factor(month,
                       levels = 1:12,
                       labels = month.abb)
  ) |>
  count(month_fct)
```

## Recoding Variables: `if_else()` and `case_when()`

`if_else()` handles a simple two-way recode:

```{r}
flights |>
  mutate(long_flight = if_else(air_time > 300, "Long", "Short")) |>
  count(long_flight)
```

`case_when()` handles three or more conditions. Each condition is written as
`condition ~ value`, and `.default` catches anything not matched:

```{r}
flights |>
  mutate(
    delay_category = case_when(
      dep_delay <= 0   ~ "On time or early",
      dep_delay <= 30  ~ "Minor delay (<= 30 min)",
      dep_delay <= 120 ~ "Moderate delay (31-120 min)",
      .default         = "Major delay (> 120 min)"
    )
  ) |>
  count(delay_category)
```

Conditions are evaluated top to bottom — the first match wins.

## Sorting: `arrange()`

```{r}
flights |>
  arrange(desc(dep_delay)) |>
  select(carrier, origin, dest, dep_delay)
```

## Grouped Summaries: `group_by()` + `summarize()`

This combination is how you compute group-level statistics:

```{r}
flights |>
  group_by(carrier) |>
  summarize(
    mean_delay = mean(arr_delay, na.rm = TRUE),
    n_flights  = n()
  ) |>
  arrange(desc(mean_delay))
```

::: {.callout-note}
Always include `na.rm = TRUE` inside summary functions like `mean()`, `sd()`,
and `median()`. Missing values in the input will return `NA` for the whole
summary otherwise.
:::

Multiple grouping variables:

```{r}
flights |>
  group_by(origin, carrier) |>
  summarize(
    mean_delay = mean(arr_delay, na.rm = TRUE),
    n          = n(),
    .groups    = "drop"
  )
```

## Putting It Together: A Pipeline

Compute the on-time arrival rate by carrier for flights longer than 500 miles,
keeping only carriers with at least 1,000 such flights:

```{r}
flights |>
  filter(distance > 500) |>
  mutate(on_time = arr_delay <= 0) |>
  group_by(carrier) |>
  summarize(
    n           = n(),
    pct_on_time = mean(on_time, na.rm = TRUE) * 100
  ) |>
  filter(n >= 1000) |>
  arrange(desc(pct_on_time))
```

---

# Joining Data

## Why Join?

It is rare that all the information you need lives in a single data frame.
Relational data — data spread across multiple linked tables — is the norm in
practice. Joins let you combine them.

We will use the `nycflights13` package, which has five related tables:

- `flights`: every flight departing NYC in 2013
- `airlines`: carrier codes and full names
- `airports`: airport codes and metadata
- `planes`: plane metadata by tail number
- `weather`: hourly weather at each NYC airport

## Keys

A **primary key** uniquely identifies each row in a table. A **foreign key** is
the matching variable in another table that refers back to the primary key.

```{r}
airlines  # primary key: carrier
airports  # primary key: faa
planes    # primary key: tailnum
```

Always verify that a claimed primary key actually uniquely identifies rows:

```{r}
planes |>
  count(tailnum) |>
  filter(n > 1)

weather |>
  count(origin, time_hour) |>
  filter(n > 1)
```

Empty results confirm both keys are valid.

## Mutating Joins

A **mutating join** adds columns from one table to another by matching on a key.

```{r}
flights2 <- flights |>
  select(year, time_hour, origin, dest, tailnum, carrier)
flights2
```

### `left_join()` — the join you will use 90% of the time

`left_join(x, y)` keeps every row in `x` and fills in matching columns from `y`.
Rows in `x` with no match in `y` get `NA` for the new columns.

```{r}
flights2 |>
  left_join(airlines)

flights2 |>
  left_join(weather |> select(origin, time_hour, temp, wind_speed))

flights2 |>
  left_join(planes |> select(tailnum, type, engines, seats))
```

::: {.callout-important}
**Why `left_join()` almost always?** Because it preserves every row in your
primary dataset. You will not silently lose observations the way `inner_join()`
can. If a join key is missing from `y`, you get `NA` — which is visible and
diagnosable. If `inner_join()` drops rows, you may not notice.
:::

### Specifying the Join Key

By default, `left_join()` joins on all columns with matching names. When key
names differ across tables, use `by`:

```{r}
flights2 |>
  left_join(airports, by = c("dest" = "faa"))

flights2 |>
  left_join(airports, by = c("origin" = "faa"))
```

### Other Mutating Joins

| Join | Keeps |
|---|---|
| `left_join(x, y)` | All rows of `x`; fills NAs where no match in `y` |
| `right_join(x, y)` | All rows of `y`; fills NAs where no match in `x` |
| `inner_join(x, y)` | Only rows with matches in **both** `x` and `y` |
| `full_join(x, y)` | All rows from both; NAs where no match on either side |

In practice: use `left_join()` as your default. Use `inner_join()` only when
you explicitly want to drop non-matching rows and have verified that is safe.

## Filtering Joins

Filtering joins do not add columns — they use `y` to decide which rows of `x`
to keep or drop.

### `semi_join()` — keep rows in `x` that have a match in `y`

```{r}
airports |>
  semi_join(flights2, by = c("faa" = "origin"))
```

### `anti_join()` — keep rows in `x` that have NO match in `y`

`anti_join()` is the most useful diagnostic tool for join problems. It finds
rows that would be silently dropped by an inner join.

```{r}
flights2 |>
  anti_join(airports, by = c("dest" = "faa")) |>
  distinct(dest)

flights2 |>
  anti_join(planes, by = "tailnum") |>
  distinct(tailnum) |>
  nrow()
```

::: {.callout-tip}
**Always run `anti_join()` before a major join.** If it returns rows you did not
expect to drop, you have a data quality problem to investigate before proceeding
with the analysis.
:::

---

# Data Visualization

We use the **Framingham Heart Study** dataset throughout this section.

```{r}
library(this.path)
setwd(this.dir())

framingham <- read_csv("framingham.csv") |>
  mutate(
    male          = factor(male,          levels = c(0,1), labels = c("Female","Male")),
    currentSmoker = factor(currentSmoker, levels = c(0,1), labels = c("Non-smoker","Smoker")),
    prevalentHyp  = factor(prevalentHyp,  levels = c(0,1), labels = c("No","Yes")),
    diabetes      = factor(diabetes,      levels = c(0,1), labels = c("No","Yes")),
    TenYearCHD    = factor(TenYearCHD,    levels = c(0,1), labels = c("No","Yes")),
    education     = factor(education,     levels = 1:4,
                           labels = c("Some HS","HS Diploma","Some College","College Degree"))
  )
str(framingham)
```

## The Grammar of Graphics

`ggplot2` is built on the idea that every plot can be described by the same
small set of components:

| Component | What it controls |
|---|---|
| **Data** | The tibble you are visualizing |
| **Aesthetics** (`aes()`) | How variables map to visual properties (x, y, color, size, shape) |
| **Geoms** | The geometric objects that represent the data (points, lines, bars, boxes) |
| **Scales** | How data values map to aesthetic values (axis limits, color palettes) |
| **Facets** | Splitting the plot into small multiples by a grouping variable |
| **Theme** | Non-data appearance (fonts, background, grid lines) |

Every `ggplot2` plot follows the same template:

```{r}
#| eval: false
ggplot(data = <DATA>) +
  <GEOM>(mapping = aes(<MAPPINGS>)) +
  <SCALES> +
  <FACETS> +
  <THEME>
```

## Visualizing Distributions

### Continuous Variables: Histograms

A histogram bins a continuous variable and counts observations per bin. Always
try several `binwidth` values — the pattern can change substantially.

```{r}
ggplot(framingham, aes(x = sysBP)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "white") +
  labs(x = "Systolic Blood Pressure (mmHg)", y = "Count",
       title = "Distribution of Systolic Blood Pressure")
```

### Continuous Variables: Density Plots

A density plot is a smoothed histogram. It is better than a histogram for
overlaying multiple groups because lines are easier to compare than bars.

```{r}
ggplot(framingham, aes(x = sysBP, fill = male)) +
  geom_density(alpha = 0.4) +
  labs(x = "Systolic Blood Pressure (mmHg)", y = "Density",
       fill = "Sex",
       title = "Systolic BP Distribution by Sex")
```

### Categorical Variables: Bar Charts

```{r}
ggplot(framingham, aes(x = education)) +
  geom_bar(fill = "steelblue") +
  labs(x = "Education Level", y = "Count",
       title = "Education Distribution in Framingham Cohort") +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
```

To plot proportions rather than counts, compute them first with `dplyr` and
use `geom_col()`:

```{r}
framingham |>
  count(education) |>
  mutate(pct = n / sum(n) * 100) |>
  ggplot(aes(x = education, y = pct)) +
  geom_col(fill = "steelblue") +
  labs(x = "Education Level", y = "Percent (%)",
       title = "Education Distribution (%)") +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
```

## Visualizing Covariation

### Continuous + Categorical: Boxplots

Boxplots show the median, IQR, and outliers compactly. They are ideal for
comparing a continuous variable across groups.

```{r}
ggplot(framingham, aes(x = male, y = sysBP)) +
  geom_boxplot(fill = "steelblue", alpha = 0.7) +
  labs(x = "Sex", y = "Systolic BP (mmHg)",
       title = "Systolic Blood Pressure by Sex")
```

When you have many groups, order them by the median to make comparisons easier:

```{r}
ggplot(framingham, aes(x = reorder(education, sysBP, FUN = median),
                       y = sysBP)) +
  geom_boxplot(fill = "steelblue", alpha = 0.7) +
  labs(x = "Education Level",
       y = "Systolic BP (mmHg)",
       title = "Systolic BP by Education Level",
       subtitle = "Groups ordered by median BP") +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
```

### Continuous + Categorical: Violin Plots

A violin plot shows the full distribution shape, not just the summary statistics.
Combine with a boxplot overlay for the best of both:

```{r}
ggplot(framingham, aes(x = male, y = sysBP)) +
  geom_violin(fill = "steelblue", alpha = 0.5) +
  geom_boxplot(width = 0.1, fill = "white", outlier.size = 0.8) +
  labs(x = "Sex", y = "Systolic BP (mmHg)",
       title = "Systolic BP by Sex: Violin + Boxplot")
```

### Two Continuous Variables: Scatterplots

```{r}
ggplot(framingham, aes(x = age, y = sysBP)) +
  geom_point(alpha = 0.3) +
  labs(x = "Age (years)", y = "Systolic BP (mmHg)",
       title = "Systolic BP vs. Age")
```

Add a smooth trend line with `geom_smooth()`:

```{r}
#| message: false
ggplot(framingham, aes(x = age, y = sysBP)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "loess", color = "firebrick", se = TRUE) +
  labs(x = "Age (years)", y = "Systolic BP (mmHg)",
       title = "Systolic BP vs. Age with Loess Smoother")
```

Map a third variable to color:

```{r}
#| message: false
ggplot(framingham, aes(x = age, y = sysBP, color = male)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "loess", se = FALSE) +
  labs(x = "Age (years)", y = "Systolic BP (mmHg)",
       color = "Sex",
       title = "Systolic BP vs. Age by Sex")
```

## Faceting

Faceting splits a plot into small multiples by a variable — often cleaner than
color coding when there are many groups.

```{r}
#| message: false
ggplot(framingham, aes(x = age, y = sysBP)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "loess", se = FALSE, color = "firebrick") +
  facet_wrap(~ male) +
  labs(x = "Age", y = "Systolic BP (mmHg)",
       title = "Systolic BP vs. Age, by Sex")
```

`facet_grid()` facets on two variables simultaneously:

```{r}
#| message: false
ggplot(framingham, aes(x = age, y = sysBP)) +
  geom_point(alpha = 0.2, size = 0.8) +
  geom_smooth(method = "loess", se = FALSE, color = "firebrick") +
  facet_grid(currentSmoker ~ male) +
  labs(x = "Age", y = "Systolic BP (mmHg)",
       title = "Systolic BP vs. Age by Sex and Smoking Status")
```

## Labels and Annotation

```{r}
framingham |>
  filter(!is.na(education)) |>
  ggplot(aes(x = reorder(education, sysBP, FUN = median),
                         y = sysBP, fill = male)) +
  geom_boxplot(alpha = 0.7) +
  labs(
    title    = "Systolic Blood Pressure by Education and Sex",
    subtitle = "Framingham Heart Study cohort",
    caption  = "Source: Framingham Heart Study",
    x        = "Education Level",
    y        = "Systolic BP (mmHg)",
    fill     = "Sex"
  ) +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
```

Add horizontal or vertical reference lines to mark thresholds:

```{r}
ggplot(framingham, aes(x = age, y = sysBP)) +
  geom_point(alpha = 0.2) +
  geom_hline(yintercept = 140, color = "blue",
             linewidth = 1, linetype = "dashed") +
  annotate("text", x = 35, y = 143,
           label = "Hypertension threshold (140 mmHg)",
           color = "blue", hjust = 0, size = 7) +
  labs(x = "Age (years)", y = "Systolic BP (mmHg)",
       title = "Systolic BP with Hypertension Threshold")
```

## Scales and Themes

Color scales: use ColorBrewer palettes for categorical variables and viridis
for continuous color, both of which are colorblind-friendly:

```{r}
ggplot(framingham, aes(x = age, y = sysBP, color = education)) +
  geom_point(alpha = 0.4) +
  scale_color_brewer(palette = "Set2") +
  labs(x = "Age", y = "Systolic BP", color = "Education")

ggplot(framingham, aes(x = age, y = sysBP, color = BMI)) +
  geom_point(alpha = 0.5) +
  scale_color_viridis_c(option = "plasma") +
  labs(x = "Age", y = "Systolic BP", color = "BMI")
```

Themes control all non-data elements: background, grid lines, fonts, and text size:

```{r}
#| message: false
base <- ggplot(framingham, aes(x = age, y = sysBP, color = male)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "loess", se = FALSE) +
  labs(x = "Age", y = "Systolic BP", color = "Sex")

base + theme_bw()
base + theme_minimal()
base + theme_classic()
```

## Saving Plots

Use `ggsave()` to write the most recent plot to disk. Always specify `width` and
`height` for reproducibility:

```{r}
#| eval: false
ggplot(framingham, aes(x = age, y = sysBP, color = male)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "loess", se = FALSE) +
  theme_minimal() +
  labs(x = "Age (years)", y = "Systolic BP (mmHg)", color = "Sex")

ggsave("figures/sysBP_vs_age.png", width = 7, height = 5, dpi = 300)
ggsave("figures/sysBP_vs_age.pdf", width = 7, height = 5)
```

::: {.callout-tip}
Use `.pdf` for vector graphics (scalable, best for publication) and `.png` for
raster graphics (best for slides and web). Always set `dpi = 300` for `.png`
to ensure print quality.
:::

## A Publication-Ready Figure

Here is a complete example building a figure from scratch with all the elements
covered above:

```{r}
#| message: false
#| fig-width: 8
#| fig-height: 5
framingham |>
  filter(!is.na(education)) |>
  ggplot(aes(x = age, y = sysBP, color = education)) +
  geom_point(alpha = 0.15, size = 0.8) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 1) +
  geom_hline(yintercept = 140, linetype = "dashed",
             color = "gray40", linewidth = 0.8) +
  annotate("text", x = 32, y = 143,
           label = "Hypertension threshold", color = "gray40",
           hjust = 0, size = 3) +
  scale_color_brewer(palette = "Set2") +
  scale_y_continuous(breaks = seq(80, 300, by = 40)) +
  facet_wrap(~ male) +
  theme_minimal() +
  theme(
    legend.position  = "bottom",
    panel.grid.minor = element_blank(),
    strip.text       = element_text(face = "bold")
  ) +
  labs(
    title    = "Systolic Blood Pressure vs. Age by Sex and Education Level",
    subtitle = "Framingham Heart Study cohort — linear trends per education group",
    caption  = "Dashed line indicates hypertension threshold (140 mmHg)",
    x        = "Age (years)",
    y        = "Systolic BP (mmHg)",
    color    = "Education level"
  )
```

---

# Introduction to Quarto

## What Is Quarto?

A **Quarto document** (`.qmd`) combines prose and R code in a single file.
When you **render** it, R runs all the code and produces a polished HTML, PDF,
or Word document with your text, code, and output together.

This is the foundation of **reproducible research**: your analysis and your
write-up live in the same place, and results update automatically if your data
or code changes.

::: {.callout-note}
**R Markdown (`.Rmd`) vs. Quarto (`.qmd`):** Quarto is the successor to R
Markdown. The syntax is nearly identical. If you have seen `.Rmd` files before,
everything you know transfers directly. We use `.qmd` in this course.
:::

## Anatomy of a Quarto Document

**1. YAML header** — at the very top, between `---` lines. Controls title,
format, and options:

```yaml
---
title: "My Analysis"
format: html
---
```

**2. Prose** written in Markdown:

```markdown
# Heading 1
## Heading 2

Regular paragraph text.

**bold**   *italic*   `inline code`

- bullet item
- another item
```

**3. Code chunks** — R code between triple backticks:

````markdown
```{{r}}
1 + 1
mean(c(4, 7, 2))
```
````

## Creating and Rendering Your First Document

**Step 1.** File → New File → Quarto Document. Give it a title, keep HTML
selected, click **Create**.

**Step 2.** Click **Render** in the editor toolbar (or `Ctrl+Shift+K` /
`Cmd+Shift+K`). The output opens in your browser or the Viewer pane.

**Step 3.** Try making a change — edit the title, add a line of prose, change
a number in a code chunk — then render again.

## Key Chunk Options

Control how each chunk behaves with options inside `#|` comments:

```{r}
#| echo: true      # show the code in the output (default)
#| eval: true      # run the code (default)
#| warning: false  # suppress warnings
#| message: false  # suppress messages like package loading notes

2 + 2
```

| Option | Default | Effect |
|--------|---------|--------|
| `echo` | `true` | Show the code in the rendered output |
| `eval` | `true` | Actually run the code |
| `warning` | `true` | Show/hide R warnings |
| `message` | `true` | Show/hide messages (e.g., from `library()`) |

Set options globally for the whole document in the YAML `execute:` block —
individual chunk options override the global defaults.

## Inline Code

Embed R results directly in prose:

```markdown
The mean age in the Framingham cohort is `{{r}} round(mean(framingham$age, na.rm = TRUE), 3)` years.
```
This renders as the following:


```markdown
The mean age in the Framingham cohort is `{r} round(mean(framingham$age, na.rm = TRUE), 3)` years.
```

The document and updates automatically if the data changes.
