---
title: "02 - Data Preprocessing"
subtitle: "Biostatistics 735 — Machine Learning in Public Health"
author: "Alexander McLain"
date: today
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    number-sections: true
    smooth-scroll: true
    theme: cosmo
    code-fold: false
    highlight-style: github
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| include: false
library(tidyverse)
library(janitor)
library(naniar)
library(rsample)
library(this.path)
setwd(this.dir())

framingham <- read_csv("framingham.csv") |>
  mutate(
    male          = factor(male,          levels = c(0,1), labels = c("Female","Male")),
    currentSmoker = factor(currentSmoker, levels = c(0,1), labels = c("Non-smoker","Smoker")),
    prevalentHyp  = factor(prevalentHyp,  levels = c(0,1), labels = c("No","Yes")),
    diabetes      = factor(diabetes,      levels = c(0,1), labels = c("No","Yes")),
    TenYearCHD    = factor(TenYearCHD,    levels = c(0,1), labels = c("No","Yes")),
    education     = factor(education,     levels = 1:4,
                           labels = c("Some HS","HS Diploma","Some College","College Degree"))
  )
```

# Overview

Raw data is almost never ready to feed into a model. Before you fit anything —
a linear model, a penalized regression, a tree — you need to decide how to
handle missing values, unusual observations, imbalanced outcomes, and variables
that are on the wrong scale or in the wrong format. These decisions are not
optional housekeeping; they change your results, and doing them incorrectly is
one of the most common sources of invalid machine learning results.

By the end of this document you will be able to:

- Diagnose and handle missing data
- Detect and reason about outliers
- Recognize class imbalance and know the basic strategies for addressing it
- Engineer and encode features so they are usable by a wide range of models
- Split data into training and test sets using the `rsample` package, and
  explain why this has to happen before (or alongside) the steps above

This document sits between **01 - Introduction to R** and **03 - Cross
Validation**. Everything here uses tools from document 1 (`dplyr`, `ggplot2`,
factors). The train/test split introduced at the end is the exact tool used at
the start of document 3 — read that section carefully, because we will not
re-explain `initial_split()` when it reappears there.

---

# Packages

```{r}
#| eval: false
install.packages(c("naniar", "rsample"))
```

| Package | Purpose |
|---|---|
| `naniar` | Summarizing and visualizing missing data |
| `rsample` | Creating train/test splits and (later) cross-validation folds |

We continue to use the cleaned `framingham` data frame from document 1
(binary/categorical variables already converted to factors).

---

# Why Preprocessing Matters

Classical statistical models (like `lm()`) are often forgiving: they will run
with missing data silently dropped, with outliers pulling a coefficient, with
a rare outcome class, and give you an answer — just possibly a misleading one.
Many machine learning methods are less forgiving in a different way: some
simply **fail** on missing data (`glmnet`, most tree ensembles via certain
implementations), some are **sensitive to scale** (penalized regression,
k-nearest neighbors, neural networks), and some are **sensitive to class
imbalance** in ways that are easy to miss (a model can achieve 98% accuracy by
always predicting "no disease" if disease is rare).

::: {.callout-important}
Preprocessing decisions should be made **before** you look at how they affect
model performance on your test set. Deciding how to impute missing data, or
which outliers to remove, based on what improves your test-set score is itself
a form of data leakage — a topic covered in depth in document 3.
:::

---

# Missing Data

## Diagnosing Missingness

Start by quantifying how much is missing, and where.

```{r}
framingham |>
  summarize(across(everything(), ~ mean(is.na(.)) * 100)) |>
  pivot_longer(everything(), names_to = "variable", values_to = "pct_missing") |>
  filter(pct_missing > 0) |>
  arrange(desc(pct_missing))
```

`naniar` gives a quick visual of the same information:

```{r}
gg_miss_var(framingham, show_pct = TRUE)
```

## Missingness Mechanisms (Briefly)

You do not need the full technical machinery to make reasonable decisions, but
it helps to know the vocabulary:

| Mechanism | Meaning | Example |
|---|---|---|
| **MCAR** — Missing Completely at Random | Missingness is unrelated to any variable, observed or not | A lab sample is dropped due to a shipping error |
| **MAR** — Missing at Random | Missingness depends on *observed* variables | Older participants are less likely to complete a cognitive test |
| **MNAR** — Missing Not at Random | Missingness depends on the *unobserved* value itself | Participants with very high blood pressure skip the BP reading |

::: {.callout-note}
Most simple imputation methods (mean/median imputation, complete-case
analysis) assume something close to MCAR. If missingness looks related to
other variables in your data (check this with `naniar::gg_miss_fct()` or by
comparing summary statistics between complete and incomplete cases), simple
methods can introduce bias. This course does not require formal MAR/MNAR
modeling — just be aware of the assumption you are making.
:::

## Handling Missing Data

**Option 1: Complete-case analysis (listwise deletion).** Drop any row with a
missing value in a variable you plan to use.

```{r}
framingham_cc <- framingham |>
  drop_na(totChol, BMI, heartRate, glucose)

nrow(framingham); nrow(framingham_cc)
```

This is simple and always valid under MCAR, but it throws away data — often a
lot of it, if missingness is spread across many variables — and can bias
results if missingness is not MCAR.

**Option 2: Simple imputation.** Replace missing values with a summary
statistic, most commonly the mean or median for numeric variables and the mode
for categorical variables.

```{r}
framingham_imp <- framingham |>
  mutate(
    glucose  = if_else(is.na(glucose),  median(glucose,  na.rm = TRUE), glucose),
    heartRate = if_else(is.na(heartRate), median(heartRate, na.rm = TRUE), heartRate)
  )
```

::: {.callout-important}
**The mean/median you impute with must come from the training data only.**
If you compute it on the full dataset (train + test) before splitting, or on
the test set itself, information from the test set leaks into your training
process. The correct order is: split first, then compute imputation values on
`train`, then apply those same values to both `train` and `test`. We will
split the data at the end of this document, and document 3 shows exactly this
pattern for centering and scaling.
:::

**Option 3: Model-based imputation.** Methods like k-nearest-neighbor
imputation or multiple imputation (`mice`, `VIM` packages) predict each
missing value from other variables rather than using a single summary
statistic. These are more accurate but add complexity and are beyond the scope
of this course; mean/median imputation or complete-case analysis will be
sufficient for our homework and projects unless stated otherwise.

---

# Outliers

## Detecting Outliers

Two common rules of thumb for a single numeric variable:

- **IQR rule:** flag values below $Q_1 - 1.5 \times \text{IQR}$ or above
  $Q_3 + 1.5 \times \text{IQR}$ (this is what `geom_boxplot()` flags as points).
- **Z-score rule:** flag values more than 3 (or so) standard deviations from
  the mean.

```{r}
ggplot(framingham, aes(y = totChol)) +
  geom_boxplot(fill = "steelblue", alpha = 0.7) +
  labs(y = "Total Cholesterol (mg/dL)",
       title = "Boxplot Flags Extreme Total Cholesterol Values")
```

```{r}
framingham |>
  mutate(
    q1 = quantile(totChol, 0.25, na.rm = TRUE),
    q3 = quantile(totChol, 0.75, na.rm = TRUE),
    iqr = q3 - q1,
    outlier_iqr = totChol < (q1 - 1.5 * iqr) | totChol > (q3 + 1.5 * iqr)
  ) |>
  count(outlier_iqr)
```

## Deciding What to Do

Flagging an outlier is not the same as knowing what to do with it. Ask:

- **Is it a data-entry error?** (e.g., height of 7 feet, age of 200) — fix or
  remove it.
- **Is it a real, extreme observation?** (e.g., a genuinely very high
  cholesterol reading) — this may be exactly the kind of patient your model
  needs to learn about. Removing it can hurt generalizability.
- **Is it distorting a distance- or scale-sensitive model** (k-NN, penalized
  regression, neural networks) without being an error? Consider a
  transformation (e.g., `log()`) or **winsorizing** (capping extreme values at
  a percentile) rather than deleting the observation.

```{r}
framingham |>
  mutate(
    totChol_capped = pmin(pmax(totChol, quantile(totChol, 0.01, na.rm = TRUE)),
                                quantile(totChol, 0.99, na.rm = TRUE))
  ) |>
  select(totChol, totChol_capped) |>
  summary()
```

::: {.callout-tip}
As with missing data, if you cap or remove outliers using percentiles
computed on the full dataset, you again risk leaking test information into
your training process. Compute the thresholds on `train` only.
:::

---

# Class Imbalance

`TenYearCHD` — whether a participant developed coronary heart disease within
10 years — is the outcome we will use in several classification examples this
semester. It is **imbalanced**: most participants did not develop CHD.

```{r}
framingham |>
  count(TenYearCHD) |>
  mutate(pct = n / sum(n) * 100)
```

## Why This Is a Problem

A classifier that always predicts "No" will be correct roughly 85% of the
time on this data — a high accuracy that tells you nothing useful. Imbalance
affects both **model fitting** (many algorithms implicitly optimize overall
accuracy and will lean toward the majority class) and **model evaluation**
(accuracy is the wrong metric to trust; sensitivity, specificity, and
precision-recall become important — covered later in the course).

## Basic Strategies

**Downsampling** the majority class:

```{r}
set.seed(1)
min_n <- framingham |> count(TenYearCHD) |> pull(n) |> min()

framingham_down <- framingham |>
  group_by(TenYearCHD) |>
  slice_sample(n = min_n) |>
  ungroup()

framingham_down |> count(TenYearCHD)
```

**Upsampling** the minority class (sampling with replacement):

```{r}
set.seed(1)
max_n <- framingham |> count(TenYearCHD) |> pull(n) |> max()

framingham_up <- framingham |>
  group_by(TenYearCHD) |>
  slice_sample(n = max_n, replace = TRUE) |>
  ungroup()

framingham_up |> count(TenYearCHD)
```

More sophisticated approaches (e.g., SMOTE, which generates synthetic
minority-class observations rather than duplicating real ones) are available
through packages like `themis`, but are not required for this course.

::: {.callout-important}
**Resampling for class balance is fit-time-only and train-only.** Never
downsample or upsample your test set — you want the test set to reflect the
real-world class distribution you will actually be predicting on. Balancing
happens on the training data, after the split, exactly like imputation and
outlier handling above.
:::

---

# Feature Engineering

## Transformations

Right-skewed variables (common in lab measurements, income, counts) often
benefit from a log transformation, especially for models sensitive to scale
and linearity.

```{r}
ggplot(framingham, aes(x = glucose)) +
  geom_histogram(bins = 40, fill = "steelblue", color = "white") +
  labs(title = "Glucose: Right-Skewed")

framingham |>
  mutate(log_glucose = log(glucose)) |>
  ggplot(aes(x = log_glucose)) +
  geom_histogram(bins = 40, fill = "steelblue", color = "white") +
  labs(title = "log(Glucose): More Symmetric")
```

## Centering and Scaling

Penalized regression, k-NN, and neural networks are sensitive to the *scale*
of predictors — a variable measured in the thousands will dominate a variable
measured in single digits, independent of its actual importance. Standardizing
(subtract the mean, divide by the standard deviation) puts variables on a
common scale.

```{r}
#| eval: false
framingham |>
  mutate(sysBP_z = as.numeric(scale(sysBP)))
```

::: {.callout-important}
Exactly like imputation: compute the mean and SD **on the training set**, and
apply those same values to both training and test data. Document 3 walks
through a full worked example of this — including what goes wrong when you
scale train and test independently — in its "Data Leakage" section.
:::

## Encoding Categorical Variables

`lm()` and similar formula-based functions handle factors automatically. Many
machine learning functions (`glmnet`, `xgboost`, k-NN implementations) instead
require a numeric matrix, so categorical variables need to be converted to
**dummy (one-hot) variables** first.

```{r}
model.matrix(~ education - 1, data = framingham) |> head()
```

Each level of `education` becomes its own 0/1 column. The `- 1` removes the
intercept column so that all levels are shown; when a model needs a reference
level instead (as `lm()` does automatically), drop one column.

Compare this to:

```{r}
model.matrix(~ education, data = framingham) |> head()
```

---

# Splitting Data: Training and Test Sets

Every preprocessing decision above — which imputation value to use, which
outlier thresholds to apply, how to balance classes, how to center and scale —
must be based on the **training data only**, then applied unchanged to the
**test data**. That guarantees the test set behaves like genuinely new,
unseen data, which is the entire point of having it.

This means the split has to happen early — conceptually, right after you've
decided what your outcome and candidate predictors are, and before you finalize
any of the steps above.

## The `rsample` Package

`rsample` (part of the `tidymodels` ecosystem) creates reproducible train/test
splits.

```{r}
set.seed(42)
split <- initial_split(framingham, prop = 0.6, strata = NULL)
train <- training(split)
test  <- testing(split)

dim(train); dim(test)
```

- `initial_split()` creates a **split object** that records which rows belong
  to training vs. test — it does not create two new data frames by itself.
- `prop = 0.6` means 60% of rows go to training, 40% to test. (Common choices
  range from 0.6–0.8 for training, depending on how much data you have.)
- `training(split)` and `testing(split)` extract the actual data frames.
- Setting `set.seed()` before splitting makes the split reproducible — you and
  a classmate running the same code will get the same rows in `train`.

::: {.callout-note}
This is exactly the code you will see reused at the start of document 3, where
`dat` replaces `framingham` as the data frame being split. Nothing new happens
there — if the code above makes sense, so will that.
:::

## The `strata` Argument

`initial_split()` has a `strata` argument, left as `NULL` above. When you
supply a variable to `strata`, the split preserves that variable's
distribution across both the training and test sets — the same logic as
stratified sampling anywhere else in statistics.

This matters most for **imbalanced outcomes**. Compare an unstratified split...

```{r}
set.seed(42)
split_plain <- initial_split(framingham, prop = 0.6, strata = NULL)
training(split_plain) |> count(TenYearCHD) |> mutate(pct = n / sum(n) * 100)
testing(split_plain)  |> count(TenYearCHD) |> mutate(pct = n / sum(n) * 100)
```

...to a stratified one:

```{r}
set.seed(42)
split_strat <- initial_split(framingham, prop = 0.6, strata = TenYearCHD)
training(split_strat) |> count(TenYearCHD) |> mutate(pct = n / sum(n) * 100)
testing(split_strat)  |> count(TenYearCHD) |> mutate(pct = n / sum(n) * 100)
```

With a rare outcome and a modest sample size, an unstratified split can, by
chance, leave the test set with a noticeably different event rate than the
training set. Stratifying removes that source of noise.

::: {.callout-tip}
Default to `strata = <your outcome variable>` for classification problems,
especially with class imbalance. For regression, `rsample` also accepts a
continuous variable for `strata` (it bins it internally) — document 3
introduces this in its cross-validation section.
:::

## Centering and Scaling Split Data

When centering and scaling training and test data, the most natural way to do it may seem like it is:

```{r}
vars_to_scale <- c("age", "sysBP", "totChol", "BMI", "heartRate", "glucose")

train_scaled <- scale(train[vars_to_scale]) 
test_scaled  <- scale(test[vars_to_scale]) 

summary(train_scaled)
summary(test_scaled)
```

**DO NOT DO THIS!!**

Centering and scaling your data based on the training data is important. Why? Consider a situation where you only had one covariate (say `age`) and the test data only had one subject. If you centered the test data individually, the mean would be the subjects value. So, they would have `age_z=0` no matter what. **Always center and scale based on the training data.**

Here's how we can do this:

```{r}
## Center and scale numeric predictors: fit on train, apply to test

vars_to_scale <- c("age", "sysBP", "totChol", "BMI", "heartRate", "glucose")

scaled_train <- scale(train[vars_to_scale]) # computes mean/SD from train

str(scaled_train)

# Extract the mean and sd from the scaled_train
mu <- attr(scaled_train, "scaled:center")
sdv <- attr(scaled_train, "scaled:scale")

train_scaled <- train
test_scaled  <- test
train_scaled[vars_to_scale] <- scaled_train
test_scaled[vars_to_scale]  <- scale(test[vars_to_scale], center = mu, scale = sdv)

summary(train_scaled[vars_to_scale])
summary(test_scaled[vars_to_scale])
```



The test data does not exactly have mean zero and that is okay. It shouldn't. The test data is supposed to represent what we could see in practice, and it will not exactly match the training data.

## What Comes Next

Once `train` and `test` exist, every step from this document gets applied in
order: diagnose and handle missingness, check outliers, address imbalance,
engineer and encode features, and center/scale — fit on `train`, applied to
both. Document 3 extends this same idea one level further, splitting `train`
itself repeatedly via cross-validation (`vfold_cv()`) to tune models honestly
without ever touching `test` until the very end.

---

# Preprocessing Checklist

1. Split into training and test sets (`strata` on the outcome if imbalanced).
2. Diagnose missingness in the training set; choose complete-case analysis or
   an imputation strategy and compute any imputation values from `train` only.
3. Check for outliers; decide fix, cap, transform, or leave, based on domain
   knowledge, not on what improves test performance.
4. Check for class imbalance in the outcome; decide whether to up/downsample
   `train` (never `test`).
5. Engineer features: transformations, interactions, encoding categorical
   variables as needed for the model you plan to use.
6. Center/scale numeric predictors if using a scale-sensitive model, fit on
   `train`, applied to `test`.
7. Apply every choice from steps 2–6 to `test` using values learned from
   `train` — never recompute them on `test`.

---

# Summary

- Missing data, outliers, and class imbalance are the norm in public health
  data, not the exception — diagnose before deciding how to handle them.
- Every preprocessing choice that involves estimating something from data
  (an imputation value, an outlier threshold, a centering/scaling parameter)
  must be estimated on the training set only.
- `rsample::initial_split()`, `training()`, and `testing()` create the
  train/test split that makes this possible, and `strata` protects you from
  an unlucky split when the outcome is imbalanced.
- This train/test split is the first line of code in document 3 — everything
  from here forward assumes you know what it does.
