---
title: "Biased Regression Methods for Prediction"
subtitle: "Instability of OLS, Ridge, Lasso, PLS, MCP, SCAD; Grouped & Fused Lasso"
author: "Alexander McLain"
date: today
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    number-sections: true
    smooth-scroll: true
    theme: cosmo
    code-fold: false
    highlight-style: github
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| include: false
set.seed(735794)

# Methods
library(glmnet)   # ridge/lasso
library(pls)      # partial least squares
library(lars)     # diabetes data (public health: diabetes progression)
library(ncvreg)   # MCP/SCAD
library(grpreg)   # grouped lasso
library(genlasso) # fused lasso / trend filtering

# Core
library(broom)
library(rsample)
library(yardstick)
library(knitr)
library(kableExtra)
library(GGally)
library(tidyverse) # It is good to load this last.Do
```

**Datasets used in this lab**

> • **Diabetes progression** (Efron et al.) via `lars::diabetes`: outcome is a quantitative measure of disease progression one year after baseline; predictors are clinical/lab measures.

> • **Air quality (Ozone)** via `airquality`: daily ozone levels in New York—used here to illustrate **fused lasso** as a denoising tool for an environmental time series.

---

# Data and Train/Test Split (Diabetes)

```{r}
data(diabetes, package = "lars")

# X: 10 predictors (matrix), y: outcome (numeric)
X <- diabetes$x
y <- diabetes$y

# Make a tibble with one column per predictor (no list-columns)
df <- tibble::as_tibble(X) %>% 
  mutate(y = as.numeric(y))

## This data is formatted strangely. I had to export and import to fix.
write.csv(df, "diabetes.csv")
df <- read_csv("diabetes.csv")[,-1]
names(df) <- c("age","sex","bmi","map","tc","ldl","hdl","tch","ltg","glu", "y")

# correlation of predictors
cor(df %>% dplyr::select(!starts_with("y"))) %>%
  round(2) %>%
  kable()
```

```{r}
set.seed(6545)
split <- initial_split(df, prop = 0.7, strata = NULL)
train <- training(split)
test  <- testing(split)

X_tr <- as.matrix(select(train, -y))
y_tr <- train$y
X_te <- as.matrix(select(test,  -y))
y_te <- test$y

rmse <- function(truth, estimate) sqrt(mean((truth - estimate)^2))
```

---

# Instability of Ordinary Least Squares (OLS)

OLS can be **unstable** when predictors are correlated and $p$ is not tiny relative to $n$. We’ll bootstrap OLS to show variability in coefficients and predictions.

```{r}
# Full OLS on training
ols <- lm(y ~ ., data = train)
summary(ols)
```

## Bootstrap variability of coefficients

```{r}
set.seed(123)
B <- 300
boot_coefs <- replicate(B, {
  idx <- sample(nrow(train), replace = TRUE)
  coef(lm(y ~ ., data = train[idx, ]))
})

boot_df <- as_tibble(t(boot_coefs)) %>%
  set_names(names(coef(ols))) %>%
  pivot_longer(everything(), names_to = "term", values_to = "beta")

boot_df %>%
  filter(term != "(Intercept)") %>%
  ggplot(aes(x = term, y = beta)) +
  geom_violin(fill = "grey90") +
  geom_hline(yintercept = 0, linetype = 2, color = "red") +
  coord_flip() +
  labs(title = "Bootstrap variability of OLS coefficients",
       x = NULL, y = "Coefficient") +
  theme_minimal(base_size = 12)
```

## OLS generalization error

```{r}
pred_ols <- predict(ols, newdata = test)
ols_rmse <- rmse(y_te, pred_ols)
tibble(Method = "OLS", Test_RMSE = ols_rmse) %>% kable(digits = 3)
```

---

# Ridge Regression (L2)

To fit ridge and lasso regresssion models we'll use the `glmnet` function. 


```{r}
set.seed(1)
cv_ridge <- cv.glmnet(
  X_tr, 
  y_tr, 
  alpha = 0, 
  standardize = TRUE, 
  nfolds = 10 
)
plot(cv_ridge)

set.seed(1)
cv_ridge <- cv.glmnet(
  X_tr, 
  y_tr, 
  alpha = 0, 
  standardize = TRUE, 
  nfolds = 10, 
  ## I usually don't specify lambda
  lambda = exp(seq(-5, 5, length.out = 200)) 
)
plot(cv_ridge)

ridge_fit_min <- glmnet(
  X_tr, 
  y_tr, 
  alpha = 0, 
  lambda = cv_ridge$lambda.min, 
  standardize = TRUE
  )

ridge_fit_1se <- glmnet(
  X_tr, 
  y_tr, 
  alpha = 0, 
  lambda = cv_ridge$lambda.1se, 
  standardize = TRUE
  )

pred_ridge_min <- predict(ridge_fit_min, X_te)
pred_ridge_1se <- predict(ridge_fit_1se, X_te)
rmse_min = rmse(y_te, pred_ridge_min) 
rmse_1se = rmse(y_te, pred_ridge_1se)

bind_cols(
  coef_ols = coef(ols),
  coef_ridge_min = coef(ridge_fit_min)[,1],
  coef_ridge_1se = coef(ridge_fit_1se)[,1]
) %>% as.matrix() %>% round(3)

```

---

# Lasso (L1)

```{r}
set.seed(2)
cv_lasso <- cv.glmnet(X_tr, y_tr, alpha = 1, standardize = TRUE, nfolds = 10)
lasso_min  <- glmnet(X_tr, y_tr, alpha = 1, lambda = cv_lasso$lambda.min,  standardize = TRUE)
lasso_1se  <- glmnet(X_tr, y_tr, alpha = 1, lambda = cv_lasso$lambda.1se,  standardize = TRUE)

pred_lasso_min <- predict(lasso_min, X_te)
pred_lasso_1se <- predict(lasso_1se, X_te)

lasso_min_rmse <- rmse(y_te, pred_lasso_min)
lasso_1se_rmse <- rmse(y_te, pred_lasso_1se)

coef(lasso_min)[,1] %>% as.matrix() %>% round(3) 
```

---

# Partial Least Squares (PLS)

PLS builds latent components that maximize covariance between $X$ and $y$.

```{r}
set.seed(3)
pls_fit <- plsr(y ~ ., data = train, validation = "CV")
plot(RMSEP(pls_fit), legendpos = "topright", main = "PLS: CV RMSEP vs #components")

# Choose ncomp with lowest CV error (or 1-SE rule)
best_comp <- which.min(RMSEP(pls_fit)$val[estimate = "CV",,, drop = TRUE])
best_comp

# Extract the coefficients
pls_coef <- coef(pls_fit, ncomp = best_comp)


pred_pls <- predict(pls_fit, newdata = test, ncomp = best_comp)[,,1]
pls_rmse <- rmse(y_te, pred_pls)
```

---

# MCP and SCAD (Nonconvex Penalties)

We use `ncvreg` for linear regression with MCP/SCAD penalties; $\lambda$ is tuned via cross-validation.

```{r}
set.seed(4)
cv_mcp  <- cv.ncvreg(X_tr, y_tr, penalty = "MCP",  family = "gaussian", nfolds = 10)
plot(cv_mcp)
cv_scad <- cv.ncvreg(X_tr, y_tr, penalty = "SCAD", family = "gaussian", nfolds = 10)
plot(cv_scad)

mcp_fit  <- cv_mcp$fit
scad_fit <- cv_scad$fit

pred_mcp  <- predict(mcp_fit,  X_te)
pred_scad <- predict(scad_fit, X_te)

mcp_rmse  <- rmse(y_te, pred_mcp)
scad_rmse <- rmse(y_te, pred_scad)


### Beta Coefficients
bind_cols(
  coef_mcp = mcp_fit$beta[,cv_mcp$min],
  coef_scad = scad_fit$beta[,cv_scad$min]
) %>% as.matrix() %>% round(3) %>% kable()

```

---

# Summary: Test RMSE and Sparsity

```{r}
summ <- tibble(
  Method = c("OLS","Ridge (min)","Ridge (1SE)","Lasso (min)","Lasso (1SE)","PLS","MCP","SCAD"),
  Test_RMSE = c(
    ols_rmse,
    rmse_min,
    rmse_1se,
    lasso_min_rmse,
    lasso_1se_rmse,
    pls_rmse,
    mcp_rmse,
    scad_rmse
  ),
  Nonzero = c(
    sum(coef(ols)[-1]!=0),
    sum(coef(ridge_fit_min )[-1]!=0),
    sum(coef(ridge_fit_1se )[-1]!=0),
    sum(coef(lasso_min )[-1]!=0),
    sum(coef(lasso_1se )[-1]!=0),
    sum(pls_coef != 0),
    sum(mcp_fit$beta[,cv_mcp$min][-1]!=0),
    sum(scad_fit$beta[,cv_scad$min][-1]!=0)
  )
) %>%
  arrange(Test_RMSE)

summ %>% kable(digits = 3, caption = "Method comparison on held-out test set") %>% kable_styling(full_width = FALSE)
```



```{r}
summ_coef <- tibble(
  Variable = attr(coef(ols)[-1], "names"),
    OLS = coef(ols)[-1],
    Ridge_min = as.vector(coef(ridge_fit_min ))[-1],
    Ridge_1SE = as.vector(coef(ridge_fit_1se ))[-1],
    Lasso_min = as.vector(coef(lasso_min ))[-1],
    Lasso_1SE = as.vector(coef(lasso_1se ))[-1],
    PLS = pls_coef[1:10],
    MCP = mcp_fit$beta[,cv_mcp$min][-1],
    SCAD = scad_fit$beta[,cv_scad$min][-1]
  )

summ_coef %>% kable(digits = 1, caption = "Comparison of Coefficients") %>% kable_styling(full_width = FALSE)
```

---

# Grouped Lasso (Groups by Clinical Domain)

The diabetes data naturally groups predictors:

* **Demographics:** `age`, `sex`
* **Anthropometrics:** `bmi`, `map`
* **Serum markers:** `tc`–`glu`

We’ll fit a **grouped lasso** to respect these groupings (encouraging selection/deselection by group).

```{r}
# Build X and groups
X <- df %>% 
  select(-"y")
grp_names <- c("age","sex","bmi","map","tc","ldl","hdl","tch","ltg","glu")
stopifnot(all(grp_names %in% names(X)))

gvec <- c(1,1, 2,2, 3,3,3,3,3,3)  # (age,sex)=1; (bmi,bp)=2; (tc..glu)=3
groups <- factor(gvec, labels = c("demo","anthro","serum"))

X_tr_g <- as.matrix(X_tr[, grp_names])
X_te_g <- as.matrix(X_te[, grp_names])

set.seed(5)
cv_grp <- cv.grpreg(X_tr_g, y_tr, group = gvec, penalty = "grLasso", nfolds = 10, family = "gaussian")
grp_fit <- grpreg(X_tr_g, y_tr, group = gvec, penalty = "grLasso", lambda = cv_grp$lambda.min)

pred_grp <- predict(grp_fit, X_te_g)
grp_rmse <- rmse(y_te, pred_grp)
grp_rmse
```

Inspect which **groups** entered:

```{r}
beta_grp <- coef(grp_fit)[-1]  # drop intercept
tibble(var = grp_names, beta = as.numeric(beta_grp),
       group = groups) %>%
  group_by(group) %>%
  summarise(any_selected = any(beta != 0), n_nonzero = sum(beta != 0)) %>%
  kable(caption = "Grouped-lasso selection by group")
```

---

# Fused Lasso (Environmental Time Series Denoising)

Fused lasso encourages **piecewise-constant** signals by penalizing the sum of
absolute differences between adjacent coefficients (a total-variation penalty).
Unlike generalized trend filtering, this is the original fused lasso estimator
(Tibshirani et al., 2005), fit directly with `genlasso::fusedlasso1d()`. We
denoise **daily Ozone** (public health air quality) from the `airquality`
dataset.

```{r}


oz <- airquality %>%
  mutate(date = as.Date(sprintf("1973-%02d-%02d", Month, Day))) %>%
  arrange(date) %>%
  select(date, Ozone) %>%
  drop_na()

plot(oz$date, oz$Ozone, type = "l", xlab = "Date", ylab = "Ozone (ppb)",
     main = "Daily Ozone (New York, 1973)")
```

Fit the fused lasso path and inspect it to choose a reasonable amount of
smoothing:

```{r}
y <- oz$Ozone

# Bona fide fused lasso: penalizes sum(|beta_i - beta_{i-1}|)
fl <- fusedlasso1d(y)

plot(fl, main = "Fused lasso solution path")

# Pick a moderately smooth fit along the path (adjust after visual inspection)
lambda_choice <- fl$lambda[50]
beta_hat <- coef(fl, lambda = lambda_choice)$beta
```

```{r}
plot(oz$date, y, type = "l", col = "gray40",
     xlab = "Date", ylab = "Ozone (ppb)",
     main = "Fused Lasso Fit to Daily Ozone")
lines(oz$date, beta_hat, col = "red", lwd = 2)
legend("topright", legend = c("Observed", "Fused lasso fit"),
       col = c("gray40", "red"), lty = 1, lwd = c(1, 2))
```

> **Interpretation:** the fused lasso identifies **regime shifts/plateaus** in
ozone levels — stretches of days the model treats as a stable pollution
level, separated by abrupt jumps. This is useful for environmental health
monitoring and change-point detection, and unlike a moving average or spline
smoother, it does not blur across a genuine jump in the underlying signal.
---

# Takeaways

* **OLS** can be unstable with correlated predictors; bootstrap variability shows why shrinkage helps.
* **Ridge** shrinks coefficients (lower variance), **Lasso/MCP/SCAD** add sparsity for interpretability.
* **PLS** is strong when predictors are highly collinear and signal is low-dimensional in the latent space.
* **Grouped lasso** encodes domain structure (e.g., demographics vs labs), selecting groups coherently.
* **Fused lasso** is powerful for ordered structures (time, space, genomic position), producing interpretable plateaus.

---

```{r}
sessionInfo()
```
