---
title: "06 - Splines and Smoothing"
subtitle: "Biostatistics 735 — Machine Learning in Public Health"
author: "Alexander McLain"
date: today
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    number-sections: true
    smooth-scroll: true
    theme: cosmo
    code-fold: false
    highlight-style: github
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| include: false
set.seed(735794)
library(tidyverse)
library(car)      # crPlots()
library(mgcv)     # gam(), s()
library(refund)   # functional regression, DTI data
```

# Overview

Splines and GAMs show up in this course in two practical situations, and we
will take each in turn:

1. **Checking the functional form of a predictor.** You already have a
   regression model with ordinary scalar predictors (age, BMI, blood
   pressure...). The question is whether each predictor should enter linearly,
   or whether the relationship is curved enough that a linear term is
   misspecifying the model.
2. **A predictor that is genuinely a curve, not a number.** Sometimes a single
   observation's "predictor" is an entire function — a growth curve, a
   biomarker trajectory, an imaging tract profile — measured at many points.
   This is not a functional-form question; the covariate itself is functional.

Both situations are handled with `mgcv::gam()` and its `s()` smooth-term
syntax. We will not construct spline basis matrices by hand — `gam()` does
that internally, chooses the smoothing parameter for you (via REML/ML), and
gives you confidence bands and significance tests directly.

---

# Checking Functional Form in a Linear Model

We use the body fat data, predicting `bodyfat` from `abdomen`, `height`, and
`age`.

```{r}
bf_df <- read_csv("bodyfat2.csv") |>
  select(bodyfat, abdomen, height, age)

head(bf_df)
```

This is the standard exploratory workflow for deciding whether a predictor
needs anything beyond a linear term: start with marginal plots, move to
partial-residual plots once other covariates are in the model, then let a GAM
tell you directly whether curvature is actually needed.

## Step 1: Marginal Scatterplots

```{r}
bf_df |>
  pivot_longer(c(abdomen, height, age), names_to = "predictor", values_to = "x") |>
  ggplot(aes(x, bodyfat)) +
  geom_point(alpha = 0.4) +
  geom_smooth(se = FALSE, color = "firebrick") +
  facet_wrap(~ predictor, scales = "free_x") +
  theme_minimal(base_size = 13) +
  labs(x = NULL, y = "Body Fat (%)",
       title = "Marginal Relationship Between Each Predictor and Body Fat")
```

::: {.callout-important}
A scatterplot like this shows the **marginal** relationship between one
predictor and the outcome — it ignores every other variable in the model.
Curvature (or the lack of it) here is a starting point, not a final answer:
once `abdomen`, `height`, and `age` are all in the same model together, the
apparent shape for any one of them can change.
:::

## Step 2: Partial-Residual Plots (`car::crPlots`)

Fit the ordinary linear model first, with every predictor entered linearly:

```{r}
lm_fit <- lm(bodyfat ~ abdomen + height + age, data = bf_df)
summary(lm_fit)
```

`car::crPlots()` gives **component-plus-residual (partial-residual) plots** —
for each predictor, the relationship with the outcome *after adjusting for*
the other predictors in the model. This is the same tool we used earlier in
the course for checking linearity assumptions.

```{r}
crPlots(lm_fit)
```

Each panel shows a dashed line (the fitted linear term) against a solid loess
line (a smooth fit to the partial residuals). When the two lines track closely,
a linear term is reasonable. When the loess line bows away from the dashed
line, that predictor's relationship with the outcome is probably not linear —
`abdomen` is the one to watch here.

## Step 3: Let `gam()` Decide

Rather than eyeballing the crPlots, fit a GAM with a smooth term for each
predictor and let `mgcv` estimate how much curvature the data actually
support.

```{r}
gam_fit <- gam(bodyfat ~ s(abdomen) + s(height) + s(age),
               data = bf_df, method = "REML")
summary(gam_fit)
```

Two things to read from the summary table:

- **`edf`** (effective degrees of freedom): `edf ≈ 1` means the smooth term
  has effectively collapsed to a straight line — no evidence of curvature.
  `edf` noticeably above 1 means the model is using real curvature to fit
  that predictor.
- **The p-value** for each smooth term tests whether that term contributes
  anything beyond a flat/no effect — not specifically whether it's nonlinear.

```{r}
plot(gam_fit, pages = 1, shade = TRUE, residuals = TRUE, pch = 1, cex = 0.5)
```

These plots are the GAM analogue of the crPlots above: partial effect of each
predictor (adjusted for the others) with a confidence band and the partial
residuals overlaid.

## Step 4: Linear vs. Smooth — a Direct Comparison

To formally check whether the smooth terms are earning their keep, compare a
fully linear model to the GAM (both fit with `method = "ML"` so the
comparison is valid):

```{r}
lm_ml   <- gam(bodyfat ~ abdomen + height + age, data = bf_df, method = "ML")
gam_ml  <- gam(bodyfat ~ s(abdomen) + s(height) + s(age), data = bf_df, method = "ML")

anova(lm_ml, gam_ml, test = "F")
```

::: {.callout-tip}
**Practical takeaway.** If `edf` is close to 1 for a predictor and the
linear-vs-smooth comparison doesn't favor the GAM, drop the smooth and use the
plain linear term — it's simpler and just as good. Reserve `s()` for the
predictors that actually show curvature, like `abdomen` here. You do not need
to smooth every predictor just because you smoothed one.
:::

---



# A Genuinely Functional Covariate

The example above still has ordinary, scalar predictors — the only question
was the *shape* of their effect. Here the predictor itself is a curve: at each
observation, we don't observe a single abdomen measurement, we observe an
entire profile measured across many positions.

We'll use a Diffusion Tensor Imaging (DTI) dataset from the `refund` package: diffusion tensor
imaging data from a study of multiple sclerosis. For each subject/visit, we
have **fractional anisotropy (FA),** FA is a number from 0 to 1 that measures how much water molecules move in a specific direction versus all directions. 

- *High FA (closer to 1):* Shows healthy, dense, and well-organized white matter tracks.
- *Low FA (closer to 0):* Shows damaged, less dense, or disorganized fibers, often caused by aging, injury, or diseases like multiple sclerosis and depression. 

Here, we have FA measured at 93 positions along the corpus callosum — a white-matter tract profile — along with a cognitive processing
score, `pasat` (Paced Auditory Serial Addition Test).

```{r}
data(DTI)
str(DTI)
```

`DTI$cca` is a matrix: one row per observation, one column per tract position.
That whole row — not any single number from it — is the predictor.

```{r}
dti0 <- DTI |>
  filter(visit == 1, !is.na(pasat), rowSums(is.na(cca)) == 0)

matplot(t(dti0$cca), type = "l", lty = 1,
        col = adjustcolor("steelblue", alpha.f = 0.3),
        xlab = "Position along corpus callosum tract",
        ylab = "Fractional anisotropy",
        main = "CCA Tract Profiles — One Curve per Subject")
```

## Scalar-on-Function Regression

We want to relate this curve to a scalar outcome (`pasat`). This is
**scalar-on-function regression**, fit here with `refund::pfr()`, which
extends the same `gam()` machinery to functional predictors via the `lf()`
("linear functional") term.

```{r}
fit_func <- pfr(pasat ~ lf(cca), data = dti0)
summary(fit_func)
```

```{r}
plot(fit_func, shade = TRUE,
     xlab = "Position along tract",
     ylab = expression(beta(t)),
     main = "Estimated Coefficient Function")
abline(h = 0, col = "red", lty = 2)
```

::: {.callout-important}
The plot above is not a partial-effect curve for one predictor among several
— it **is** the model's entire estimated relationship between the tract
profile and PASAT score. $\hat\beta(t)$ tells you, at each position $t$ along
the tract, how strongly FA at that position is associated with the outcome.
Where the confidence band excludes zero, that region of the tract is driving
the association; where it doesn't, that region isn't contributing much.
:::

This is a fundamentally different use of smoothing than Part 1: there, `s()`
let us relax a linearity assumption on an otherwise ordinary predictor. Here,
smoothness is doing double duty — it is how we represent a predictor that has
no natural "linear" form to begin with, since there is no single number to
enter into a regression.

::: {.callout-note}
**Connection to the fused lasso.** The scalar-on-function regression above
estimates $\hat\beta(t)$ by imposing a *smoothness* penalty (roughness of the
coefficient curve, via the spline basis in `pfr()`). The fused lasso example
from the penalized-regression material imposes a different kind of structure:
instead of penalizing curvature, it penalizes $\sum_t |\beta_t - \beta_{t-1}|$,
which encourages $\hat\beta(t)$ to be **piecewise constant** rather than
smoothly varying.

Both are reasonable choices for a coefficient that's indexed by an ordered
position (here, location along the tract) rather than a arbitrary variable:

| | Smoothing spline / GAM (`pfr`) | Fused lasso |
|---|---|---|
| Assumes | $\beta(t)$ varies smoothly | $\beta(t)$ is flat within regions, with abrupt jumps between them |
| Penalty | On curvature (2nd derivative) | On adjacent differences ($\ell_1$) |
| Estimated shape | Smooth curve | Step function / plateaus |
| Best suited to | Gradual biological gradients | Genuine regime changes — e.g., a tract segment that matters and a neighboring segment that doesn't |
| Coefficients can be exactly zero? | Only approximately, over a whole smooth region | Yes — the $\ell_1$ penalty can zero out entire flat segments |

For the DTI tract data, a smooth $\hat\beta(t)$ is the more natural
assumption — anatomy doesn't usually have sharp discontinuities. But if you
believed only a specific, contiguous stretch of the tract mattered and the
rest was truly irrelevant, a fused lasso on $\beta(t)$ (fusing adjacent
positions, with an $\ell_1$ penalty that can drive whole segments to exactly
zero) would recover that block structure more directly than a smoother would.
:::

---


# Using two-dimensional tensor regression

We'll now look at an example which fits a multidimensional smoothing spline using _tensor product basis_.  This action will be performed using the \texttt{mgcv} package we discussed in the last example.  

```{r printr.help.sections=c('description', 'usage', 'arguments','details')}
#| comment: ''
#| tidy: false
#| error: false

#Fit an ANOVA type tensor product interaction
b2 <- gam(bodyfat~ ti(abdomen) + ti(height) + ti(abdomen,height),
          data=bf_df, 
          method = "REML")
summary(b2)
#Estimated smoothing parameters for the model.
b2$sp

#Plot the results.
vis.gam(b2, theta = 50);title("tensor anova")
```


---

# Summary

- Use marginal scatterplots for a first look, but remember they ignore every
  other covariate in the model.
- `car::crPlots()` gives the partial-residual view once other predictors are
  accounted for — the tool we've already been using for linearity checks.
- `mgcv::gam()` with `s()` terms lets the data decide how much curvature is
  needed; check `edf` and compare linear vs. smooth models with `anova()`
  rather than smoothing every predictor by default.
- When the predictor itself is a curve (a tract profile, a growth trajectory,
  a biomarker measured densely over time), that's a different problem —
  scalar-on-function regression (`refund::pfr()` with `lf()`) — not a
  functional-form question about a scalar predictor.
- There's nothing written in the sky saying that the relationship between variables and outcomes will be simple. There can be two- and even three-way non-linear interactions. 
