---
title: "07 - Mock Analysis"
subtitle: "Biostatistics 735 — Machine Learning in Public Health"
author: "Alexander McLain"
date: today
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    number-sections: true
    smooth-scroll: true
    theme: cosmo
    code-fold: false
    highlight-style: github
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| include: true
set.seed(735794)
library(tidyverse)
library(janitor)
library(rsample)   # group_vfold_cv(), rolling_origin()
library(mgcv)      # gam()
library(genlasso)  # fusedlasso1d()
library(glmnet)
library(grpreg)     # group lasso
library(nhanesA)    # NHANES data pull
library(BGLR)       # wheat data
library(kableExtra)

select <- dplyr::select 
```


# Scenario 1: NHANES 

**Goal:** Predict systolic blood pressure from
demographic and body-measurement variables using a recent NHANES cycle.

**Intended CV issue:** NHANES uses a complex, clustered survey design.
Participants are sampled within primary sampling units (`SDMVPSU`), nested
within strata (`SDMVSTRA`). Observations within the same PSU are more similar
to each other than to observations in a different PSU (shared geography,
recruitment site, local population characteristics). A random row-level CV
split puts people from the same PSU in both training and validation folds,
which overstates performance — the same leakage mechanism as splitting
repeated visits on the same subject, just from geographic clustering instead
of time.

**Intended method:** GAM, since age and BMI effects on blood pressure are
plausibly nonlinear (we just covered `s()` terms) and NHANES is large enough
to estimate smooth terms comfortably.

## Pulling the Data

```{r}
demo <- nhanes("DEMO_J")   # 2017-2018 cycle: demographics, SDMVPSU, SDMVSTRA
bpx  <- nhanes("BPX_J")    # blood pressure exam
bmx  <- nhanes("BMX_J")    # body measures

str(demo[, c("SEQN", "RIDAGEYR", "RIAGENDR", "SDMVPSU", "SDMVSTRA")])
```

```{r}
# SDMVPSU only takes values 1–2 and is nested within each stratum, 
# not a standalone cluster ID. To get a cluster ID we need to combine
# psu and strata.
nhanes_df <- demo |>
  select(SEQN, age = RIDAGEYR, sex = RIAGENDR, race.eth = RIDRETH3, 
         educ = DMDEDUC2, married = DMDHRMAZ, 
         psu = SDMVPSU, strata = SDMVSTRA) |>
  left_join(bpx |> select(SEQN, sysBP = BPXSY1), by = "SEQN") |>
  left_join(bmx |> select(SEQN, BMI = BMXBMI), by = "SEQN")  |> 
  mutate(cluster = interaction(strata, psu, drop = TRUE)) |>
  drop_na()

nrow(nhanes_df)
n_distinct(nhanes_df$cluster)   # should be 30 (15 strata x 2 psu)
table(nhanes_df$cluster)

```


## The Wrong Way: Naive Row-Level CV

```{r}
set.seed(1)
folds_naive <- vfold_cv(nhanes_df, v = 5)

rmse_naive <- map_dbl(folds_naive$splits, function(s) {
  tr <- analysis(s); va <- assessment(s)
  fit <- gam(sysBP ~ s(age, by = sex) + s(BMI, by = sex) + sex + 
               race.eth + educ + married, data = tr, method = "REML")
  sqrt(mean((va$sysBP - predict(fit, newdata = va))^2))
})

mean(rmse_naive)
```

## The Right Way: Grouping by Cluster

```{r}
set.seed(1)
folds_psu <- group_vfold_cv(nhanes_df, group = cluster, v = 5)

rmse_grouped <- map_dbl(folds_psu$splits, function(s) {
  tr <- analysis(s); va <- assessment(s)
  fit <- gam(sysBP ~ s(age, by = sex) + s(BMI, by = sex) + sex + 
               race.eth + educ + married, data = tr, method = "REML")
  sqrt(mean((va$sysBP - predict(fit, newdata = va))^2))
})

mean(rmse_grouped)
```

```{r}
tibble(
  CV_design = c("Naive (row-level)", "Grouped by Cluster"),
  Mean_RMSE = c(mean(rmse_naive), mean(rmse_grouped))
)
```

**Debrief point:** if the naive CV RMSE is meaningfully lower than the grouped
CV RMSE, that gap *is* the leakage — it's an estimate of how much the model
was "cheating" by seeing correlated observations from the same PSU on both
sides of the split.


## A New Question: Does Error Vary by Race/Ethnicity

In this particular dataset, the overall naive vs. grouped RMSE comparison
earlier didn't show much of a gap — so it might be tempting to assume the
subgroup breakdown above is fine as is. However, a small or absent gap in the *overall* RMSE doesn't guarantee a small
gap in every *subgroup's* RMSE, since leakage severity can vary by how
concentrated each group is across PSUs. The only way to know is to rerun the
breakdown on `folds_psu` and compare directly, I've done that and there still isn't a difference. However, here, we'll still use the correct `folds_psu`.


```{r}
set.seed(1)
grouped_preds <- map_dfr(folds_psu$splits, function(s) {
  tr <- analysis(s); va <- assessment(s)
  fit <- gam(sysBP ~ s(age, by = sex) + s(BMI, by = sex) + sex +
               race.eth + educ + married, data = tr, method = "REML")

  va |>
    mutate(pred = predict(fit, newdata = va)) |>
    select(race.eth, sysBP, pred)
})

rmse_by_race_grouped <- grouped_preds |>
  mutate(sq_err = (sysBP - pred)^2) |>
  group_by(race.eth) |>
  summarize(
    n       = n(),
    avg_obs = mean(sysBP),
    avg_pred = mean(pred),
    mse     = mean(sq_err),
    se_mse  = sd(sq_err) / sqrt(n),
    rmse    = sqrt(mse),
    se_rmse = se_mse / (2 * rmse),        # delta method: SE(sqrt(x)) ~= SE(x) / (2*sqrt(x))
    ci_lower = rmse - 1.96 * se_rmse,
    ci_upper = rmse + 1.96 * se_rmse,
    .groups = "drop"
  ) |>
  arrange(desc(rmse))

rmse_by_race_grouped |>
  transmute(
    `Race/Ethnicity` = race.eth,
    `n`              = n,
    `Avg Obs` = round(avg_obs,1),
    `Avg Pred` = round(avg_pred,1),
    `Avg Diff` = round(avg_obs - avg_pred,2),
    `RMSE`           = round(rmse, 2),
    `SE`             = round(se_rmse, 3),
    `95% CI`         = paste0("(", round(ci_lower, 2), ", ", round(ci_upper, 2), ")")
  ) |>
  kable(align = "lrrrr", caption = "RMSE by Race/Ethnicity with Approximate 95% CIs")
```


---

# Scenario 2: Testing Volume and Percent Positive

**Goal:** Using weekly test-volume and
percent-positive data from the five most populous states, predict (a) total
weekly specimens tested and (b) percent positive, for future weeks.

**CV issue:** Two, layered. First, temporal ordering — the outcome
in week $t$ is correlated with weeks $t-1, t-2, \dots$, so a random split lets
the model see the future. CV has to be a rolling-origin / forward-chaining
design. Second, if state is entered as a fixed effect, the fitted model
cannot generalize to a state it never saw in training — worth surfacing even
though we aren't holding out a state here, because it constrains what this
model can and can't be used for later.

**Intended methods:** GAM and fused lasso (captures the piecewise trend / regime
shifts in each state's series — a direct callback to the fused-lasso lesson). For `state` we can use a fixed-effects linear model since there are only 5 states.


## Data Setup

This uses CDC's ILINet clinical-lab surveillance file directly. I downloaded this from the [web](https://gis.cdc.gov/grasp/fluview/fluportaldashboard.html). 

```{r}
flu_raw <- read_csv("ICL_NREVSS_Clinical_Labs.csv", skip = 1, na = "X") |>
  clean_names()

top5_states <- c("California", "Texas", "Florida", "New York", "Pennsylvania")

flu_df <- flu_raw |>
  rename(state = region,
         total_tests  = total_specimens,
         pct_positive = percent_positive) |>
  filter(state %in% top5_states) |>
  arrange(state, year, week) |>          # chronological order WITHIN each state
  select(state, year, week, total_tests, pct_positive) |>
  group_by(state) |>
  mutate(t = row_number()) |>            # sequential time index that respects year + week
  ungroup() 

flu_df |> count(state)
```

::: {.callout-note}
If I was modeling this data, I would log-transform the pct_positive variable. Here, for clean interpretations, we won't do this.
:::


::: {.callout-important}
`week` alone resets every season (1–52/53), so ordering by `week` without
`year` first silently interleaves different flu seasons — week 40 of 2015
would sort next to week 40 of 2019. `arrange(state, year, week)` fixes the
chronology; `t` is then a clean sequential index per state that both the
rolling-origin split and the fused lasso can rely on.
:::

## Why the Two "Obvious" Approaches Don't Actually Answer the Question

The handout goal is genuine forecasting — total specimens and percent
positive for a **future** week/year, e.g., California, 2026, week 50. That
requirement rules out a couple of approaches:

- **Fused lasso on the raw time series** estimates one coefficient per
  *observed* time point. There is no coefficient for "week 50 of 2026"
  because that row doesn't exist in the fitted path — the model has nothing
  to extrapolate with. It's a good *descriptive* tool for finding past
  regime shifts, but not a predictive one for a genuinely new time point.
- **`week_num` as a sequential index fed into a fixed-effects model** has the
  same problem in a different place: it's tied to the training data's
  specific rows, not a reusable calendar concept. There's no principled way
  to hand the model "week_num = (whatever a 2026 week-50 row would have
  been)."

What actually generalizes to a new year/week is a model with **separate,
interpretable terms for week and year**: a cyclic seasonal effect for week
(defined for all 52 weeks, so no extrapolation needed there at all) and a
linear trend for year (which *can* be extrapolated, honestly, a few years
out). `state` re-enters as a plain fixed effect — a factor term.

## Model: GAM with Cyclic Week Smooth + Linear Year + State Fixed Effect

```{r}
gam_fit <- gam(
  pct_positive ~ state + s(week, bs = "cc", k = 10) + year,
  data   = flu_df,
  knots  = list(week = c(0.5, 52.5)),   # ties week 52 back to week 1
  method = "REML"
)

summary(gam_fit)
plot(gam_fit, select = 1, shade = TRUE, main = "Seasonal effect of week (cyclic)")
```

`s(week, bs = "cc")` is a **cyclic** cubic smooth — it forces the fitted
curve at week 52 to connect back to week 1, the right assumption for a
seasonal effect (there's no real discontinuity between the last week of one
flu season and the first week of the next). `year` enters linearly, so it
contributes a single extrapolable slope rather than a value tied to specific
observed years. `state` is a plain factor term — a fixed effect exactly as
requested, fit directly in `gam()`.

::: {.callout-note}
`refund::pfr()` would fit this identical model — it's built directly on top
of `mgcv::gam()` and accepts the same formula syntax when there's no
functional predictor involved. `gam()` is used here because nothing in this
formula is actually functional (unlike the DTI example); `pfr()` becomes
necessary once you have a genuine curve as a predictor, not before.
:::

## Honest Error Estimation: Rolling-Origin CV, Actually Used

The `roll` object above illustrated the *idea* of rolling-origin splits on a
single state. The number we actually need — the expected forecast error of
the model above — requires applying that same logic to the full pooled panel,
refitting at each cutoff and scoring on the weeks immediately after.

```{r}
flu_df <- flu_df |>
  mutate(calendar_t = (year - min(year)) * 52 + week) |>
  arrange(calendar_t, state)

weeks_sorted  <- sort(unique(flu_df$calendar_t))
initial_weeks <- 104   # ~2 seasons before the first forecast
horizon_weeks <- 4     # forecast 4 weeks ahead at each cutoff

cutoffs <- weeks_sorted[initial_weeks:(length(weeks_sorted) - horizon_weeks)]
cutoffs[1:5]

cv_results <- map_dfr(cutoffs, function(cut) {
  train <- flu_df |> filter(calendar_t <= cut)
  test  <- flu_df |> filter(calendar_t > cut, calendar_t <= cut + horizon_weeks)
  if (nrow(test) == 0) return(NULL)

  fit <- gam(pct_positive ~ state + s(week, bs = "cc", k = 10) + year,
             data = train, knots = list(week = c(0.5, 52.5)), method = "REML")

  tibble(
    cutoff = cut,
    rmse   = sqrt(mean((test$pct_positive - predict(fit, newdata = test))^2, na.rm = TRUE))
  )
})

head(cv_results)

mean(cv_results$rmse)

min_year <- min(flu_df$year)

cv_results <- cv_results |>
  mutate(
    approx_year = min_year + (cutoff - 1) %/% 52,
    approx_week = (cutoff - 1) %% 52 + 1,
    # Approximate date for the start of that MMWR week (good enough for plotting)
    approx_date = as.Date(paste0(approx_year, "-01-01")) + (approx_week - 1) * 7
  )

ggplot(cv_results, aes(approx_date, rmse)) +
  # Shade Dec-Feb each year so winter/flu-season is visible at a glance
  annotate("rect",
           xmin = as.Date(paste0(min_year:(min_year + 10), "-12-01"))[-c(1,2)],
           xmax = as.Date(paste0((min_year + 1):(min_year + 11), "-03-01"))[-c(1,2)],
           ymin = -Inf, ymax = Inf, fill = "steelblue", alpha = 0.08) +
  geom_line() +
  geom_point() +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  labs(title = "Rolling-Origin Forecast Error Over Time",
       x = "Training cutoff (date)", y = "RMSE (4-week-ahead forecast)") +
  theme_minimal(base_size = 13)
```

This refits `gam()` at every cutoff using only data available up to that
point, then evaluates on the following month — exactly how the model would be
used in practice. `mean(cv_results$rmse)` is the number to report as the
model's expected forecast error, not any single train/test split.

::: {.callout-important}
This loop refits the model dozens of times, which is the honest cost of doing
CV properly for a forecasting model. Resist the temptation to shortcut it
with a single holdout at the end — one split gives a noisy point estimate,
not a distribution you can trust.
:::

## Forecasting a Genuinely Future Observation

```{r}
final_fit <- gam(pct_positive ~ state + s(week, bs = "cc", k = 10) + year,
                  data = flu_df, knots = list(week = c(0.5, 52.5)), method = "REML")

future_point <- tibble(state = "California", year = 2026, week = 50)
predict(final_fit, newdata = future_point, se.fit = TRUE)
```

::: {.callout-important}
This prediction is only as trustworthy as the linear extrapolation in `year`.
The seasonal `week` effect is well-supported (fit on the full observed
cycle), but the `year` coefficient is extrapolating past the last observed
year and cannot know about a genuine future disruption (a new variant,
another pandemic-scale event). Report the rolling-origin CV RMSE alongside
any such forecast as the honest uncertainty, and treat a multi-year-out
extrapolation as considerably shakier than a same-season forecast.
:::

The same approach — cyclic week smooth, linear year, state fixed effect,
rolling-origin CV — applies to `total_tests`: refit with
`total_tests ~ state + s(week, bs = "cc", k = 10) + year` and repeat the CV
loop above.

::: {.callout-important}
**The limitation to flag explicitly.** A fixed effect for `state` is
estimated *only* for the states in the training data. There is no mechanism
in this model for predicting a sixth state that wasn't part of the fit — the
model has literally no parameter for it. If the eventual goal is a model that
generalizes to new states (not just the five here), a fixed-effects approach
is the wrong tool; you'd want something that treats state as a random effect
or borrows strength across states (e.g., a hierarchical model), which is
beyond what we've covered so far. Worth having the group articulate this
limitation even though it doesn't change what they fit today.
:::

### Repeating for log-transformed outcome

Below, I'm repeating the entire thing for a log transformed outcome:

```{r}
flu_df <- flu_df |>
  mutate(l_pct_positive = log(if_else(pct_positive==0, 1/total_tests*100, pct_positive))) |>
  arrange(calendar_t, state)

weeks_sorted  <- sort(unique(flu_df$calendar_t))
initial_weeks <- 104   # ~2 seasons before the first forecast
horizon_weeks <- 4     # forecast 4 weeks ahead at each cutoff

cutoffs <- weeks_sorted[initial_weeks:(length(weeks_sorted) - horizon_weeks)]

cv_results <- map_dfr(cutoffs, function(cut) {
  train <- flu_df |> filter(calendar_t <= cut)
  test  <- flu_df |> filter(calendar_t > cut, calendar_t <= cut + horizon_weeks)
  if (nrow(test) == 0) return(NULL)

  fit <- gam(l_pct_positive ~ state + s(week, bs = "cc", k = 10) + year,
             data = train, knots = list(week = c(0.5, 52.5)), method = "REML")

  tibble(
    cutoff = cut,
    rmse   = sqrt(mean((test$pct_positive - predict(fit, newdata = test))^2, na.rm = TRUE))
  )
})

mean(cv_results$rmse)

min_year <- min(flu_df$year)

cv_results <- cv_results |>
  mutate(
    approx_year = min_year + (cutoff - 1) %/% 52,
    approx_week = (cutoff - 1) %% 52 + 1,
    # Approximate date for the start of that MMWR week (good enough for plotting)
    approx_date = as.Date(paste0(approx_year, "-01-01")) + (approx_week - 1) * 7
  )

ggplot(cv_results, aes(approx_date, rmse)) +
  # Shade Dec-Feb each year so winter/flu-season is visible at a glance
  annotate("rect",
           xmin = as.Date(paste0(min_year:(min_year + 10), "-12-01"))[-c(1,2)],
           xmax = as.Date(paste0((min_year + 1):(min_year + 11), "-03-01"))[-c(1,2)],
           ymin = -Inf, ymax = Inf, fill = "steelblue", alpha = 0.08) +
  geom_line() +
  geom_point() +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  labs(title = "Rolling-Origin Forecast Error Over Time",
       x = "Training cutoff (date)", y = "RMSE (4-week-ahead forecast)") +
  theme_minimal(base_size = 13)

final_fit <- gam(l_pct_positive ~ state + s(week, bs = "cc", k = 10) + year,
                  data = flu_df, knots = list(week = c(0.5, 52.5)), method = "REML")

future_point <- tibble(state = "California", year = 2026, week = 50)
pred_vals <- predict(final_fit, newdata = future_point, se.fit = TRUE)

tibble(pred = exp(pred_vals$fit), se.fit = pred_vals$se.fit*exp(pred_vals$fit))
```


## Fused Lasso

For fun, I wanted to fit this model with the fused LASSO. What's cool about this is that I can fuse the coefficients in two directions `year` and `week`. Let's check it out:


```{r}
# ============================================================
# Step 1: Fit the 2D fused lasso on years up to 2024 (holding out 2025)
# ============================================================
fused_train <- flu_df |>
  filter(state == "California") |>
  filter(year > 2015 & year < 2025 & week < 53) |>   # complete years, 2025 excluded
  arrange(year, week)                                 # week must vary fastest within each year

dim1 <- length(unique(fused_train$week))   # 52 weeks
dim2 <- length(unique(fused_train$year))   # years 2016-2024

fl_train <- fusedlasso2d(fused_train$pct_positive, dim1 = dim1, dim2 = dim2)

# Grab a fine path of lambdas so the validation curve below isn't lumpy
a_train <- coef(fl_train, nlam = 30)

# ============================================================
# Keep the 3x3 plot of coefficients from the TRAINING-ONLY fit
# (same 9-lambda snapshot as before, just refit without 2025)
# ============================================================
a9 <- coef(fl_train, nlam = 9)

par(mar = c(1, 1, 2, 1), mfrow = c(3, 3))
cols <- rainbow(100)
zlim <- range(a9$beta)
for (i in 1:9) {
  image(matrix(a9$beta[, i], nrow = dim1), col = cols, zlim = zlim, axes = FALSE)
  mtext(bquote(lambda == .(sprintf("%.3f", a9$lambda[i]))))
}
```


```{r}
# ============================================================
# Step 2: Predict 2025 from the training fit, at every lambda,
# and see which lambda gives the lowest holdout error
# ============================================================

# The fused lasso has no coefficient beyond the training grid, so the natural
# forecast for a year it never saw is the LAST fitted year's curve
# (total-variation regularization assumes adjacent cells are similar --
# "next year looks like last year" is the model's own implied forecast).

weeks_grid <- sort(unique(fused_train$week))

holdout_2025 <- flu_df |>
  filter(state == "California", year == 2025, week < 53) |>
  arrange(week)

forecast_rmse <- map_dbl(seq_along(a_train$lambda), function(j) {
  beta_mat <- matrix(a_train$beta[, j], nrow = dim1, ncol = dim2)
  last_year_col <- beta_mat[, dim2]                # fitted 2024 curve
  names(last_year_col) <- weeks_grid

  pred_2025 <- last_year_col[as.character(holdout_2025$week)]
  sqrt(mean((holdout_2025$pct_positive - pred_2025)^2, na.rm = TRUE))
})

lambda_err <- tibble(lambda = a_train$lambda, rmse = forecast_rmse)

ggplot(lambda_err, aes(lambda, rmse)) +
  geom_line() + geom_point() +
  scale_x_log10() +
  labs(title = "Validation Error by Lambda (2025 Held Out)",
       x = expression(lambda ~ "(log scale)"),
       y = "RMSE, 2025 forecast vs. actual (log percent positive)") +
  theme_minimal(base_size = 13)

lambda_best <- lambda_err$lambda[which.min(lambda_err$rmse)]
lambda_best
```

```{r}
# ============================================================
# Step 3: Refit at lambda_best using ALL data (now including 2025),
# display the coefficients with ggplot
# ============================================================
fused_full <- flu_df |>
  filter(state == "California") |>
  filter(year > 2015 & year < 2026 & week < 53) |>
  arrange(year, week)

dim1_full <- length(unique(fused_full$week))
dim2_full <- length(unique(fused_full$year))

fl_full <- fusedlasso2d(fused_full$pct_positive, dim1 = dim1_full, dim2 = dim2_full)

beta_full <- coef(fl_full, lambda = lambda_best)$beta

plot_full <- fused_full |>
  mutate(Beta = as.vector(beta_full),
         year_f = factor(year))

ggplot(plot_full, aes(week, Beta, color = year_f)) +
  geom_line() + geom_point(size = 0.8) +
  labs(title = paste0("2D Fused Lasso Coefficients, Full Data (\u03bb = ",
                       round(lambda_best, 3), ")"),
       x = "Week", y = "Beta", color = "Year") +
  theme_minimal(base_size = 13)
```

```{r}
# ============================================================
# Step 4: Predict 2025 week 50 from the full refit, then carry that
# forward as the prediction for 2026 week 50
# ============================================================
weeks_grid_full <- sort(unique(fused_full$week))
years_grid_full  <- sort(unique(fused_full$year))

beta_mat_full <- matrix(beta_full, nrow = dim1_full, ncol = dim2_full,
                         dimnames = list(weeks_grid_full, years_grid_full))

pred_2025_wk50 <- beta_mat_full[as.character(50), as.character(2025)]
pred_2026_wk50 <- pred_2025_wk50   # no coefficient exists beyond the observed grid

tibble(
  year = c(2025, 2026),
  week = 50,
  predicted_pct_positive = c(pred_2025_wk50, pred_2026_wk50))
```

::: {.callout-important}
Notice the 2026 prediction is identical to the 2025 one — the fused lasso literally has no parameter for a cell outside the fitted grid, so "predicting 2026" here just means assuming next year looks like the last observed year. That's a fine baseline, but it's a much weaker extrapolation than the year-as-linear-term GAM from earlier in this document, which can actually project a trend forward. Worth having students compare the two 2026 forecasts side by side.
:::


---

# Scenario 3: BGLR Wheat

**Goal:** Using genetic marker data on a panel
of wheat lines, predict yield in a new growing environment using a model
trained on other environments.

**Intended CV issue:** This is a target-context mismatch, not a random-split
problem. The same lines are phenotyped in multiple environments, and the
genetic effect on yield can differ by environment (G×E interaction) — a model
fit on environments 1–3 may not transfer cleanly to environment 4, which is
structurally the same issue as a model failing to generalize across
populations, sites, or cohorts. The honest evaluation is to hold out an
*entire environment*, never a random subset of line-by-environment
combinations.

**Intended method:** Lasso, this will provide variable selection and prediction of new values.

## Data

```{r}
data(wheat)
str(wheat.X[, 1:5])
str(wheat.Y)

X <- wheat.X
Y <- wheat.Y
colnames(Y)
```


::: {.callout-note}
`wheat.Y` has one column per environment. The code below assumes the first column is the held-out one and the last three are the training
environments.
:::

## Held-Out Environment Design

```{r}
train_envs <- 2:4
test_env   <- 1

y_train <- rowMeans(Y[, train_envs], na.rm = TRUE)  # average across training environments
y_test  <- Y[, test_env]

keep <- complete.cases(X) & !is.na(y_train) & !is.na(y_test)
X_fit  <- X[keep, ]
y_fit  <- y_train[keep]
y_eval <- y_test[keep]
```

## Lasso

```{r}
cv_fit <- cv.glmnet(X_fit, y_fit, nfolds = 10, alpha = 1)
plot(cv_fit)

lambda_best <- cv_fit$lambda.min
lambda_1se <- cv_fit$lambda.1se

lasso_min  <- glmnet(X_fit, y_fit, alpha = 1, lambda = lambda_best,  standardize = TRUE)
lasso_1se  <- glmnet(X_fit, y_fit, alpha = 1, lambda = lambda_1se,  standardize = TRUE)
```

## Evaluating Transfer to the Held-Out Environment

```{r}
pred_test <- predict(lasso_min, X_fit)
pred_test_1se <- predict(lasso_1se, X_fit)

tibble(
  metric = c("RMSE", "Correlation", "Non-zero Coeffs"),
  Best  = c(
    sqrt(mean((y_eval - pred_test)^2)),
    cor(y_eval, pred_test), 
    sum(coef(lasso_min )[-1]!=0)
  ),
  One_SE  = c(
    sqrt(mean((y_eval - pred_test_1se)^2)),
    cor(y_eval, pred_test_1se), 
    sum(coef(lasso_1se )[-1]!=0)
  )
)
```

```{r}
tibble(observed = y_eval, predicted = as.numeric(pred_test)) |>
  ggplot(aes(observed, predicted)) +
  geom_point(alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, color = "firebrick", linetype = "dashed") +
  labs(title = "Held-Out Environment: Observed vs. Predicted",
       x = "Observed yield (env 4)", y = "Predicted yield (trained on envs 1-3)") +
  theme_minimal(base_size = 13)
```

## Naive Comparison: Ordinary CV Ignoring Environment

Rather than holding out an entire environment, we could instead pool every
line-by-environment observation into one long dataset and run standard
random k-fold CV on it — treating each row as if it were independent. This
is the version many people reach for by default; here's why it's misleading
for this data.

```{r}
# Stack ALL environments (not just the training ones) into a long dataset:
# one row per line-by-environment combination. Random folds below don't
# know that the same genetic line shows up once per environment.

env_cols <- colnames(Y)[1:4]

y_long  <- as.vector(Y[, env_cols])
X_long  <- X[rep(seq_len(nrow(X)), times = length(env_cols)), ]
line_id <- rep(seq_len(nrow(X)), times = length(env_cols))

keep_long <- complete.cases(X_long) & !is.na(y_long)
X_long_fit  <- X_long[keep_long, ]
y_long_fit  <- y_long[keep_long]
line_long_fit <- line_id[keep_long]

nrow(X_long_fit)
```

```{r}
set.seed(1)
cv_naive <- cv.glmnet(X_long_fit, y_long_fit, nfolds = 10, alpha = 1)
plot(cv_naive)

# cv.glmnet's cvm at the chosen lambda is mean CV squared error
# for a gaussian family, so sqrt(cvm) is directly comparable to the
# RMSE reported for the honest environment-transfer evaluation above.
rmse_naive_cv <- sqrt(min(cv_naive$cvm))
rmse_naive_cv
```

```{r}
tibble(
  Design = c("Naive CV (pooled, random folds, ignores environment)",
             "Honest environment transfer (train envs 1-3, test env 4)"),
  RMSE = c(rmse_naive_cv, sqrt(mean((y_eval - pred_test)^2)))
)

# Percent difference
round((rmse_naive_cv- sqrt(mean((y_eval - pred_test)^2))) /rmse_naive_cv, 3)*100
```

::: {.callout-important}
**Why the naive number looks better.** Each line appears once per
environment, so it contributes up to four rows to `X_long_fit` — one per
environment — all sharing the same genotype. Random folds don't keep a
line's rows together, so a line's environment-4 phenotype can end up in the
validation fold while that *same line's* environment-1/2/3 phenotypes (and
identical `X` row) sit in the training fold. The model can partly recognize
the line from its genotype and lean on what it already "knows" about that
line's phenotype in other environments, rather than learning a genotype
effect that genuinely transfers to a new environment. The gap between these
two RMSEs is an estimate of exactly that — the same leakage mechanism
flagged for NHANES, just clustered by genetic line instead of by PSU.
:::

---


