---
title: "Naive Bayes with Mixed Feature Types — Wisconsin Breast Cancer (WDBC)"
subtitle: "Biostatistics 735/794 — Machine Learning in Public Health (R)"
author: "Alexander McLain"
date: today
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    number-sections: true
    smooth-scroll: true
    theme: cosmo
    code-fold: false
    highlight-style: github
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
lang: en
---

```{r}
#| include: false
set.seed(735794)

# Core
library(tidyverse)
library(rsample)
library(yardstick)
library(knitr)
library(kableExtra)

# Models
library(e1071)   # naiveBayes (Gaussian / mixed-type)
library(klaR)    # NaiveBayes with usekernel = TRUE (kernel-density NB)

select <- dplyr::select 

# Small helper: nice tables
kbl2 <- function(x, digits = 3, ...) kable(x, align = "lrrrrrr", digits = digits, ...) %>% kable_styling(full_width = FALSE)
```

# Setup

We reuse the WDBC data and the train/test split from the previous two
examples.

```{r}
url_data  <- "https://archive.ics.uci.edu/ml/machine-learning-databases/breast-cancer-wisconsin/wdbc.data"
raw <- read.csv(url_data, header = FALSE)

names(raw) <- c("ID","Diagnosis",
                "radius_mean","texture_mean","perimeter_mean","area_mean","smoothness_mean",
                "compactness_mean","concavity_mean","concave_points_mean","symmetry_mean","fractal_dimension_mean",
                "radius_se","texture_se","perimeter_se","area_se","smoothness_se",
                "compactness_se","concavity_se","concave_points_se","symmetry_se","fractal_dimension_se",
                "radius_worst","texture_worst","perimeter_worst","area_worst","smoothness_worst",
                "compactness_worst","concavity_worst","concave_points_worst","symmetry_worst","fractal_dimension_worst")

bc <- raw %>%
  as_tibble() %>%
  mutate(Diagnosis = factor(Diagnosis, levels = c("B","M"))) %>%  # B = negative, M = positive
  select(-ID)

set.seed(42)
split <- initial_split(bc, prop = 0.8, strata = Diagnosis)
train <- training(split)
test  <- testing(split)
```

A helper to compute the metrics we've discussed (AUC is threshold-free; others use a threshold, default 0.5):

```{r}
eval_metrics <- function(truth, prob_pos, thr = 0.5) {
  truth <- factor(truth, levels = c("B","M"))     # "M" is event
  cls   <- factor(ifelse(prob_pos >= thr, "M","B"), levels = c("B","M"))
  tibble(
    AUC  = roc_auc_vec(truth, prob_pos, event_level = "second"),
    Sens = sens_vec(truth, cls,        event_level = "second"),
    Spec = spec_vec(truth, cls,        event_level = "second"),
    PPV  = ppv_vec (truth, cls,        event_level = "second"),
    NPV  = npv_vec (truth, cls,        event_level = "second"),
    Acc  = accuracy_vec(truth, cls)
  )
}
```

---

# Part A — **Gaussian Naive Bayes** (treat all predictors as continuous)

`e1071::naiveBayes` models numeric predictors with a Gaussian density per class and factors with multinomial/Bernoulli. Here we keep **all 30 features numeric** → Gaussian NB.

```{r}
nb_gauss <- naiveBayes(Diagnosis ~ ., data = train, laplace = 0)  # no smoothing needed for continuous

p_tr_g <- predict(nb_gauss, newdata = train, type = "raw")[,"M"]
p_te_g <- predict(nb_gauss, newdata = test,  type = "raw")[,"M"]

metrics_gauss <- bind_rows(
  eval_metrics(train$Diagnosis, p_tr_g) %>% mutate(Set = "Train"),
  eval_metrics(test$Diagnosis,  p_te_g) %>% mutate(Set = "Test")
) %>% relocate(Set)

kbl2(metrics_gauss, caption = "Gaussian Naive Bayes: Performance (threshold = 0.5)")
```

ROC curve (train vs test):

```{r}
train$estimate <- p_tr_g
test$estimate  <- p_te_g
roc_train_g <- yardstick::roc_curve(data = train, Diagnosis, estimate, event_level = "second") %>% mutate(Set = "Train")
roc_test_g  <- yardstick::roc_curve(data = test, Diagnosis,  estimate, event_level = "second") %>% mutate(Set = "Test")

bind_rows(roc_train_g, roc_test_g) %>%
  ggplot(aes(x = 1 - specificity, y = sensitivity, color = Set)) +
  geom_path(linewidth = 1) + geom_abline(slope = 1, intercept = 0, linetype = 3) +
  coord_equal() + theme_minimal(base_size = 13) +
  labs(title = "Gaussian Naive Bayes — ROC", x = "1 - Specificity (FPR)", y = "Sensitivity (TPR)")
```

---

# Part B — **Mixed-Type Naive Bayes** (some binary, some multinomial, some continuous)

Here we transform a few predictors to **binary** (median split) and a few to **multinomial** (tertiles), and leave the rest **continuous**. `naiveBayes` will combine the appropriate likelihoods under the Naive assumption.

**Choices for this demo**

* Binary: `radius_mean`, `texture_mean`, `perimeter_mean` (High vs Low, median split)
* Multinomial (3-level): `area_mean`, `smoothness_mean` (Low/Med/High via tertiles)
* Continuous: all remaining features stay numeric

```{r}
make_mixed <- function(df) {
  out <- df

  # Binary via median split
  for (v in c("radius_mean","texture_mean","perimeter_mean")) {
    med <- median(out[[v]], na.rm = TRUE)
    out[[paste0(v,"_bin")]] <- factor(ifelse(out[[v]] > med, "High","Low"))
    out[[v]] <- NULL
  }

  # Multinomial via tertiles
  for (v in c("area_mean","smoothness_mean")) {
    qs <- quantile(out[[v]], probs = c(1/3, 2/3), na.rm = TRUE)
    cutv <- cut(out[[v]],
                breaks = c(-Inf, qs[1], qs[2], Inf),
                labels = c("Low","Med","High"),
                right = TRUE)
    out[[paste0(v,"_ter")]] <- factor(cutv)
    out[[v]] <- NULL
  }

  out
}

train_mx <- make_mixed(train)
test_mx  <- make_mixed(test)
```

Fit Naive Bayes with **Laplace smoothing** (recommended when categorical levels can be rare):

```{r}
nb_mixed <- naiveBayes(Diagnosis ~ ., data = train_mx, laplace = 1)

p_tr_mx <- predict(nb_mixed, newdata = train_mx, type = "raw")[,"M"]
p_te_mx <- predict(nb_mixed, newdata = test_mx,  type = "raw")[,"M"]

metrics_mixed <- bind_rows(
  eval_metrics(train_mx$Diagnosis, p_tr_mx) %>% mutate(Set = "Train"),
  eval_metrics(test_mx$Diagnosis,  p_te_mx) %>% mutate(Set = "Test")
) %>% relocate(Set)

kbl2(metrics_mixed, caption = "Mixed-Type Naive Bayes (binary + multinomial + continuous): Performance (threshold = 0.5)")
```

ROC curve (train vs test) for the mixed model:

```{r}
train$estimate_mx <- p_tr_mx
test$estimate_mx  <- p_te_mx
roc_train_m <- yardstick::roc_curve(data = train, Diagnosis, estimate_mx, event_level = "second") %>% mutate(Set = "Train")
roc_test_m  <- yardstick::roc_curve(data = test, Diagnosis,  estimate_mx, event_level = "second") %>% mutate(Set = "Test")

bind_rows(roc_train_m, roc_test_m) %>%
  ggplot(aes(x = 1 - specificity, y = sensitivity, color = Set)) +
  geom_path(linewidth = 1) + geom_abline(slope = 1, intercept = 0, linetype = 3) +
  coord_equal() + theme_minimal(base_size = 13) +
  labs(title = "Mixed-Type Naive Bayes — ROC", x = "1 - Specificity (FPR)", y = "Sensitivity (TPR)")
```

---

## Summary So Far

* **Naive Bayes is a framework**: multiply class-conditional likelihoods per feature under conditional independence.
* With `e1071::naiveBayes`, **numeric** variables are modeled as **Gaussian**; **factors** are modeled as **multinomial/Bernoulli** with optional **Laplace smoothing**.
* You can **mix feature types** in the same model (as we did with binary + multinomial + continuous).
* Report both **threshold-free (AUC)** and **thresholded (Se/Sp/PPV/NPV/Acc)** performance.

---

# Part C — **Kernel-Density Naive Bayes** (non-parametric for continuous features)

Not all of the variables have a normal distribution. For example:

```{r}
ggplot(data = train) + geom_histogram(aes(radius_mean))
```

We now replace the Gaussian assumption with a **kernel density estimate
(KDE)** per class, for every continuous feature. This lets us capture
skewness, heavy tails, and multi-modality that a Normal model can't. `klaR::NaiveBayes()` fits exactly this
model with `usekernel = TRUE`: a per-class KDE for every numeric predictor,
combined under the same Naive independence assumption as Parts A and B.

```{r}
nb_kde <- NaiveBayes(Diagnosis ~ ., data = train, usekernel = TRUE)

p_tr_kde <- predict(nb_kde, newdata = train)$posterior[, "M"]
p_te_kde <- predict(nb_kde, newdata = test)$posterior[, "M"]

metrics_kde <- bind_rows(
  eval_metrics(train$Diagnosis, p_tr_kde) %>% mutate(Set = "Train"),
  eval_metrics(test$Diagnosis,  p_te_kde) %>% mutate(Set = "Test")
) %>% relocate(Set)

kbl2(metrics_kde, caption = "Kernel-Density Naive Bayes: Performance (threshold = 0.5)")
```

ROC curves for KDE-NB:

```{r}
train$estimate_kde <- p_tr_kde
test$estimate_kde  <- p_te_kde
roc_train_kde <- yardstick::roc_curve(data = train, Diagnosis, estimate_kde, event_level = "second") %>% mutate(Set = "Train")
roc_test_kde  <- yardstick::roc_curve(data = test, Diagnosis,  estimate_kde, event_level = "second") %>% mutate(Set = "Test")

bind_rows(roc_train_kde, roc_test_kde) %>%
  ggplot(aes(x = 1 - specificity, y = sensitivity, color = Set)) +
  geom_path(linewidth = 1) + geom_abline(slope = 1, intercept = 0, linetype = 3) +
  coord_equal() + theme_minimal(base_size = 13) +
  labs(title = "KDE-Naive Bayes — ROC", x = "1 - Specificity (FPR)", y = "Sensitivity (TPR)")
```

The bandwidth (`bw`) and kernel (`kernel`) arguments to `NaiveBayes()` control
how smooth each class-conditional density is — the KDE analogue of a tuning
parameter, worth adjusting if a density looks too wiggly or too flat.

## Compare Gaussian NB vs Mixed NB vs KDE-NB (test set)

```{r}
compare_tbl <- list(
  `Gaussian NB` = eval_metrics(test$Diagnosis,     p_te_g),
  `Mixed NB`    = eval_metrics(test_mx$Diagnosis,  p_te_mx),
  `KDE-NB`      = eval_metrics(test$Diagnosis,     p_te_kde)
) %>%
  imap_dfr(~ mutate(.x, Model = .y)) %>%
  relocate(Model)

kbl2(compare_tbl, caption = "Test-set comparison: Gaussian NB vs Mixed NB vs KDE-NB (threshold = 0.5)")
```

**Notes**

* KDE-NB **drops the Normality assumption** for continuous features; it can adapt to skewed or multi-modal FNA features.
* Trade-offs: KDE adds flexibility but may overfit with tiny class-specific samples; the bandwidth acts like a **smoothness** tuning parameter, much like `s()` did for GAMs.

---

# Part D — **Naive Bayes with Cost-Sensitive Thresholds**

We are going to use the predicted probabilities from Part C, but now choose the classification
threshold to reflect asymmetric costs rather than defaulting to 0.5.

**Screening context.** Suppose a false negative (calling a diseased patient
healthy) is 10 times as costly as a false positive. When a false negative is
$K$ times as costly as a false positive, the cost-minimizing threshold is
$1/(K+1)$ — here, $1/11$.

**Confirmation context.** Suppose instead a false positive (calling a healthy patient
diseased) is 10 times as
costly as a false negative. When a false positive is $K$ times as costly,
the cost-minimizing threshold is $K/(K+1)$ — here, $10/11$.

```{r}

metrics_overall <- bind_rows(
  eval_metrics(train$Diagnosis, p_tr_kde, thr = 0.5) %>% mutate(Set = "Train"),
  eval_metrics(test$Diagnosis,  p_te_kde, thr = 0.5) %>% mutate(Set = "Test")
) %>% mutate(Type = "Overall", Threshold = 0.5)

metrics_screen <- bind_rows(
  eval_metrics(train$Diagnosis, p_tr_kde, thr = 1/11) %>% mutate(Set = "Train"),
  eval_metrics(test$Diagnosis,  p_te_kde, thr = 1/11) %>% mutate(Set = "Test")
) %>% mutate(Type = "Screen", Threshold = 1/11)

metrics_confirm <- bind_rows(
  eval_metrics(train$Diagnosis, p_tr_kde, thr = 10/11) %>% mutate(Set = "Train"),
  eval_metrics(test$Diagnosis,  p_te_kde, thr = 10/11) %>% mutate(Set = "Test")
) %>% mutate(Type = "Confirm", Threshold = 10/11)

metrics_cost <- bind_rows(metrics_overall, metrics_screen, metrics_confirm) %>%
  relocate(Set, Type, Threshold)

kbl2(metrics_cost %>% filter(Set == "Train"),
     caption = "Gaussian Naive Bayes: Overall vs. Cost-Sensitive Thresholds (Test Set)", digits = 3)

kbl2(metrics_cost %>% filter(Set == "Test"),
     caption = "Gaussian Naive Bayes: Overall vs. Cost-Sensitive Thresholds (Test Set)", digits = 3)
```

**Reading the comparison:**

Under Gaussian NB, none of the test-set predicted probabilities fell between
$1/11$ and $10/11$ — the Gaussian model is confident enough that moving the
threshold anywhere in that range reclassifies nobody, so the overall,
screen, and confirm rows are identical. That's itself worth pointing out:
it's a sign of how peaked the Gaussian posterior tends to be, not a bug.

Switching to KDE-NB, a handful of predicted probabilities do fall in that
band, so the three thresholds now produce different classifications. On the
training set, the results look textbook: moving from screen to confirm
pushes specificity to a perfect 1.00 and PPV to 1.00, with sensitivity and
NPV essentially unchanged. That's the theoretical trade-off working as
expected, and it's not surprising — training data is where the model fits
best, and there's enough data for the general pattern to show through
cleanly.

The test set is the interesting case. Going from screen ($1/11$) to confirm
($10/11$) drops sensitivity from 0.884 to 0.860, while specificity stays
exactly flat at 0.958 — and PPV, NPV, and accuracy all *decrease* slightly
instead of PPV increasing the way theory predicts. This isn't a contradiction
of the theory, it's a consequence of how few observations are actually
driving it. If only one or two malignant cases sit in that borderline
probability range, raising the threshold reclassifies just those one or two
patients from "predicted M" to "predicted B" — which mechanically lowers
sensitivity — while zero benign cases happen to cross the line in the other
direction, so specificity doesn't move at all. With so few borderline cases,
a single reclassified patient can dominate the metrics and push them in a
direction that doesn't match the population-level trade-off the theory
describes.

The lesson: the sensitivity/specificity/PPV/NPV trade-off from moving a
threshold is an *expected* pattern over enough borderline cases, not a
guarantee in any particular finite sample. A test set this size, with only a
few predictions near the cutoff, is exactly where you should expect noisy,
occasionally counter-intuitive movement — and it's a good reminder to look
at *how many* observations a threshold change actually affects before
reading much into the direction of the shift.


```{r}
par(mfrow = c(2, 2))
hist(p_tr_kde[train$Diagnosis == "B"], breaks = 100, main = "Histogram of pi's for Healthy (train)", xlim = c(0,1))
abline(v = c(1/11,1/2,10/11), col = c(3,1,4))
hist(p_te_kde[test$Diagnosis == "B"], breaks = 100, main = "Histogram of pi's for Healthy (test)", xlim = c(0,1))
abline(v = c(1/11,1/2,10/11), col = c(3,1,4))
hist(p_tr_kde[train$Diagnosis == "M"], breaks = 100, main = "Histogram of pi's for Diseased (train)", xlim = c(0,1))
abline(v = c(1/11,1/2,10/11), col = c(3,1,4))
hist(p_te_kde[test$Diagnosis == "M"], breaks = 100, main = "Histogram of pi's for Diseased (test)", xlim = c(0,1))
abline(v = c(1/11,1/2,10/11), col = c(3,1,4))
```

**Notes**

* **For screening:** decreasing the threshold results in more FPs, fewer
  FNs, **higher sensitivity**, lower specificity, **higher NPV**, and lower
  PPV. *The test prioritizes not missing diseased individuals — ruling
  people out.*
* **For confirmation:** increasing the threshold results in fewer FPs, more
  FNs, lower sensitivity, **higher specificity**, lower NPV, and **higher
  PPV.** *The test prioritizes not calling healthy individuals diseased —
  ruling people in.*
* **For overall predictive accuracy** with no clear cost asymmetry, a
  threshold of 0.5 is the natural default.
  
  
  
# GAM for the Breast Cancer Data



```{r}
library(mgcv)

gam_fit <- gam(
  Diagnosis ~ s(radius_worst) + s(radius_se)  + s(texture_mean) + s(concavity_mean),
  data   = train,
  family = binomial
)

summary(gam_fit)
plot(gam_fit, pages = 1, shade = TRUE, seWithMean = TRUE)
```


  