---
title: "CART Examples"
subtitle: "Biostatistics 735 — Machine Learning in Public Health"
author: "Alexander McLain"
date: today
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    number-sections: true
    smooth-scroll: true
    theme: cosmo
    code-fold: false
    highlight-style: github
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| include: false

library(tidyverse)
library(knitr)
library(kableExtra)
library(rpart)
library(rpart.plot)
library(caret)
library(mlbench)
library(naivebayes)   # for Naive Bayes

# Small helper: nice tables
kbl2 <- function(x, ...) kable(x, align = "lrrrrrr", ...) %>% kable_styling(full_width = FALSE)
```

# Continuous Outcome: Communities and Crime Data

We will predict violent crime rate using socioeconomic predictors with the Communities and Crime dataset.

```{r}
df <- read_csv("communities+and+crime/communities.data", col_names = FALSE, na = "?")

names(df) <- c("state","county", "community","communityname", "fold","population", "householdsize","racepctblack", "racePctWhite","racePctAsian", "racePctHisp","agePct12t21", "agePct12t29","agePct16t24", "agePct65up","numbUrban", "pctUrban","medIncome", "pctWWage","pctWFarmSelf", "pctWInvInc","pctWSocSec", "pctWPubAsst","pctWRetire", "medFamInc","perCapInc", "whitePerCap","blackPerCap", "indianPerCap","AsianPerCap", "OtherPerCap","HispPerCap", "NumUnderPov","PctPopUnderPov", "PctLess9thGrade","PctNotHSGrad", "PctBSorMore","PctUnemployed", "PctEmploy","PctEmplManu", "PctEmplProfServ","PctOccupManu", "PctOccupMgmtProf","MalePctDivorce", "MalePctNevMarr","FemalePctDiv", "TotalPctDiv","PersPerFam", "PctFam2Par","PctKids2Par", "PctYoungKids2Par","PctTeen2Par", "PctWorkMomYoungKids","PctWorkMom", "NumIlleg","PctIlleg", "NumImmig","PctImmigRecent", "PctImmigRec5","PctImmigRec8", "PctImmigRec10","PctRecentImmig", "PctRecImmig5","PctRecImmig8", "PctRecImmig10","PctSpeakEnglOnly", "PctNotSpeakEnglWell","PctLargHouseFam", "PctLargHouseOccup","PersPerOccupHous", "PersPerOwnOccHous","PersPerRentOccHous", "PctPersOwnOccup","PctPersDenseHous", "PctHousLess3BR","MedNumBR", "HousVacant","PctHousOccup", "PctHousOwnOcc","PctVacantBoarded", "PctVacMore6Mos","MedYrHousBuilt", "PctHousNoPhone","PctWOFullPlumb", "OwnOccLowQuart","OwnOccMedVal", "OwnOccHiQuart","RentLowQ", "RentMedian","RentHighQ", "MedRent","MedRentPctHousInc", "MedOwnCostPctInc","MedOwnCostPctIncNoMtg", "NumInShelters","NumStreet", "PctForeignBorn","PctBornSameState", "PctSameHouse85","PctSameCity85", "PctSameState85","LemasSwornFT", "LemasSwFTPerPop","LemasSwFTFieldOps", "LemasSwFTFieldPerPop","LemasTotalReq", "LemasTotReqPerPop","PolicReqPerOffic", "PolicPerPop","RacialMatchCommPol", "PctPolicWhite","PctPolicBlack", "PctPolicHisp","PctPolicAsian", "PctPolicMinor","OfficAssgnDrugUnits", "NumKindsDrugsSeiz","PolicAveOTWorked", "LandArea","PopDens", "PctUsePubTrans","PolicCars", "PolicOperBudg","LemasPctPolicOnPatr", "LemasGangUnitDeploy","LemasPctOfficDrugUn", "PolicBudgPerPop","ViolentCrimesPerPop")

df_nomiss <- df %>% 
  select(-c(state:fold,LemasSwornFT:PolicAveOTWorked, PolicCars:LemasGangUnitDeploy, PolicBudgPerPop)) %>% 
  filter(complete.cases(.))

head(df_nomiss)
```

Distribution of outcome by candidate predictors

```{r}
ggplot(df_nomiss, aes(x = PctIlleg, y = ViolentCrimesPerPop)) + geom_point(alpha = 0.6) +
  geom_smooth(method = "loess") +
  labs(title = "Outcome distribution by house hold size")

ggplot(df_nomiss, aes(x = PctFam2Par, y = ViolentCrimesPerPop)) + geom_point(alpha = 0.6) +
  geom_smooth(method = "loess") +
  labs(title = "Outcome vs. Median Income")
```

Fit CART with cross-validation to select CP

```{r}
set.seed(735)

fit.n <- rpart(ViolentCrimesPerPop~., 
               data=df_nomiss, 
               method="anova", 
               control = list(minbucket = 50,
                              minsplit = 70,
                              cp = 0.0000002, 
                              xval = 50))

Fit_table.n <- as.data.frame(fit.n$cptable)
kbl2(Fit_table.n, caption = "50-Fold CV Results for CART")

plot(Fit_table.n$nsplit,Fit_table.n$xerror,
     xlab = "Number of splits",ylab="CV error",
     main = "CV error by number of splits",
     type = "l", xlim=c(0,min(c(max(Fit_table.n$nsplit),100))))
abline(h=1,lty=3)


Best_mod <- which.min(Fit_table.n$xerror)
Best_mod_1sd <- which.max(Fit_table.n$xerror < c(Fit_table.n$xerror+Fit_table.n$xstd) [which.min(Fit_table.n$xerror)])
abline(v = Fit_table.n$nsplit[Best_mod] ,lwd=3)
abline(v = Fit_table.n$nsplit[Best_mod_1sd], lwd = 3, col = 4)
```

## Final tree

Here, we'll use the 1-sd criteria to get our final tree:

```{r}
j <- Best_mod_1sd
kbl2(Fit_table.n[j,])
# Get Cp for the best model
cp <- Fit_table.n$CP[j]
# Get the tree for the best model
fit.best <- prune(fit.n,cp=cp)
rpart.plot(fit.best,extra = 101)

```

Some of the variables:

- PctIlleg: percentage of kids born to never married (numeric - decimal)
- PctFam2Par: percentage of families (with kids) that are headed by two parents (numeric - decimal)
- HousVacant: number of vacant households (numeric - decimal)
- MalePctDivorce: percentage of males who are divorced (numeric - decimal)
- PctKids2Par: percentage of kids in family housing with two parents (numeric - decimal)
- NumIlleg: number of kids born to never married (numeric - decimal)
- PctPersDenseHous: percent of persons in dense housing (more than 1 person per room) (numeric - decimal)
- NumStreet: number of homeless people counted in the street (numeric - decimal)

⸻

# Binary Outcome: Pima Indians Diabetes

Next, we classify diabetes status using the Pima Indians Diabetes dataset.

```{r}
PimaIndiansDiabetes <- read_csv("pima.indians.diabetes3.csv")
bin_df <- na.omit(PimaIndiansDiabetes) |> 
  rename(diabetes = class) |> 
  select(-classdigit)
head(bin_df)
```


Distribution of outcome by candidate predictors

```{r}
ggplot(bin_df, aes(x = diabetes, y = glucose)) + geom_boxplot() +
  labs(title = "Glucose by Diabetes Status")

ggplot(bin_df, aes(x = diabetes, y = age)) + geom_boxplot() +
  labs(title = "Age by Diabetes Status")
```

## Using `caret` for Model Tuning

So far, we have fit models directly using functions such as `rpart()`. Here we will use the `caret` package.

`caret` provides a common framework for:

1. fitting a model,
2. choosing tuning parameters using cross-validation, and
3. comparing predictive performance across models.

The first step is to tell `caret` **how we want model performance to be evaluated**.

```{r}
set.seed(735)

train_control <- trainControl(
  method = "cv",
  number = 10,
  classProbs = TRUE,
  summaryFunction = twoClassSummary
)
```

`trainControl()` does **not fit a model**. It defines the resampling procedure that `caret` will use later.

Here we are asking for:

* `method = "cv"`: use cross-validation.
* `number = 10`: use 10 folds.
* `classProbs = TRUE`: calculate predicted probabilities in addition to predicted classes.
* `summaryFunction = twoClassSummary`: calculate ROC AUC, sensitivity, and specificity within the validation folds.

We can now use this same resampling plan to fit different classification models.

## Fitting and Tuning CART

```{r}
cart_class <- train(
  diabetes ~ .,
  data = bin_df,
  method = "rpart",
  trControl = train_control,
  metric = "ROC",
  tuneLength = 10
)

cart_class
```

The `train()` function coordinates the entire model-fitting and tuning process.

Here:

* `diabetes ~ .` defines the outcome and predictors.
* `method = "rpart"` tells `caret` which type of model to fit.
* `trControl = train_control` tells `caret` to use the 10-fold CV procedure defined above.
* `metric = "ROC"` tells `caret` to choose the tuning parameter that maximizes cross-validated ROC AUC.
* `tuneLength = 10` asks `caret` to consider 10 candidate values of the tuning parameter.

For CART, the tuning parameter is the complexity parameter:

\[
cp.
\]

Conceptually, `caret` is doing the following:

1. Select candidate values of $cp$.
2. Divide the data into 10 folds.
3. For each value of $cp$:
   * fit CART using 9 folds,
   * predict observations in the remaining fold,
   * calculate ROC AUC,
   * repeat until every fold has served as the validation set.
4. Average the ROC AUC across the 10 folds.
5. Select the value of $cp$ with the largest cross-validated ROC AUC.
6. Fit the final CART model to the full dataset using the selected $cp$.

Thus, `caret` separates two important decisions:

> **What model should we fit?**  
> `method = "rpart"`

and

> **How should we tune and evaluate it?**  
> `trControl = train_control` and `metric = "ROC"`

## Looking at the Tuning Results

The cross-validation results for the candidate values of $cp$ are stored in:

```{r}
cart_class$results
```

We can also plot them:

```{r}
plot(cart_class)
```

The selected tuning parameter is:

```{r}
cart_class$bestTune
```

## Final CART Tree

After selecting $cp$ using cross-validation, `caret` refits CART using the full dataset.

The resulting `rpart` model is stored in:

```{r}
best_tree_class <- cart_class$finalModel

rpart.plot(
  best_tree_class,
  main = "CART Classification Tree",
  extra = 101
)
```

The important distinction is:

* `cart_class` contains the **entire caret analysis**, including the CV results and tuning information.
* `cart_class$finalModel` contains the **final fitted CART model** selected by that procedure.
⸻

# Compare CART vs Naive Bayes for Pima Indians Diabetes

## Using the Same Framework for Naive Bayes

One advantage of `caret` is that the same general workflow can be used for many different machine learning methods.

For CART we used:

```r
method = "rpart"
```

For Naive Bayes we simply change the modeling method:

```{r}
nb_grid <- expand.grid(
  usekernel = c(TRUE, FALSE),
  laplace   = c(0, 1, 2),
  adjust    = c(0.5, 1, 1.5, 2)
) |> 
  # adjust only matters when usekernel = TRUE
  filter(!(!usekernel & adjust > 0.5))

set.seed(735)

nb_model <- train(
  diabetes ~ .,
  data = bin_df,
  method = "naive_bayes",
  trControl = train_control,
  metric = "ROC",
  tuneGrid = nb_grid
)
```

Notice that we use the **same `train_control` object**.

Thus, CART and Naive Bayes are evaluated using the same:

* 10-fold cross-validation strategy,
* prediction metric (ROC AUC), and
* general model-selection framework.

What changes is the statistical model being fit.

```{r}
nb_model$results|> kbl2(digits = 3)
nb_model$bestTune |> kbl2()
```


## Compare ROC AUC

```{r}
results <- resamples(list(CART = cart_class, NB = nb_model))
```

**Code Explanation**

•	`resamples()` is a caret function that takes multiple fitted caret models (here cart_class and nb_model) and extracts the resampling results (cross-validation scores).

•	`list(CART = cart_class, NB = nb_model)` names each model so they can be compared side by side.

```{r}
summary(results)
bwplot(results, main = "CART vs Naive Bayes")
```

⸻

Key Takeaways

•	CART flexibly partitions the predictor space for continuous or binary outcomes.

•	Cross-validation is essential for tuning tree complexity (cp).

•	CART trees are interpretable but may not always perform as well as other classifiers.

•	Naive Bayes provides a useful baseline for classification problems and often complements CART in comparison.

