---
title: "12 - Random Forests"
subtitle: "Biostatistics 735 — Machine Learning in Public Health"
author: "Alexander McLain"
date: today
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    number-sections: true
    smooth-scroll: true
    theme: cosmo
    code-fold: false
    highlight-style: github
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  cache: true
---

```{r}
#| include: false
library(tidyverse)
library(kableExtra)
library(caret)
library(randomForest)
library(mlbench)
library(pdp) # for partial dependence plots
library(yardstick)
kbl2 <- function(x, ...) kable(x, align = "lrrrrrr", ...) %>% kable_styling(full_width = FALSE)
```

# Random Forests for Regression

We begin with the Communities and Crime dataset.  
Random forests improve CART by averaging many trees, each grown on a bootstrap sample and using a random subset of predictors at each split.

```{r}
# Load Communities and Crime data 
crime <- readRDS("community_crime.RDS")

set.seed(735)
train_index <- createDataPartition(crime$ViolentCrime, p = 0.7, list = FALSE)
train <- crime[train_index, ]
test  <- crime[-train_index, ]
```

## Comparing test error across `mtry` and number of trees

```{r}
results <- data.frame()

for (m in c(2, 3, 4, 6, 8, 10)) {
  for (nt in c(50, 100, 200, 500, 1000, 2000, 3000)) {
    rf_fit <- randomForest(ViolentCrime ~ ., data = train,
                           mtry = m, ntree = nt)
    pred <- predict(rf_fit, newdata = test)
    test_mse <- mean((test$ViolentCrime - pred)^2)
    results <- rbind(results,
                     data.frame(mtry = m, ntree = nt, TestMSE = test_mse))
  }
}

kbl2(results %>% arrange(TestMSE) %>% filter(row_number() <= 15), )
```

```{r}
ggplot(results, aes(x = ntree, y = TestMSE, color = factor(mtry))) +
  geom_line() + geom_point() +
  labs(title = "Test Error across mtry and ntree",
       color = "mtry") +
  theme_minimal()
```


## Cross-validation to select mtry with caret

```{r}
set.seed(123)

# Define training control for 10-fold CV
ctrl <- trainControl(method = "cv", number = 10)

# Train random forest model with CV to tune mtry
rf_caret <- train(ViolentCrime ~ ., 
                  data = train,
                  method = "rf",
                  metric = "RMSE",  
                  trControl = ctrl,
                  tuneGrid = data.frame(mtry = 1:7)) # grid of mtry values

rf_caret
plot(rf_caret)
```

Explanation of the code:

•	`trainControl(method = "cv", number = 10)`: runs 10-fold CV.
•	`metric = "RMSE"`: we’re using RMSE to pick the best mtry.
•	`tuneGrid = data.frame(mtry = 1:7)`: defines the candidate values of mtry to try (from 1 up to 7, since Pima has 8 predictors).
•	`plot(rf_caret)`: shows how AUC changes as mtry changes.

## Variable Importance in Regression

```{r}
rf_reg <- rf_caret$finalModel
varImpPlot(rf_reg, main = "Variable Importance (Regression)")
```

## Partial Dependence Plots (PDPs)

Partial dependence plots (PDPs) help us understand how a single predictor (or a pair of predictors) influences the model’s predictions, while averaging out the effects of all other predictors.  

- For regression models, the y-axis shows the average predicted outcome as the predictor changes.  
- For classification models, the y-axis typically shows the average predicted probability of the positive class.  
- PDPs are useful for interpreting "black box" models like random forests, since they highlight nonlinear effects and interactions that may not be obvious from variable importance alone.  

In short, PDPs answer: *If we vary one predictor, holding everything else constant on average, how do predictions change?*

```{r}
p1 <- partial(rf_reg, pred.var = "PctUnemployed", train = train)
autoplot(p1, main = "PDP: % Unemployed")

p2 <- partial(rf_reg, pred.var = "PctPopUnderPov", train = train)
autoplot(p2, main = "PDP: % Poverty")
```

---

# Random Forests for Classification

We now turn to the Pima Indians Diabetes dataset.

```{r}
PimaIndiansDiabetes <- read_csv("pima.indians.diabetes3.csv")
pima <- na.omit(PimaIndiansDiabetes) |> 
  rename(diabetes = class) |> 
  mutate(diabetes = as.factor(diabetes)) |> 
  select(-classdigit)

set.seed(735)
train_index <- createDataPartition(pima$diabetes, p = 0.7, list = FALSE)
train <- pima[train_index, ]
test  <- pima[-train_index, ]
```

## Fit Random Forest Classifier

First, let's just fit for a fixed `mtry` and number of trees.

```{r}
rf_class <- randomForest(diabetes ~ ., data = train,
                         mtry = 3, ntree = 500, importance = TRUE)
rf_class
```


## Cross-validation to select mtry with caret

```{r}
set.seed(123)

# Define training control for 10-fold CV
ctrl <- trainControl(method = "cv", number = 10, 
                     classProbs = TRUE, 
                     summaryFunction = twoClassSummary)

# Train random forest model with CV to tune mtry
rf_caret <- train(diabetes ~ ., 
                  data = pima,
                  method = "rf",
                  metric = "ROC",   # use AUC as performance measure
                  trControl = ctrl,
                  tuneGrid = data.frame(mtry = 1:7)) # grid of mtry values

rf_caret
plot(rf_caret)
```

## Variable Importance for Final Model

```{r}
rf_finalModel <- rf_caret$finalModel
varImpPlot(rf_finalModel, main = "Variable Importance (Classification)")
```

## Partial Dependence Plots (Classification)

```{r}
p3 <- partial(rf_finalModel, pred.var = "glucose", train = train, which.class = "diabetic")
autoplot(p3, main = "PDP: Glucose (Diabetes Probability)")

p4 <- partial(rf_finalModel, pred.var = "age", train = train, which.class = "diabetic")
autoplot(p4, main = "PDP: Age (Diabetes Probability)")
```

## Test Error (or lack thereof)

```{r}
pred <- predict(rf_finalModel, newdata = test, type =  "prob")
table(pred[,1] < 0.5, test$diabetes)

```


---

# Key Takeaways

- Random forests reduce the variance of CART by averaging many trees.  
- Performance depends on `mtry` (predictor subset size) and the number of trees.  
- Variable importance measures highlight the most influential predictors.  
- Partial dependence plots show how predictors influence the predicted outcome.  
- In public health, random forests are widely used for both prediction and feature selection.
```