---
title: "Homework One"
subtitle: "Biostatistics 735 — Machine Learning in Public Health"
date: August 18, 2026 
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    number-sections: true
    smooth-scroll: true
    theme: cosmo
    code-fold: false
    highlight-style: github
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---


# Instructions

This homework is **due August 31, 2026**. Here is how to get started:

**Step 1.** Open a new Quarto document: File → New File → Quarto Document.
Give it a title like `"Homework 1"` and select HTML. Click **Create**.

**Step 2.** Delete the template content that Quarto pre-fills and replace it
with a YAML header and setup chunk matching the one from Part 1:

````markdown
---
title: "Homework One"
subtitle: "Biostatistics 735 — Machine Learning in Public Health"
date: today
format:
  html:
    toc: true
    theme: cosmo
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

```{{r}}
library(tidyverse)
library(this.path) 
this.dir <- this.dir()
setwd(this.dir)

framingham <- read_csv("framingham.csv")
```
````


**Step 3.** Work through Exercises 1--4 in order. For each one:

- Add a new code chunk (`Ctrl+Alt+I` / `Cmd+Option+I`)
- Write and run your code
- Add a sentence or two below the chunk describing what you found in plain
  language --- not just what the numbers are, but what they mean

**Step 4.** When you have finished all exercises, click **Render**
(`Ctrl+Shift+K` / `Cmd+Shift+K`) and confirm the document compiles to HTML
without errors. Fix any errors before moving on.

---

# Exercises

**Exercise 1.**
Read `framingham.csv` using `read_csv()`. The dataset uses blank cells for missing values, which `read_csv()` handles automatically.After reading in:

a. Use `str()` to check column types. Which variables are stored as `<dbl>` but
   should be factors?
b. Convert at least four binary or categorical variables to factors with
   meaningful labels using `mutate()` and `factor()`.
c. Write the cleaned dataset to a new file called `framingham_clean.csv` using
   `write_csv()`.

```{r}
# your code here
```

 **Exercise 2.**
Build a single pipeline that:

(a) Keeps only participants aged 40--65
(b) Creates a new variable `hyp_smoker`: `"Yes"` if the participant has
    prevalent hypertension **and** is a current smoker, `"No"` otherwise
    (use `if_else()` or `case_when()`)
(c) Computes, for each combination of `hyp_smoker` and `male`: number of
    participants, mean systolic BP, and 10-year CHD rate
(d) Keeps only groups with at least 50 participants
(e) Sorts by CHD rate descending

What combination of characteristics is associated with the highest CHD risk
in this age group?

```{r}
# your code here
```


**Exercise 3 **

a. Create a dataset called `flight_avg_delay` by:

   - Grouping `flights` by destination (`dest`)
   - Computing the mean arrival delay (`avg_delay`) for each destination,
     ignoring NAs
   - Joining in the latitude (`lat`), longitude (`lon`), and full airport name
     (`name`) from `airports` using `left_join()` with `by = c("dest" = "faa")`
   - Dropping destinations with no matching airport record using `filter()`
     and `!is.na(lat)`

b. How many destinations have no match in `airports`? Use `anti_join()` to
   find them before building the dataset.

c. Once you have `flight_avg_delay`, run this code to plot average arrival delay
   on a US map:

```{r}
#| eval: false

# This requires the maps package.
install.packages("maps") 
library(maps)

flight_avg_delay |>
  ggplot(aes(lon, lat, color = avg_delay)) +
  borders("state") +
  geom_point(size = 2) +
  coord_quickmap() +
  scale_color_gradient2(low = "blue", mid = "white", high = "red",
                        midpoint = 0) +
  labs(title = "Average Arrival Delay by Destination Airport",
       color  = "Avg delay (min)")
```

Which regions of the country have the worst average arrival delays?

```{r}
# your code here
```



**Exercise 4 **

Build a single polished figure that answers this question: **How does the
10-year coronary heart disease rate in the Framingham cohort vary by age group,
sex, and hypertension status?**

Steps:

a. Create an age group variable using `case_when()`:
   "40--44", "45--49", "50--54", "55--59", "60--64", "65+"
   (or similar grouping of your choosing). Make it a factor with levels in
   the right order.

b. Compute the CHD rate (proportion with `TenYearCHD == "Yes"`) for each
   combination of age group, sex, and hypertension status using
   `group_by()` and `summarize()`. Keep only groups with at least 20
   participants.

c. Plot the result. A grouped bar chart (`geom_col()` with `position = "dodge"`)
   or a line plot (`geom_line()` + `geom_point()`) with age group on the x-axis
   works well here. Color by hypertension status, facet by sex.

d. Add a full set of labels, a meaningful color palette, and a clean theme.

e. Save the figure with `ggsave()` at 300 dpi.

f. Write a 3--4 sentence caption summarizing the main findings as if this were
   going into a manuscript.

```{r}
# your code here
```
